---
format: 
  html:
    embed-resources: true
execute:
  freeze: true
bibliography: survival_marginaleffects.bib
---

# Survival analysis

Survival analysis is concerned with analyzing the (causal) effects of one or more variables on a time-to-event outcome, such as time until death or time until the occurrence of a disease. One of the main issues of survival analysis in practice is that the time-to-event outcomes are often right-censored, meaning that it is only known that the event of interest did not occur up until some point in time for some individuals. The most popular class of models that directly deals with right-censoring are Cox proportional hazards models [@Cox1972]. Although very useful, these models are not marginal models. They only produce hazard ratios (HR), which are conditional quantities. As such, they do not directly estimate practically relevant causal estimands by themselves [@Hernan2010]. This makes the parameter estimates produced by such models quite difficult to interpret in practical settings, an issue that is shared by a majority of other time-to-event models as well.

To interpret results from survival models, authors have recommended reporting marginal estimands, such as the difference or ratio between counterfactual survival probabilities [@Klein2007; @Royston2013; @Uno2014]. These counterfactual probabilities may also be presented visually as "adjusted survival curves," allowing for easier presentation of the results [@Denz2023]. A popular and statistically efficient way to calculate these quantities is the use of g-computation [@Robins1986]. 

The `marginaleffects` package is uniquely suited to perform these tasks. It allows users to compute a wide array of intuitive and meaningful quantities of interest, and to compare them to one another. The package also allows users to easily specify complex hypothesis tests, to study interactions or effect modification.

Crucially, the syntax and approach used to interpret survival models is exactly the same as for any of the 100+ model types supported by `marginaleffects`. Users who have read the (free) *Model to Meaning* will feel right at home. 

This vignette illustrates how to use `marginaleffects` to estimate various marginal causal estimands in survival analysis on the basis of time-to-event models. Our goal is mainly to give a non-technical overview of the capabilities of the package. For more technical and detailed explanations of the shown material we recommend consulting standard textbooks on survival analysis.

## Example Data and Model

Let's consider a classic real-data example from the survival analysis literature: the `rotterdam` dataset from the `survival` R package [@Hofman1991]. The `rotterdam` dataset, obtained from the Rotterdam tumor bank, contains information on 2982 breast cancer patients. It includes variables related to patient demographics, clinical characteristics, and survival outcomes. The dataset is used in survival analysis to study the relationship between predictors and survival time.

For the purposes of this vignette, we will try to analyze the causal effect of a hormonal treatment on the time until death. In the `rotterdam` dataset, the hormonal treatment is a binary variable called `hormon`, where `0` indicates no treatment and `1` means that the patient did receive a treatment. The time until death is coded as a classic time-to-event variable with two columns: `dtime` and `death`. Here, `dtime` is the time until death or right-censoring in days and `death`is the event status indicator, where `0` indicates that the patient was still alive at `dtime` and `1` indicates that the patient died at `dtime`. The study did not randomize the `hormon` treatment, meaning that we have to adjust for confounding as is usual in observational data.

The first 6 rows of the dataset look like this:

```{r}
#| include: false
options(marginaleffects_print_type = FALSE)
options(marginaleffects_print_column_names = FALSE)
options(width = 300)
``` 
```{r}
library(marginaleffects)

# model-fitting packages
library(survival)
library(splines)

# plotting packages and theme
library(tinyplot)
tinytheme("ipsum")

# parallel processing for bootstrap confidence intervals
library(future.apply)
plan(multisession)
options(marginaleffects_parallel_inferences = TRUE)
options(marginaleffects_parallel_packages = c("survival", "splines"))

head(rotterdam)
```

For the analysis of the data we will first fit a standard Cox proportional hazards model with the time to death or right-censoring as the response variable and the hormonal treatment as independent variable. Additionally, we will include `grade`, which is an ordinal variable reflecting the level of tumor differentiation, and `age` at surgery, a continuous variable including the age in years as further independent variables. To account for a possible non-linear relationship between `age` and the time until death we model this variable using a natural spline with two degrees of freedom, as shown below:

```{r}
model <- coxph(
    Surv(dtime, death) ~ hormon + grade + ns(age, df = 2),
    data = rotterdam
)

summary(model)
```

From the `summary()` output alone it looks like the `hormonal` treatment leads to a faster time until death on average with a HR of ~ 1.34, even after adjustment for `age` and `grade`. We will investigate this further below.

::: {.callout-warning}
This notebook reports uncertainty estimates obtained via non-parametric bootstrapping, using the `vcov="rsample"` argument. This is important because the default standard errors in `marginaleffects` may be severely anti-conservative. See the section on "Uncertainty" below for an important note on this topic.
:::

## Average Counterfactual Predictions

We start by performing [G-Computation](https://marginaleffects.com/chapters/gcomputation.html) to compute and plot [average counterfactual predictions](https://marginaleffects.com/chapters/predictions.html) for each of the two possible values of `hormon` as a function of time. The process is as follows:

1. Replicate the entire dataset once for each combination of `hormon` and `dtime` that we wish to consider.
2. Compute the predicted survival probability for every row in this grid.
3. Take the average of the predicted survival probabilities for each combination of `hormon` and `dtime`.

The function we use to implement these steps is `avg_predictions()`. We set `type="survival"` to obtain the average counterfactual predictions on the survival probability scale. Other options would be to use `type="risk"` to obtain the predictions on a death probability scale instead. 

The `by` argument indicates that we want to marginalize survival probabilities over each unique combination of `hormon` and `dtime`.

The `vcov` argument is set to `"rsample"` to use non-parametric bootstrapping to compute the standard errors, implemented by the `rsample` package.

Because we want to calculate counterfactual predictions, we also had to define the `newdata` argument using a call to the `datagrid()` function. As usual, this argument controls over which values of the predictors the predictions should be made. Since there are many `dtime` values, replicating the full dataset would be very memory- and computationally-intensive. Therefore, we specify a grid with 25 equally spaced points in time between the first and the last occurring event time. Then, we feed the grid to the `avg_predictions()` function.

```{r}
p <- avg_predictions(model,
    type = "survival",
    by = c("dtime", "hormon"),
    vcov = "rsample",
    newdata = datagrid(
        hormon = 0:1,
        dtime = seq(36, 7043, length.out = 25),
        grid_type = "counterfactual"
    )
)
 
tail(p)
```

The `Estimate` column shows the estimated average "counterfactual" survival probability at `dtime` for `hormon = 0` and `hormon = 1`. These can be interpreted as:

> the fraction of individuals that we would expect to survive, on average, up to `dtime`, if every individual in the population had or had not received hormonal treatment.

Consider this subset of estimates, where `dtime` is at its maximum value:

```{r}
last <- p[p$dtime == max(p$dtime),]
last
```

The first row suggests that if we had *not* intervened on any individual in the dataset, our model expects that about `r sprintf("%.0f%%", last$estimate[1] * 100)` percent of individuals would still be alive by `dtime`=`r last$dtime[1]`. In contrast, the second row suggests that if we had intervened on every individual in the dataset, our model expects that only about `r sprintf("%.0f%%", last$estimate[2] * 100)` percent of individuals would still be alive by `dtime`=`r last$dtime[1]`.

::: {.callout-warning}
This counterfactual interpretation of the estimated average survival probabilities only holds if the model is correctly specified, and if four fundamental causal identification assumptions hold:

1. positivity 
2. counterfactual consistency 
3. no-interference 
4. conditional exchangeability 

These assumptions are described in detail in @Hernan2020 and many other books and articles on causal inference based methods.
:::

### Plot

Since the output of `avg_predictions()` is a data frame, we can use it directly to plot the results, using `ggplot2`, base `R`, or any other plotting package. Here, we use the `tinyplot` package to display the adjusted survival curves:

```{r}
#| warnings: false
with(p, {
    tinyplot(
        x = dtime,
        y = estimate,
        ymin = conf.low,
        ymax = conf.high,
        by = hormon,
        type = "ribbon",
        ylab = "Adjusted Survival Probability"
    )
})
```

The resulting plot shows the survival curves for both treatment options of `hormon`, adjusted for `age` and `grade`. As can be seen quite clearly from the plot alone, the confidence intervals do not overlap for the majority of the shown time, but not all points in time. Using only the HR from the model, we would have been unable to make such distinctions or plot this graphic.

### Hypothesis Testing

The plot in the previous section showed a difference between the two `hormon` groups at almost every point in time. The confidence intervals did not overlap much, which could lead the analyst to believe that the difference between surivival probabilities in the treatment and control groups are distinguishable. In some situations, however, it might be necessary to formally compare the survival probabilities at a specific point in time using a statistical hypothesis test [@Klein2007]. In other words, one may wish to check if we can reject this null hypothesis:

> At time $t$, the difference between the estimated survival probability for patients in group `hormon=1` and `hormon=1` is equal to zero. 

This kind of null hypothesis can be tested easily with the `avg_predictions()` function. In the code below, we specify a new grid with three different vales of `dtime`: 1000, 4000, and 7000. Then, we use [the convenient formula syntax](https://marginaleffects.com/chapters/hypothesis.html) of the `hypothesis` argument to compute differences in estimated survival probabilities for each `dtime`.

```{r}
p <- avg_predictions(model,
  hypothesis = difference ~ reference | dtime,
  type = "survival",
  by = c("dtime", "hormon"),
  vcov = "rsample",
  newdata = datagrid(
    hormon = 0:1,
    dtime = c(1000, 4000, 7000),
    grid_type = "counterfactual"
  )
)
p
```

Note that the syntax is essentially identical to the syntax used when estimating the adjusted survival curves, with the only difference being that we focus on three points in time in the `datagrid()` call, and that we additionally specified the `hypothesis` argument. 

The `Estimate` column of the output now includes the estimated difference between counterfactual survival probabilities at $t\in\{1000,4000,7000\}$. For all three time points, the difference in expected survival probabilities for the two `hormon` groups is negative, and the confidence intervals exclude zero. This means that, for all three time points, we can reject the null hypothesis that the survival probabilities are equal for patients receiving hormonal treatment and those not receiving it.

Another option would be to use the *ratio* of the two survival probabilities instead. To do this, we only change the left-hand side of the `hypothesis` formula.

```{r}
p <- avg_predictions(model,
  hypothesis = ratio ~ reference | dtime,
  type = "survival",
  by = c("dtime", "hormon"),
  vcov = "rsample",
  newdata = datagrid(
    hormon = 0:1,
    dtime = c(1000, 4000, 7000),
    grid_type = "counterfactual"
  )
)
p
```

Here, the `Estimate` argument shows the ratio of the two survival probabilities at $t$ instead of the difference, again with associated 95% confidence interval. The three intervals exclude 1, which means we can reject the null hypothesis that estimated survival is identical in both `hormon` groups.

As noted in [the hypothesis testing chapter](https://marginaleffects.com/chapters/hypothesis.html), users may define much more complicated hypotheses or equivalence tests, to compare subgroups and look into interactions or treatment effect heterogeneity.

### Interactions and heterogeneous treatment effects

In the real world, we often encounter heterogeneous treatment effects, meaning that the causal effect of one variable on an outcome is not necessarily the same in different groups of individuals. For example, we might be interested in assessing whether the effect of the hormonal treatment differs by values of `grade`. The classic approach to assess such heterogeneity is to include a multiplicative interactions in the Cox model formula:

```{r}
model <- coxph(
  Surv(dtime, death) ~ hormon * factor(grade) + ns(age, df=2), 
  data=rotterdam
)

summary(model)
```

Although there is nothing inherently wrong with this approach, the results from the Cox model alone are not easy to interpret. In this case the HR of `hormon` no longer refers to the HR of the hormonal treatment in general, but only to the HR of `hormon` *in the `grade` = 2 group*, which can be confusing. Additionally, the HRs only show conditional, not marginal, effects. Using g-computation we can instead visualize adjusted survival curves for `hormon` by `grade` and perform additional hypothesis tests.

Even though the model and the estimand changed, the syntax required to estimate the counterfactual survival probabilities stays essentially the same. The only change we have to perform is that we have to add `grade` to the `datagrid()` call and to the `by` argument. To make the output a little more comprehensible we focus three points in time: $t\in\{1000, 4000, 7000\}$.

```{r}
p <- avg_predictions(model,
  type = "survival",
  by = c("dtime", "hormon", "grade"),
  vcov = "rsample",
  newdata = datagrid(
    hormon = unique,
    grade = unique,
    dtime = c(1000, 4000, 7000),
    grid_type = "counterfactual"
  )
)
p
```

The output now contains twelve rows, one for each combination of `dtime`, `hormon` and `grade`. For example, the first row shows the estimated average survival probability at $t=1000$ would be `r sprintf("%.0f%%", p$estimate[1] * 100)`, in a counterfactual world where every patient has `hormon=0` and `grade = 2`.

Now, imagine we want to test a complex hypothesis such as:

> What is the difference in expected (counterfactual) survival probabilities between two specific types of individuals?

a. Row 1: `hormon=0` and `grade=2` and `dtime=1000`
b. Row 7: `hormon=0` and `grade=2` and `dtime=4000`

We could first compute the estimated differences manually as:

```{r}
p[c(1, 7),]

diff(p$estimate[c(1, 7)])
```

Alternatively, we could use add the `hypothesis` argument to the previous call, which gives us the same point estimate plus confidence intervals.

```{r}
avg_predictions(model,
  hypothesis = "b1 - b7 = 0",
  type = "survival",
  by = c("dtime", "hormon", "grade"),
  vcov = "rsample",
  newdata = datagrid(
    hormon = unique,
    grade = unique,
    dtime = c(1000, 4000, 7000),
    grid_type = "counterfactual"
  )
)
```

````{comment}
::: {.callout-note}
If we were instead interested in the **effect modification** of the effect of `hormon` by `grade`, we could use almost the same code. We would only have to remove the `grade` specification from the `datagrid()` call (see @VanderWeele2009 for an explanation of the distinction).
:::
````

We can of course plot counterfactual predictions for different subgroups. Here, we consider 25 equally spaced points in time, with 4 different combinations of `hormon` and `grade`.

```{r}
#| warnings: false
p <- predictions(model,
  type = "survival",
  by = c("dtime", "hormon", "grade"),
  vcov = "rsample",
  newdata = datagrid(
    hormon = unique,
    grade = unique,
    dtime = seq(36, 7043, length.out = 25),
    grid_type = "counterfactual"
  )
)

with(p, {
    tinyplot(
        estimate ~ dtime | hormon + grade,
        ymin = conf.low,
        ymax = conf.high,
        type = "ribbon",
        ylab = "Adjusted Survival Probability"
    )
})
```

Instead of one adjusted survival curve per `hormon` group, we now get four adjusted survival curves in total. In the resulting graph we get some indication of an interaction effect between `grade` and `hormon`. For `grade = 2` the curves of `hormon = 0` and `hormon = 1` are very similar, whereas they do seem to be somewhat further apart for `grade = 3`. It is, however, difficult to make a solid judgement here because the confidence intervals of all groups are quite wide and mostly overlapping.

## Counterfactual comparisons

So far, we have focused on the average counterfactual predictions (on the survival scale) for different combinations of predictors. The `marginaleffects` package also allows us to take a more explicitly "interventional" approach, focused on the estimation of treatment effects. Indeed, [the Counterfactual Comparisons chapter](https://marginaleffects.com/chapters/comparisons.html) explains how to use the `comparisons()` and `avg_comparisons()` functions to answer a series of key questions.

To start, we ask: What is the estimated average treatment effect of `hormon` on the survival probability of people in our sample, at time $t=5000$.

```{r}
cmp <- avg_comparisons(model,
    variables = "hormon",
    type = "survival",
    vcov = "rsample",
    newdata = datagrid(
        dtime = 5000,
        grid_type = "counterfactual"
    )
)
cmp
```

On average, moving from `hormon = 0` to `hormon = 1` is associated with a `r round(cmp$estimate[1] * 100)` percentage point change in the expected survival probability at $t=5000$.

We can express the same result as a ratio of survival probabilities by using the `comparison` argument.

```{r}
avg_comparisons(model,
    comparison = "ratio",
    variables = "hormon",
    type = "survival",
    vcov = "rsample",
    newdata = datagrid(
        dtime = 5000,
        grid_type = "counterfactual"
    )
)
```

### Heterogeneous treatment effects

Now, we ask whether the estimated effect of `hormon` differs between `grade` groups. To do this, we add the `by` argument.

```{r}
cmp <- avg_comparisons(model,
    variables = "hormon",
    type = "survival",
    by = "grade",
    vcov = "rsample",
    newdata = datagrid(
        dtime = 5000,
        grid_type = "counterfactual"
    )
)
cmp
```

For individuals in `grade = 2`, moving from `hormon = 0` to `hormon = 1` is associated with a `r round(cmp$estimate[1] * 100)` percentage point change in the expected survival probability at $t=5000$. For individuals in `grade = 3`, the corresponding change is `r round(cmp$estimate[2] * 100)` percentage points.

At first glance, these two estimates seem different. Are they really? To test this formally, we add the `hypothesis` argument to the previous call.

```{r}
cmp <- avg_comparisons(model,
    hypothesis = difference ~ reference,
    variables = "hormon",
    type = "survival",
    by = "grade",
    vcov = "rsample",
    newdata = datagrid(
        dtime = 5000,
        grid_type = "counterfactual"
    )
)
cmp
```

The difference in estimated survival probabilities between `hormon = 0` and `hormon = 1` is different for different values of `grade`. However, the confidence interval on the difference (in differences) covers zero. This means that we *cannot* reject the null hypothesis that the effect of `hormon` on survival is identical for people with `grade=2` and those with `grade=3`.

As explained elsewhere [in the package documentation](https://marginaleffects.com/chapters/hypothesis.html), the `hypothesis` argument allows very flexible input types. For example, we could identify estimates by row number and specify hypothesis tests like `hypothesis = "b1 - b2 = 0"` or `hypothesis = "b1 / b2 = 1"`. Alternatively, we could supply custom functions or formulas to test even more complicated hypotheses about heterogeneity.

# Uncertainty

Before concluding, it is useful to briefly discuss the challenge of estimating uncertainty around marginal estimates in the types of models considered here. One important challenge arises because the Cox proportional‐hazards model the log‑likelihood naturally factorizes into a *partial* component that involves only the regression coefficients and a *baseline* component that involves the cumulative hazard.  

$$
h(t\mid x)=h_{0}(t)\,\exp\{x^{\top}\beta\},
$$

with baseline hazard $h_{0}(t)$ and coefficients $\beta$.

The key innovation that underpins this model is that we can that we do not need to optimize the likelihood of the full model. Instead, if we are willing to accept some assumptions (i.e., proportional hazard), we can proceed with the partial likelihood.

Once we fit the model in this way, we can compute predictions on the linear-predictor scale easily.

$$
\hat\eta = x^{\top}\hat\beta,
\qquad
\widehat{\text{Var}}(\hat\eta)=x^{\top}\widehat{\text{Var}}(\hat\beta)\,x,
$$

The variance of those predictions can be obtained via the delta method, through the estimated variance–covariance matrix of the coefficients alone. Unfortunately, other quantities of interest are functions of more than just the estimated coefficients. For functions that depend on the baseline hazard (e.g., survival or cumulative hazard curves) a second source of variability enters through the estimation of $H_{0}(t)$. 

````{comment}
For any subject with covariates $x$, the marginal survivor estimate is

$$
\hat S(t\mid x)=\exp\!\bigl[-\hat H_{0}(t)\,\exp\{x^{\top}\hat\beta\}\bigr],
$$

Its variance decomposes as

$$
\text{Var}[\hat S]=
\underbrace{\Bigl(\partial S/\partial\beta\Bigr)^{\!\top}
\widehat{\text{Var}}(\hat\beta)\,
\Bigl(\partial S/\partial\beta\Bigr)}_{\text{coefficient part}}
\;+\;
\underbrace{\Bigl(\partial S/\partial H_{0}\Bigr)^{\!\top}
\widehat{\text{Var}}\bigl[\hat H_{0}(t)\bigr]\,
\Bigl(\partial S/\partial H_{0}\Bigr)}_{\text{baseline part}}
\;+\;2\,\text{Cov}\bigl(\hat\beta,\hat H_{0}\bigr), 
$$

with a baseline term that can be evaluated with different strategies.
````

`marginaleffects` applies the multivariate delta method to propagate uncertainty from $\hat\beta$ to any user‑defined function $g(\hat\beta)$. For Cox models the current implementation ignores the extra variability in $\hat H_{0}(t)$.  When effect estimates depend on the baseline this omission produces anti‑conservative confidence intervals---often by a non‑trivial margin.

A simple remedy is the non‑parametric bootstrap: resample subjects (or clusters) with replacement, refit the Cox model in each replicate, and re‑calculate the target function.  Because every replicate re‑estimates both $\beta$ *and* $H_{0}(t)$, the resulting empirical distribution captures their joint variability without analytic derivations.  While computationally heavier, this method usually restores nominal coverage and works seamlessly with `marginaleffects` via the `vcov = "rsample"` argument or the `inferences()` function.

# Conclusion

This vignette has shown that the `marginaleffects` can be used to easily compute very useful marginal estimands, that greatly facilitate the interpretation of survival models. For users who are already familiar with the `marginaleffects` package this is particularly easy, because the syntax required for survival analysis is exactly the same syntax that can be used for non-survival models. Additionally, the included capabilities to test complex hypotheses dealing with interaction and effect modification is, to the best of our knowledge, not implemented elsewhere.

Note, however, that although g-computation is one of the most popular and most efficient methods to estimate the causal estimands described here, it is not the only method and may not always be the best choice [@Denz2023]. Due to its' sole reliance on a time-to-event model, the resulting estimates are very susceptible to model misspecification. Other methods, such as augmented inverse probability of treatment weighting [@Wang2018] or targeted maximum likelihood estimation [@Cai2020], additionally use a model for the treatment assignment process and are doubly-robust in the sense that only one of the models needs to be correctly specified to obtain unbiased estimates. Additionally, we recommend that users keep in mind that the causal interpretation of the produced estimates of any method are only valid given the causal identifiability assumptions, which do not always hold in practice.

## Thanks

The `marginaleffects` authors gratefully acknowledge Robin Denz for writing the first version of this vignette.

## References
