---
output: github_document
title: "`meffects`: Marginal effects using `R`, automatic differentiation, and the delta method"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

`meffects` stands for "Marginal Effects."

`meffects` also stands for "Model Effects."

This package is still experimental. *Use with caution!*

# What is this?

This package is (essentially) a clone of the `margins` package. It allows users to easily and quickly compute "marginal effects" for a wide variety of models. A "marginal effect" is a measure of the association between a change in the regressors and a change in the response variable. More formally, [the `margins` vignette](https://cran.r-project.org/web/packages/margins/index.html) defines "marginal effects" as follows:

> "Marginal effects are partial derivatives of the regression equation with respect to each variable in the model for each unit in the data; average marginal effects are simply the mean of these unit-specific partial derivatives over some sample. In ordinary least squares regression with no interactions or higher-order term, the estimated slope coefficients are marginal effects. In other cases and for generalized linear models, the coefficients are not marginal effects at least not on the scale of the response variable. margins therefore provides ways of calculating the marginal effects of variables to make these models more interpretable."

# Supported models

```{r, echo=FALSE, warning=FALSE}
library(tibble)
library(kableExtra)

tab <- 
'"Model",           "Effect", "Variance", "Stata",      "margins"
"stats::lm",        "✓"      , "✓"        , "✓"         , "✓"
"stats::glm",       "✓"      , "✓"        , "✓"         , "✓"
"stats::loess",     "✓"      ,            ,             , "✓"
"AER::ivreg",       "✓"      , "✓"        , "✓"         , "✓"
"betareg::betareg", "✓"      , "✓"        , "✓"         , "✓"
"fixest::feols",    "✓"      , "✓"        ,             ,
"fixest::feglm",    "✓"      , "✓"        ,             ,
"ivreg::ivreg",     "✓"      , "✓"        , "✓"         , "✓"
"lme4::lmer",       "✓"      , "✓"        ,             , "dydx only"
"lme4::glmer",      "✓"      , "✓"        ,             , "dydx only"
"MASS::polr",       "✓"      ,            , "✓"         ,
"ordinal::clm",     "✓"      ,            ,             , "✓"
"survey::svyglm",   "✓"      , "✓"        ,             , "✓"
"survey::svyglm",   "✓"      , "✓"        ,             , "✓"'
tab <- read.csv(text = tab)
colnames(tab) <- c("Model", "Support: Effect", "Support: Std.Errors", "Validity: Stata", "Validity: margins")
kbl(tab, format = "pipe") %>% kable_styling() 
#%>% add_header_above(c(" " = 1, "Support" = 2, "Validity" = 2))
```

# Installation

You can install the released version of meffects from Github:

```{r, eval=FALSE}
remotes::install_github("vincentarelbundock/meffects")
```

# Examples

## Logit regression with a multiplicative interaction term

Load the library, simulate data, and estimate a logistic regression model with interaction terms:

```{r}
library(meffects)

N <- 1000
dat <- data.frame(
    x2 = rnorm(N),
    x1 = rnorm(N),
    x3 = rnorm(N),
    x4 = rnorm(N),
    e = rnorm(N))
dat$y <- rbinom(N, 1, plogis(
    dat$x1 + dat$x2 + dat$x3 + dat$x4 + dat$x3 * dat$x4))

mod <- glm(y ~ x1 + x2 + x3 * x4, data = dat, family = binomial)

coef(mod)
```

Compute unit-level marginal effects and standard errors:

```{r}
mfx <- meffects(mod)
head(mfx)
```

Notice that the results are presented in "tidy" format: each row of the original dataset gets a unique `rowid` value, each unit-level marginal effect appears on a distinct row, and metadata appears neatly in separate columns. This makes it easy to operate on the results programmatically.

We can obtain similar (but arguably messier) results with the `margins` package:

```{r, eval=FALSE}
library(margins)

mar <- margins(mod, unit_ses = TRUE)

head(data.frame(mar), 2)
```

## Conditional marginal effects with `ggplot2`

```{r, echo=FALSE}
library(ggplot2)
theme_set(theme_minimal())
```

We can use the `newdata` argument to do a "counterfactual" analysis. Here, we create a new dataset with some factor and logical variables. Then, we use the `at` argument to specify counterfactual values that we want to inspect. When a variable is not specified in `at`, its median or mode is automatically selected:

```{r, eval=FALSE}
tmp <- mtcars
tmp$am <- as.logical(tmp$am)
tmp$gear <- as.factor(tmp$gear)
mod <- lm(mpg ~ hp * wt + am + gear, data = tmp)

meffects(mod, 
          variables = "hp",
          at = list(hp = c(100, 110),
                    gear = c(3, 4),
                    am = TRUE))
```

The nice thing about the `tidy` output format is that we can pipe the output of the `meffects` function directly to `ggplot2`:

```{r, eval=FALSE}
library(tidyverse)

meffects(mod, 
          variables = "hp",
          at = list(hp = c(100, 110),
                    gear = c(3, 4),
                    am = TRUE))


    mutate(conf.low = dydx - 1.96 * std.error,
           conf.high = dydx + 1.96 * std.error) |>
    ggplot(aes(x = x4, 
               y = dydx, 
               ymin = conf.low, 
               ymax = conf.high)) +
    geom_ribbon(alpha = .1) +
    geom_line() + 
    facet_wrap(~term)
```

# Benchmarks

Here are a couple naive benchmarks to compare the speed of computation with the `meffects` and `margins` packages. Since unit-level standard errors can be expensive to compute, we run the benchmarks with and without standard errors.

## Marginal effects and standard errors (unit-level)

In this naive benchmark, computing marginal effects with their unit-level standard errors is over 300x faster.

```{r, eval=FALSE}
b1 = bench::mark(
    margins(mod, unit_ses = TRUE),
    meffects(mod, variance = vcov(mod)),
    check = FALSE,
    max_iterations = 3)
b1
```

## Marginal effects only

In this naive benchmark, computing marginal effects *without* their unit-level standard errors is over 40% faster.

```{r, eval=FALSE}
b2 = bench::mark(
    margins(mod, unit_ses = FALSE),
    meffects(mod, variance = NULL),
    check = FALSE,
    max_iterations = 3)
b2
```
