# Elasticity

```{r, include = FALSE}
## this vignette is in .Rbuildignore because lme4 is not available on old CRAN
## test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```


::: {.callout-warning}
Slopes and elasticities can only be calculated for continuous numeric variables. The `slopes()` functions will automatically revert to `comparisons()` for binary or categorical variables.
:::

In some contexts, it is useful to interpret the results of a regression model in terms of elasticity or semi-elasticity. One strategy to achieve that is to estimate a log-log or a semilog model, where the left and/or right-hand side variables are logged. Another approach is to note that $\frac{\partial ln(x)}{\partial x}=\frac{1}{x}$, and to post-process the marginal effects to transform them into elasticities or semi-elasticities.

For example, say we estimate a linear model of this form:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \varepsilon$$

Let $\hat{y}$ be the adjusted prediction made by the model for some combination of covariates $x_1$ and $x_2$. The slope with respect to $x_1$ (or "marginal effect") is:

$$\frac{\partial \hat{y}}{\partial x_1}$$

We can estimate the slope ("dydx"), semi-elasticities ("dyex" and "eydx"), or elasticity ("eyex") with respect to $x_1$ as follows:

$$\eta_1=\frac{\partial \hat{y}}{\partial x_1}$$
$$\eta_2=\frac{\partial \hat{y}}{\partial x_1}\cdot x_1$$
$$\eta_3=\frac{\partial \hat{y}}{\partial x_1}\cdot \frac{1}{\hat{y}}$$ 
$$\eta_4=\frac{\partial \hat{y}}{\partial x_1}\cdot \frac{x_1}{\hat{y}}$$

with interpretations roughly as follows:

1. "dydx": Increasing the $x_1$ variable by 1 unit is associated with a change of $\eta_1$ units in $y$.
2. "dyex": Increasing the $x_1$ variable by 100% of its baseline value is associated with a change of $\eta_2$ units in $y$.
3. "eydx": Increasing the $x_1$ variable by 1 unit is associated with a change of $\eta_3$% of $y$'s baseline value.
4. "eyex": Increasing $x_1$ by 1% of its baseline value is associated with a change of $\eta_4$% in the baseline value of $y$.


With the `marginaleffects` package, these quantities are easy to compute with the `slopes()` function and the `slope` argument. 

::: {.panel-tabset group="language"}
### R
```{r}
library(marginaleffects)
dat <- transform(mtcars, y = mpg, x = wt)
mod <- lm(y ~ x, data = dat)

pct_change_x <- mean(dat$x) / 100
p0 <- predict(mod)
p1 <- predict(mod, newdata = transform(dat, x = x + pct_change_x))
mean(p1 - p0)

s <- avg_slopes(mod, slope = "dyex")
s$estimate
```
### Python
```{python}
import numpy as np
import statsmodels.formula.api as smf
from marginaleffects import avg_slopes
from statsmodels.datasets import get_rdataset

mtcars = get_rdataset("mtcars").data

dat = mtcars.assign(
    y=mtcars["mpg"],
    x=mtcars["wt"]
)

mod = smf.ols("y ~ x", data=dat).fit()

pct_change_x = dat["x"].mean() / 100

p0 = mod.predict(dat)
p1 = mod.predict(dat.assign(x=dat["x"] + pct_change_x))

print(np.mean(p1 - p0))

s_dyex = avg_slopes(mod, slope="dyex")
print(float(s_dyex["estimate"][0]))
```
:::


Increasing the `x` variable by 100% of its baseline value is associated with a change of $\eta_1=`{r} s$estimate`$ units in $y$. Notice that the `slopes()` estimates are scaled differently than in the manual computation.

::: {.panel-tabset group="language"}
### R
```{r}
yhat <- predict(mod)
p0 <- predict(mod, newdata = dat)
p1 <- predict(mod, newdata = transform(dat, x = x + 1))
mean((p1 - p0) / yhat)

s <- avg_slopes(mod, slope = "eydx")
s$estimate
```
### Python
```{python}
yhat = mod.predict(dat)

p0 = mod.predict(dat)
p1 = mod.predict(dat.assign(x=dat["x"] + 1))

print(np.mean((p1 - p0) / yhat))

s_eydx = avg_slopes(mod, slope="eydx")
print(float(s_eydx["estimate"][0]))
```
:::

Increasing the `x` variable by 1 unit is associated with a change of $\eta_2=`{r} s$estimate`\%$ of $y$'s baseline value.

::: {.panel-tabset group="language"}
### R
```{r}
pct_change_x <- 1 / dat$x
pct_change_y <- (p1 - p0) / yhat
mean(pct_change_y / pct_change_x)

s <- avg_slopes(mod, slope = "eyex")
s$estimate
```
### Python
```{python}
pct_change_x = 1 / dat["x"]
pct_change_y = (p1 - p0) / yhat

print(np.mean(pct_change_y / pct_change_x))

s_eyex = avg_slopes(mod, slope="eyex")
print(float(s_eyex["estimate"][0]))
```
:::


Increasing `x` by 1% of its baseline value is associated with a change of $\eta_3=`{r} s$estimate`\%$ in the baseline value of $y$.
