
# Extensions


```{r, include = FALSE}
options(width = 1000)
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .4,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

This vignette shows how to add support for new models and add new functionality for supported models.

## Support a new model type

It is very easy to add support for new models in `marginaleffects`.  All we need is to set a global option and define 4 very simple functions.

If you add support for a class of models produced by a CRAN package, please consider submitting your code for inclusion in the package: https://github.com/vincentarelbundock/marginaleffects

If you add support for a class of models produced by a package hosted *elsewhere* than CRAN, you can submit it for inclusion in the *unsupported* user-submitted library of extensions: Currently

* [`countreg` package](https://github.com/vincentarelbundock/marginaleffects/blob/main/sandbox/methods_countreg.R). Thanks to Olivier Beaumais.
* [`censreg` package](https://github.com/vincentarelbundock/marginaleffects/blob/main/sandbox/methods_censReg.R). Thanks to Oleg Komashko.

The rest of this section will illustrate how to add support a model: first a frequentist model, and then a Bayesian model.

### A Frequentist Model

To begin, we define a function which fits a model. Normally, this function will be supplied by a modeling package published on CRAN. Here, we create a function called `lm_manual()`, which estimates a linear regression model using simple linear algebra operators, and returns an object of class `lm_manual`:

```{r}
lm_manual <- function(f, data, ...) {
  # design matrix
  X <- model.matrix(f, data = data)
  # response matrix
  Y <- data[[as.character(f[2])]]
  # coefficients
  b <- solve(crossprod(X)) %*% crossprod(X, Y)
  Yhat <- X %*% b
  # variance-covariance matrix
  e <- Y - Yhat
  df <- nrow(X) - ncol(X)
  s2 <- sum(e^2) / df
  V <- s2 * solve(crossprod(X))
  # model object
  out <- list(
    d = data,
    f = f,
    X = X,
    Y = Y,
    V = V,
    b = b
  )
  # class name: lm_manual
  class(out) <- c("lm_manual", "list")
  return(out)
}
```

Important: The custom fit function must assign a new class name to the object it returns. In the example above, the model is assigned to be of class `lm_manual` (see the penultimate line of code in the function).

Our new function replicates the results of `lm()`:

```{r}
model <- lm_manual(mpg ~ hp + drat, data = mtcars)
model$b

model_lm <- lm(mpg ~ hp + drat, data = mtcars)
coef(model_lm)
```

To extend support in `marginaleffects`, the first step is to tell the package that our new class is supported. We do this by defining a global option:

```{r}
library(marginaleffects)

options("marginaleffects_model_classes" = "lm_manual")
```

Then, we define 4 methods:

1. `get_coef()`
    - Mandatory arguments: `model`, `...`
    - Returns: named vector of parameters (coefficients).
1. `set_coef()`
    - Mandatory arguments: `model`, `coefs` (named vector of coefficients), `...`
    - Returns: A new model object in which the original coefficients were replaced by the new vector.
    - [Example](https://github.com/vincentarelbundock/marginaleffects/blob/main/R/set_coef.R)
1. `get_vcov()`
    - Mandatory arguments: `model`, `...`.
    - Optional arguments: `vcov`
    - Returns: A named square variance-covariance matrix.
1. `get_predict()`
    - Mandatory arguments: `model`, `newdata` (data frame), `...`
    - Option arguments: `type` and other model-specific arguments.
    - Returns: A data frame with two columns: a unique `rowid` and a column of `estimate` values.

Note that each of these methods will be named with the suffix `.lm_manual` to indicate that they should be used whenever `marginaleffects` needs to process an object of class `lm_manual`.

```{r}
get_coef.lm_manual <- function(model, ...) {
  b <- model$b
  b <- setNames(as.vector(b), row.names(b))
  return(b)
}

set_coef.lm_manual <- function(model, coefs, ...) {
  out <- model
  out$b <- coefs
  return(out)
}

get_vcov.lm_manual <- function(model, ...) {
  return(model$V)
}

get_predict.lm_manual <- function(model, newdata, ...) {
  newX <- model.matrix(model$f, data = newdata)
  Yhat <- newX %*% model$b
  out <- data.frame(
    rowid = seq_len(nrow(Yhat)),
    estimate = as.vector(Yhat)
  )
  return(out)
}
```

The methods we just defined work as expected:

```{r}
get_coef(model)

get_vcov(model)

get_predict(model, newdata = head(mtcars))
```

Now we can use the `avg_slopes` function:

```{r}
avg_slopes(model, newdata = mtcars, variables = c("hp", "drat"))

predictions(model, newdata = mtcars) |> head()
```

Note that, for custom models, we typically have to supply values for the `newdata` and `variables` arguments explicitly.

### A Bayesian Model

Adding support for a Bayesian model is even easier!

Bayesian models are typically estimated using MCMC methods, and so all of the inferential statistics are derived directly from the samples of the posterior distribution. Therefore, the only thing we need to do is to define a `get_predict()` method that returns posterior samples of predictions.

Let's illustrate this with a simple example using the `rjags` package. First, let's estimate a simple logistic regression model:

```{r}
#| results: hide
library(rjags)

dat <- get_dataset("Titanic", "Stat2Data")
dat <- dat[!is.na(dat$Age), ]

model_string <- "
model {
  # Likelihood
  for (i in 1:N) {
    y[i] ~ dbern(p[i])
    logit(p[i]) <- b0 + b1 * x1[i] + b2 * x2[i]
  }

  # Priors
  b0 ~ dnorm(0, 0.001)
  b1 ~ dnorm(0, 0.001)
  b2 ~ dnorm(0, 0.001)
}"

data_list <- list(
  y = dat$Survived,
  x1 = dat$Age,
  x2 = dat$SexCode,
  N = nrow(dat)
)

model <- jags.model(
  textConnection(model_string),
  data = data_list,
  n.chains = 3
)

update(model, 1000) # burn-in

samples <- coda.samples(
  model,
  c("b0", "b1", "b2"),
  n.iter = 1000
)
```

The `samples` object contains posterior samples of the parameters. It's not a "model" in the traditional sense, but it contains all the information we need to compute predictions for new data.

For example, the below code reutrns 3000 MCMC samples of the posterior distribution of the predicted probability of survival for a 36 year old male:

```{r}
#| eval: false
Age <- 36
SexCode <- 0

lodds <- as.matrix(samples) %*% c(1, Age, SexCode)
probs <- exp(lodds) / (1 + exp(lodds))
```

We can generalize this code to compute predictions for any new data frame.

```{r}
class(samples) <- c("jags_samples", class(samples))

get_predict.jags_samples <- function(
  model,
  newdata,
  type = c("response", "link"),
  ...
) {
  type <- match.arg(type)

  # Compute predictions for each row of newdata and each posterior draw
  X <- cbind(1, newdata$Age, newdata$SexCode)
  est <- as.matrix(model) %*% t(X) # log-odds predictions
  if (type == "response") {
    est <- exp(est) / (1 + exp(est)) # probability predictions
  }

  # Make an empty data frame with the right number of rows
  out <- data.frame(
    rowid = seq_len(nrow(newdata)),
    estimate = 0
  )

  # Append the posterior draws as an attribute of the output data frame
  attr(out, "posterior_draws") <- t(est)

  return(out)
}

# Not strictly necessary, but defining this function silences a warning about
# missing variance-covariance matrix.
get_vcov.jags_samples <- function(model, vcov = NULL, ...) {
  return(NULL)
}
```

And... that's it!

Let's see how it works within `marginaleffects`:

```{r}
# tell the package that we are adding support for a new class of models:
options("marginaleffects_model_classes" = "jags_samples")

avg_predictions(samples, variables = "SexCode", newdata = dat)

avg_slopes(samples, variables = "Age", by = "SexCode", newdata = dat)

# etc...
```

See what you can do with Bayesian models in `marginaleffects` in the [Bayes chapter](brms.html).

## Merge your extension into the main package

If you wrote a working extension, please consider contributing back to the community by submitting your code for inclusion in the main package. To do this, all you need to do is follow the steps above, and then:

1. Add the package to the permanent list of supported models, to avoid calling `options()` in every session: `R/sanitize_model.R`
2. Add the package to the vignette of supported models by editing: `website/data/raw/supported_models.csv`
3. Add a bullet point to the news file: `NEWS.md`

## Modify or extend supported models

Let's say you want to estimate a model using the `mclogit::mblogit` function.  That package is already supported by `marginaleffects`, but you want to use a `type` (scale) of predictions that is not currently supported: a "centered link scale."

To achieve this, we would need to override the `get_predict.mblogit()` method. However, it can be unsafe to reassign methods supplied by a package that we loaded with `library`. To be safe, we assign a new model class to our object ("customclass") which will inherit from `mblogit`. Then, we define a `get_predict.customclass` method to make our new kinds of predictions.

Load libraries, estimate a model:

```{r}
library(mclogit)
library(data.table)

model <- mblogit(
  factor(gear) ~ am + mpg,
  data = mtcars,
  trace = FALSE
)
```

Tell `marginaleffects` that we are adding support for a new class model models, and assign a new inherited class name to a duplicate of the model object:

```{r}
options("marginaleffects_model_classes" = "customclass")

model_custom <- model

class(model_custom) <- c("customclass", class(model))
```

Define a new `get_predict.customclass` method. We use the default `predict()` function to obtain predictions. Since this is a multinomial model, `predict()` returns a matrix of predictions with one column per level of the response variable.

Our new `get_predict.customclass` method takes this matrix of predictions, modifies it, and reshapes it to return a data frame with three columns: `rowid`, `group`, and `estimate`:

```{r}
get_predict.customclass <- function(model, newdata, ...) {
  out <- predict(model, newdata = newdata, type = "link")
  out <- cbind(0, out)
  colnames(out)[1] <- dimnames(model$D)[[1]][[1]]
  out <- out - rowMeans(out)
  out <- as.data.frame(out)
  out$rowid <- seq_len(nrow(out))
  out <- data.table(out)
  out <- melt(
    out,
    id.vars = "rowid",
    value.name = "estimate",
    variable.name = "group"
  )
}
```

Finally, we can call any `slopes` function and obtain results. Notice that our object of class `customclass` now produces different results than the default `mblogit` object:

```{r}
avg_predictions(model)

avg_predictions(model_custom)
```
