
# FAQ 


```{r, include = FALSE}
options(width = 1000)
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
url <- "https://raw.githubusercontent.com/vincentarelbundock/marginaleffects/main/website/data/raw/supported_models.csv"
dat <- read.csv(url)
n_support <- nrow(dat)
```

## Questions & Answers

* [Error: Matrix columns are not supported](https://stackoverflow.com/questions/78568706/marginaleffects-error-matrix-columns-are-not-supported)
* [Wiggly (non-smooth) confidence intervals in plots.](https://stackoverflow.com/questions/77556969/wiggly-confidence-interval-with-marginaleffectsplot-slopes/77620185#77620185)
* [`plot_predictions()` over a range of unobserved values](https://stackoverflow.com/questions/72723687/plot-cap-response-curve-for-counterfactual-data)
* [Plot the marginal effects from a `plm` package model](https://stackoverflow.com/questions/73126378/marginal-effects-plot-of-plm/73127507#73127507)
* [Models with demeaned, polynomials, or transformed variables](https://stackoverflow.com/questions/73303108/marginal-effects-for-de-meaned-polynomials-in-mixed-models/73305398#73305398)
* [`nlme::lme` problem with character predictors](https://stackoverflow.com/questions/77516330/marginaleffects-package-refuses-service-with-nlme/77517278#77517278)
* [Controlling facet rows and columns in `plot_*()` functions.](https://github.com/vincentarelbundock/marginaleffects/issues/1107)
* [Bayesian diagnostics (Rhat, etc.) with `brms`](https://stackoverflow.com/questions/78479179/diagnostic-of-marginal-effects-from-a-brms-model)
* [Pooling predictions with multiple imputation](https://stackoverflow.com/questions/77686706/how-to-pool-average-predictions-from-a-multinomial-regression-per-category-in-r/77688581#77688581)


## Calling `marginaleffects` in functions, loops, environments, or after re-assigning variables

Functions from the `marginaleffects` package can sometimes fail when they are called inside a function, loop, or other environments. To see why, it is important to know that `marginaleffects` often needs to operate on the original data that was used to fit the model. To extract this original data, we use the `get_data()` function from the `insight` package.

In most cases, `get_data()` can extract the data which is stored inside the model object created by the modeling package. However, some modeling packages do *not* save the original data in the model object (in order to save memory). In those cases, `get_data()` will parse the call to find the name of the data object, and will search for that data object in the global environment. When users fit models in a different environment (e.g., function calls), `get_data()` may not be able to retrieve the original data.

A related problem can arise if users fit a model, but then assign a new value to the variable that used to store the dataset.

Recommendations:

1. Supply your dataset explicitly to the `newdata` argument of `slopes` functions.
2. Avoid assigning a new value to a variable that you use to store a dataset for model fitting.

Another alternative is to re-assign relevant variables explicitly inside the function, to ensure that `marginaleffects` gets access to them. For instance,

```r
get_int_eff = function(formula, data, boot = FALSE){
  assign("formula", formula, envir = parent.frame())
  mod = glm(formula = formula, data = data)
  int_eff = avg_comparisons(
    model = mod,
    variables = "am",
    type = "response")
  if (boot){
    int_eff = inferences(int_eff, method = "boot", R = 100, conf_type = "perc")
  }
  return(int_eff)
}
```

## Why you should not modify the original data after fitting

`marginaleffects` frequently needs access to the original dataset even when you supply `newdata`. This is because most model objects do not store the full data, or they store a `model.frame` that cannot always be inverted back to the original variables (some in-formula transformations are non-invertible). As a result, `insight::get_data()` often searches the environment for a data frame with the same name used in the model call.

If you overwrite or mutate that original object after fitting, `marginaleffects` may:

- Throw confusing errors about missing variables or factor levels.
- Use altered values silently, leading to incorrect estimates.

To avoid this, keep the original data object unchanged between model fitting and `marginaleffects` calls. If you must change it, use a new object name, refit the model, or pass an explicit dataset via `newdata`/`modeldata` when possible.


## Equivalence between `avg_comparisons()` and `avg_predictions()`

Say we estimate a simple logit regression and want to compute the average log-odds ratio associated with a change from 0 to 1 on the `vs` variable. We can proceed as follows using the `avg_comparisons()` function:

```{r}
library(marginaleffects)
mod <- glm(am ~ vs + scale(drat), family = binomial, mtcars)

avg_comparisons(mod,
    variables = "vs",
    comparison = "lnoravg",
    type = "response")
```

We can specify the function to compare "hi" predictions to "lo" predictions manually and get the same results:


```{r}
fun <- function(hi, lo) log((mean(hi) / (1 - mean(hi))) / (mean(lo) / (1 - mean(lo))))

comparisons(mod,
    variables = "vs",
    comparison = fun,
    type = "response")
```

The same results can be obtained using the `predictions()` function. First, we replicate the entire dataset, and substitute the values of `vs`. Then, we make predictions. Finally, we compute the log odds ratios:

```{r}
dat0 <- transform(mtcars, vs = 0)
dat1 <- transform(mtcars, vs = 1)

p0 <- predictions(mod, newdata = dat0, type = "response")
p1 <- predictions(mod, newdata = dat1, type = "response")

fun(p1$estimate, p0$estimate)
```

Notice that the following command gives us a different result. This is because instead of duplicating the full dataset, and computing the "hi" and "lo" as the mean predictions over the full data, we only compute the averages on the subsets of rows where `vs` is 0 or 1, respectively:

```{r}
p <- avg_predictions(mod, by = "vs", type = "response")
fun(p$estimate[2], p$estimate[1])
```

Which is equivalent to:

```{r}
dat0 <- subset(mtcars, vs == 0)
dat1 <- subset(mtcars, vs == 1)

p0 <- predictions(mod, newdata = dat0, type = "response")
p1 <- predictions(mod, newdata = dat1, type = "response")

fun(p1$estimate, p0$estimate)
```

## Dates

Date predictors are not supported by default in `marginaleffects`. There are two main reasons for that. First, many quantities of interest are only formally defined for numeric variables (ex: derivatives). Second, an object like "2024-11-04" does not have a natural arithmetic. What does it mean to take a sum like: $\pi$ + "2024-11-04"? In that spirit, `marginaleffects` requires that the user transforms the date variable into a numeric variable before fitting the model. This makes operations more explicit.

In this example, we fit a model with a date and another with a integer version of the date. We call `predict()` to illustrate the fact that the two models yield exactly identical predictions:

```{r}
#| error: true
library(lubridate)
library(marginaleffects)
set.seed(123)

N <- 100
Date <- ymd("2024-01-01") + days(sample(0:30, N, replace = TRUE))
Days <- as.integer(Date)
Response <- pi * as.integer(Date) + rnorm(N)

mod <- lm(Response ~ Date)
predict(mod) |> head()
avg_slopes(mod)

mod <- lm(Response ~ Days)
predict(mod) |> head()
avg_slopes(mod)
```

## More plot customization

As [described in the Plot vignette,](plots.html) there are currently three ways to customize plots in `marginaleffects`:

1. Using the arguments provided by `plot_*()` functions in the `marginaleffects` package.
2. Modifying the plot objects using `ggplot2` or an extension package like `ggExtra`.
3. Setting `draw=FALSE` to extract the plotting data, and then feeding that data to a standalone plotting software like `ggplot2`, `plot`, or `lattice`.

Options 2 and 3 are extremely flexible, and make it trivial to reproduce the default plots or to draw highly customized visualizations.

Of course, some users might still prefer to use function arguments directly available in `marginaleffects::plot_*()`, and the unavailability of some arguments can be frustrating. Nevertheless, a decision was made to keep the number of arguments in `plot_*()` functions quite low. This is because `ggplot2` functions like `geom_*()`, `facet_*()`, and `aes()` each accept a very large number of options, and because users' needs are incredibly varied. This makes it difficult to accomodate all needs and possibilities, without developing unwieldy functions with dozens of arguments.^[Note that we cannot pass visual arguments to `...`, because that mechanism is already used to pass arguments to the `predict()` methods, which is required for some model types.]

Thus, the `marginaleffects` package is designed with limited scope, focusing on *computation* while offering a few *visualization* options for quick exploration. Users who need highly customized or publication-ready figures are encouraged to use the `draw=FALSE` argument with standalone plotting software.

In an ideal world, someone else would build a richer visualization package on top of `marginaleffects`. I think this would be a very cool project, and I would be happy to collaborate, advise, and help if I can. In the meantime, feel free to keep the discussion going by opening an issue on Github. I will be happy to consider counter-arguments, suggestions, or pull requests.


## What’s the difference between `condition` and `by` in `plot_*()` functions?

The two arguments control how other variables in the model are handled when plotting predictions, comparisons, or slopes.

#### `condition`: Conditional estimates for a single (average or modal) hypothetical individual

Answers this question: 

> How would the estimated quantity change if we changed the focal predictors specified in `condition`, while holding all other covariates at their mean or modal values?

- Builds a *grid* over your focal predictors, while holding **all other variables** (including random effects in mixed models) at a **fixed reference value**—typically their mean, mode, or a user-defined value.
- Equivalent to `newdata = datagrid(...)` in `predictions()`.

Appropriate when:

- We are substantively interested in the "average person," that is, in a person or unit that is exactly average or modal on all dimensions.
- Computation is expensive and we need fast results.
- The focal predictor we want to display on the x-axis is continuous, which makes averaging estimates for each unique value of that continuous predictor difficult. (This often results in a jagged-looking plot.)

Inappropriate when:

- The hypothetical average individual does not exist, or is not particularly interesting.

#### `by`: Average (marginal) predictions, slopes. and comparisons

Answers this question: 

> What is the average prediction/comparison/slope in the subgroups defined by the focal predictors specified in `by`?

- Computes predictions for **each row in the original data**, using their actual covariate values, then **aggregates/averages** within subgroups defined by `by`.
- Equivalent to using `by` in `predictions()` or `avg_predictions()`.

Appropriate when:

- The sample is roughly representative of the target population.
- We are interested in an average prediction or an average effect, marginalized for different subgroups of the variables specified in `by`.

Inappropriate when:

- The x-axis variable is continuous. Since we average estimates for every value of the x-axis variable, we get averages computed in very small bins, with different "mix" of control variables in each bin. This can lead to jagged-looking plots, which are not very informative.

FWIW, I (Vincent) personally tend to nearly always use `by`, unless the x-axis variable is continuous and I want a clean, smooth line.


## How to pass arguments to `datagrid()` functions

There is not a direct way to pass arguments to functions in datagrid(). But it’s quite easy to achieve the result you need with a function factory:

```{r}
library(marginaleffects)
seq.range = \(n) \(x) seq(min(x), max(x), length.out = n)

grid <- datagrid(
  Petal.Width = seq.range(5), 
  Petal.Length = seq.range(10), 
  newdata = iris)

dim(grid)
```

## `inferences()` inside a wrapper

One challenge is that `inferences(method = "boot")` re-evaluates a stored call. If a wrapper passes local symbols instead of concrete values, bootstrap can fail when those symbols are not found.

A solution is to build `avg_comparisons()` with evaluated arguments using `do.call()`. This stores literal values in the call, so bootstrap re-evaluation works.

```{r}
set.seed(1)
library(marginaleffects)

dat <- data.frame(
  period = rep(c(0, 1), each = 100),
  treatment = rep(c(0, 1), times = 100),
  n = sample(10:100, size = 200, replace = TRUE)
)
dat$y <- round(runif(nrow(dat), 0.05, 0.95) * dat$n)

mod <- glm(cbind(y, n - y) ~ treatment * period, data = dat, family = "binomial")

fun <- function(model, var_trt) {
  args <- list(model = model, variables = var_trt) 
  out <- do.call(avg_comparisons, args)
  inferences(out, method = "boot", R = 10, conf_type = "perc")
}

fun(mod, "treatment")
```

