# GAM

Generalized Additive Models (GAMs) are a flexible class of statistical models that extend traditional linear regression by allowing non-linear relationships between predictors and the response variable. Instead of assuming straight-line relationships, GAMs use smooth functions called splines that can capture complex patterns in the data. 

This flexibility makes GAMs particularly useful for modeling real-world phenomena where relationships may be curved, seasonal, or otherwise non-linear. For example, the relationship between temperature and energy consumption often follows a U-shape that would be poorly captured by linear regression. 

While GAMs are powerful and flexible, their estimated parameters can be difficult to interpret directly since they involve complex spline basis functions. Fortunately, the `marginaleffects` package provides tools to understand and visualize GAM results using the same intuitive workflow as simpler models - through predictions, counterfactual comparisons, and slopes.


```{r, include = FALSE}
## this vignette is in .Rbuildignore because lme4 is not available on old CRAN
## test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

library(ggplot2)

theme_clean <- function() {
  theme_minimal() +
    theme(panel.grid.minor = element_blank(),
          strip.text = element_text(size = rel(1), hjust = 0),
          strip.background = element_blank(),
          legend.position = "bottom")
}
ggplot2::theme_set(theme_clean())
```

## Estimation

We will estimate a GAM model using the `mgcv` package and the `simdat` dataset distributed with the `itsadug` package:

```{r}
library(marginaleffects)
library(itsadug)
library(mgcv)

simdat$Subject <- as.factor(simdat$Subject)

dim(simdat)
head(simdat)
```

Fit a model with a random effect and group-time smooths:

```{r}
model <- bam(Y ~ Group + s(Time, by = Group) + s(Subject, bs = "re"),
             data = simdat)

summary(model)
```

## Predictions

Compute adjusted predictions for each observed combination of regressor in the dataset used to fit the model. This gives us a dataset with the same number of rows as the original data, but new columns with predicted values and uncertainty estimates:

```{r}
pred <- predictions(model)
dim(pred)
head(pred)
```

We can easily plot adjusted predictions for different values of a regressor using the `plot_predictions()` function:

```{r}
plot_predictions(model, condition = "Time")
```

## Slopes

Marginal effects are slopes of the prediction equation. They are an observation-level quantity. The `slopes()` function produces a dataset with the same number of rows as the original data, but with new columns for the slope and uncertainty estimates:

```{r}
mfx <- slopes(model, variables = "Time")
head(mfx)
```

We can plot marginal effects for different values of a regressor using the `plot_slopes()` function.
This next plot shows the slope of the prediction equation, that is, the slope of the previous plot, at every value of the `Time` variable.

```{r}
plot_slopes(model, variables = "Time", condition = "Time")
```

The marginal effects in this plot can be interpreted as measuring the change in `Y` that is associated with a small increase in `Time`, for different baseline values of `Time`.

## Excluding terms

The `predict()` method of the `mgcv` package allows users to "exclude" some smoothing terms, using the `exclude` argument. You can pass the same argument to any function in the `marginaleffects` package:

```{r}
predictions(model, newdata = "mean", exclude = "s(Subject)")
```

See the documentation in `?mgcv:::predict.bam` for details.

## Warning

It may not always be appropriate to compute aggregated (aka, average or marginal) for models with splines. See @Woo2017 for a technical reference and [this Cross Validated post](https://stats.stackexchange.com/questions/654784/how-to-decide-which-spline-to-use-when-conducting-g-computation-after-weighting/654791#654791) for a discussion.
