# Data grids

This notebook presents some advanced `datagrid()` use. Readers should be familiar with the content of @sec-framework_grid before diving in.

## The `by` argument

Sometimes we want to hold variables at constant values *within* groups. The `by` argument in `datagrid()` allows us to create grids where certain variables are fixed at specific values within subgroups.

For example, we might want to compare predictions at the mean of a continuous variable. In this example, the `Sepal.Width` column is held at the *global* mean in every row of the grid:

```{r}
library(marginaleffects)
mod <- lm(Petal.Width ~ Sepal.Width + Species, data = iris)

predictions(mod,
  newdata = datagrid(Species = unique, Sepal.Width = mean)
)
```

Now, we hold `Sepal.Width` at its mean value within each `Species` group. This allows us to compare the effect of species while controlling for the typical sepal width in each group:

```{r}
predictions(mod,
  newdata = datagrid(Species = unique, Sepal.Width = mean, by = "Species")
)

aggregate(Sepal.Width ~ Species, data = iris, FUN = mean)
```

## Sampling from large grids with `grid_type = "dataframe"`

When working with many categorical variables, a balanced grid can become prohibitively large. For instance, with 5 variables each taking 26 possible values, a full factorial grid would have 26^5 = 11,881,376 rows.

The `grid_type = "dataframe"` argument combined with a custom `FUN` allows us to sample uniformly from the full factorial space. The resulting grid will "approach" the true factorial grid, but have a much smaller footprint.

```{r}
# Simulate data with 5 categorical variables
set.seed(48103)
n <- 10000
dat <- data.frame(
  x1 = sample(letters, n, replace = TRUE),
  x2 = sample(letters, n, replace = TRUE), 
  x3 = sample(letters, n, replace = TRUE),
  x4 = sample(letters, n, replace = TRUE),
  x5 = sample(letters, n, replace = TRUE),
  y = rnorm(n)
)

# Fit model
mod <- lm(y ~ x1 + x2 + x3 + x4 + x5, data = dat)

# Sample 5 values uniformly from each variable's unique entries
sample_unique <- \(x, n) sample(unique(x), n, replace = TRUE)
grid_sampled <- datagrid(
  model = mod,
  grid_type = "dataframe",
  FUN = \(x) sample_unique(x, 5)
)

grid_sampled
```

Now we can make predictions on the much smaller sampled grid:

```{r}
predictions(mod, newdata = grid_sampled)
```

Note that the `grid_type="dataframe"` argument tells `datagrid()` to construct the grid by applying `FUN` to each variable individually, and then by stacking the sampled values vertically using `cbind()`. This is different from other grids like `balanced` or `counterfactual`, which take the cartesian product of supplied values, with `expand.grid()`.
