# Multinomial Logit

This vignette shows how to estimate and interpret **multinomial logit (mlogit)** models with the *marginaleffects* package. Multinomial logit models are useful when we wish to model choices among three or more unordered alternatives—for example, commuters selecting a mode of transport (car, bus, bicycle, walk). In these models the data are arranged in "long" format, with one row for every *unit–option* combination. Because this structure differs from the single‑row‑per‑unit layout used by most other models, current support in *marginaleffects* is limited to the `predictions()` and `avg_predictions()` functions. The examples below show that, by using the `hypothesis` argument, we can still obtain marginal effects and treatment effects that answer substantive questions about policy interventions.

## Note on implementation

This tutorial demonstrates how to work with multinomial logit models using `predictions()` together with custom hypothesis functions, rather than `comparisons()`. While `comparisons()` would be more complex to implement for multinomial models, the `predictions()` approach with custom hypotheses provides a flexible and powerful alternative. With minimal additional work, this approach can be extended to support more advanced analyses while maintaining the core functionality needed for most use‑cases.

## Fit the model

```{r}
#| message: false
#| warning: false
library(dplyr)
library(mlogit)
library(marginaleffects)

# Fishing mode‑choice data from mlogit
 data(Fishing)
 Fish <- dfidx(Fishing, varying = 2:9, shape = "wide", choice = "mode")

 m <- mlogit(mode ~ price + catch | income, data = Fish)
```

## Contrast analysis

Assume we want to know how a 100‑dollar increase in beach‐trip prices affects mode choice. We create two replicas of the original data: one with the observed prices (`lo`) and one with the higher beach price (`hi`). The `term` variable tags each scenario; later we will contrast them.

```{r}
# Prepare data for the two scenarios
lo <- transform(as.data.frame(Fish), term = "lo")
lo$id1 <- Fish$idx$id1
lo$id2 <- Fish$idx$id2
lo$idx <- NULL

hi <- transform(lo,
    price = ifelse(id2 == "beach", price + 100, price),
    term = "hi")

dat <- rbind(lo, hi)
dat$term <- factor(dat$term, levels = c("lo", "hi"))
```

## Average predicted probabilities

The call below obtains unit‑level fitted values and then averages them by choice *group* (the alternative) and *term* (scenario):

```{r}
#| warning: false
predictions(m,
    newdata = dat,
    by = c("group", "term"))
```

## Alternative: custom hypothesis function

The same result can be produced with a custom summary function supplied to the `hypothesis` argument. Custom functions offer considerable flexibility:

```{r}
#| warning: false
h <- \(x) aggregate(estimate ~ term + group, data = x, FUN = mean)
predictions(m, newdata = dat, hypothesis = h)
```

Instead of using the `aggregate()` function from base R, we can use the `dplyr` package to summarize the data.

```{r}
#| eval: false
library(dplyr) 
h <- function(x) {
  x |> summarize(estimate = mean(estimate), .by = c("term", "group"))
}
predictions(m, newdata = dat, hypothesis = h)
```

## Average marginal effects (risk differences)

Finally, we estimate the average effect of the price increase on the expected probability of each mode. We proceed in three steps inside a custom hypothesis function:

1. Average the unit‑level fitted values within each *term* and *group*.
2. Compute the difference between the `hi` and `lo` averages.
3. Rename the grouping variable to `term`, as expected by *marginaleffects*.

```{r}
#| warning: false
h <- function(x) {
    x |>
        aggregate(estimate ~ term + group, data = _, FUN = mean) |>
        aggregate(estimate ~ group, data = _, FUN = diff) |>
        setNames(c("term", "estimate"))
}
predictions(m, newdata = dat, hypothesis = h)
```

Interpretation: a 100‑dollar increase in beach‑trip prices lowers the probability of choosing the beach by roughly ten percentage points while modestly raising the probabilities of the other modes.

Or the same results can be obtained using `dplyr`:

```{r}
#| eval: false
h <- function(x) {
  x |>
    summarize(estimate = mean(estimate), .by = c("term", "group")) |>
    summarize(estimate = diff(estimate), .by = "group") |>
    rename(term = group)
}
predictions(m, newdata = dat, hypothesis = h)
```

