# Performance


## Why is `marginaleffects` slow, sometimes?

There are four primary reasons why some `marginaleffects` calls are slow:

1. Slow `predict()` function.
2. Too many comparisons.
3. Large grid.
4. Expensive standard errors.

Let's consider each of these points in turn, along with potential solutions.

```{r}
library(marginaleffects)
library(microbenchmark)
library(tictoc)
options(width = 300)

dat <- get_dataset("airbnb")

mod <- lm(
    price ~ bathrooms * bedrooms * unit_type * Breakfast * Gym, 
    data = dat)
```

## Slow `predict()` function

To compute predictions, comparisons, and slopes, `marginaleffects` must call `predict()` repeatedly. This function is supplied by the package that implements the model object, and its speed can vary widely. 

Unfortunately, since this function is not supplied by `marginaleffects` itself, there is not much that can be done to speed it up. Therefore, we move on directly to the other sources of sluggishness.

## Too many comparisons

When using the `slopes()` or `comparisons()` families of functions, the default is to compute the "effect" of all focal preditors one after the other. When there are many predictors, this can be expensive.

### Solution: `variables` argument

A simple solution is to specify the `variables` argument to only compute estimates for the predictors of interest. Compare:

```{r}
tic()
cmp <- avg_comparisons(mod, variables = "bathrooms")
toc()

tic()
cmp <- avg_comparisons(mod)
toc()
```

## Large grid

There are two main reasons why a grid might be large. 

First, the dataset used to fit the model has many rows, and you have not specified a `newdata` argument. In the fitted example above, the `airbnb` dataset is large. Therefore, if we do not specify a `newdata` argument, `marginaleffects` will compute one estimate (and standard error) for each row in the dataset. This can be very costly.

```{r}
p <- predictions(mod)
dim(p)
```

The second situation in which a grid can be large occurs when users have specified a grid in the `newdata` argument, perhaps using the `datagrid()` helper, and that grid is very large. 

Consider a simple dataset with 4 categorical variables, each with 26 possible levels:

```{r}
let <- data.frame(
    A = sample(letters, 1000, replace = TRUE),
    B = sample(letters, 1000, replace = TRUE),
    C = sample(letters, 1000, replace = TRUE),
    D = sample(letters, 1000, replace = TRUE)
)
nrow(let)
```

This dataset has 1000 rows. However, if we use the `datagrid()` function to create a "balanced grid", we end up with 26^4^=456,976 rows:

```{r}
dg <- datagrid(newdata = let, grid_type = "balanced")
nrow(dg)
```

### Solution A: Estimates at representative values

Instead of compute predictions, comparisons, or slopes for every row of a large grid, it is much faster to compute just a few estimates. For example, you may want to compute predictions for an individual with average (or modal) characteristics, or for a unit with specific values of interest.

```{r}
predictions(mod, newdata = "mean")

predictions(mod, newdata = datagrid(Gym = 1, bathrooms = 2, bedrooms = 6))
```

### Solution B: Random subset of the grid

An alternative is to sample a subset of rows at random from the large grid. If the subset is large enough, the statistics we compute on that basis will often be very similar to the full-sample values.

Here, we use `avg_comparisons()` to compute the "average effect" of `bathrooms` on `price`, and find that the estimates are reasonably close when we use the full grid or a subset.

```{r}
set.seed(48103)
# random subset of the grid
sub <- dat[sample(nrow(dat), 1000), ]
avg_comparisons(mod, variables = "bathrooms", newdata = sub)

# full grid
avg_comparisons(mod, variables = "bathrooms")
```

### Solution C: Weighted grid

When all the predictor variables are categorical, we can use a third approach: use a weighted grid of unique rows. Consider again our `airbnb` model. There are five categorical predictors in this model:

```{r}
pred <- insight::find_predictors(mod, flatten = TRUE)
pred
```

What we can do is create a grid of unique rows, and assign weights to each row based on how many times it appears in the original dataset.

```{r}
#| message: false
#| warning: false
library(dplyr)
grid <- dat |>
    select(all_of(pred)) |>
    summarise(weight = n(), .by = all_of(pred))
```

This grid is much smaller than the original dataset:

```{r}
nrow(dat)
nrow(grid)
```

Now we can compute a quantity of interest with `marginaleffects`, using the `wts` argument to "expand" the grid and obtain the same result as if we had used the full dataset, but much faster.

```{r}
# full dataset
tic()
p_full <- avg_predictions(mod)
toc()
```

```{r}
# weighted grid
tic()
p_subset <- avg_predictions(mod, newdata = grid, wts = "weight")
toc()
```

```{r}
p_full
p_subset
```

## Bayesian analyses

Bayesian analyses can require a lot of memory because `marginaleffects` often needs to store one prediction for every row of the grid and for every posterior draw. With `brms` models, the `ndraws` argument is passed forward to `posterior_epred()` and can be used to evaluate only a subset of the posterior draws. This makes the prediction matrices smaller and can make RAM requirements more reasonable in some applications.

```{r}
#| eval: false
avg_comparisons(
    mod,
    variables = "treatment",
    ndraws = 1000
)
```

## Standard errors are expensive

The final reason why some `marginaleffects` calls take time is that standard errors are expensive to compute. This is especially true when the fitted model includes many parameters (coefficients), because the default strategy to compute standard errors involves calling `predict()` at least once per coefficient.

### Solution A: No standard errors

It is always much faster to compute estimates without standard errors. If you do not need them, set `vcov=FALSE`.

```{r}
tic()
p <- avg_predictions(mod, vcov = FALSE)
toc()

tic()
p <- avg_predictions(mod)
toc()
```

### Solution B: Analytic Jacobians

For many models, `marginaleffects` computes standard errors from an exact
closed-form Jacobian rather than by finite differences. This is both faster and
more accurate: the default strategy calls `predict()` once per coefficient,
whereas the analytic path evaluates a single matrix expression, and it carries
no step-size error at all.

This happens automatically whenever the model and the estimand are eligible; there
is nothing to install and nothing to enable. The `jacobian_method` component
reports which path actually produced the numbers.

```{r}
p <- avg_predictions(mod)
components(p, "jacobian_method")
```

Eligible models include `lm`, `glm`, `MASS::glm.nb()`, `MASS::rlm()`,
`brglm2::brglmFit()`, `rms::ols()`, `rms::lrm()`, `quantreg::rq()`,
`ivreg::ivreg()`, `geepack::geeglm()`, and `survey::svyglm()`, on both the link
and response scales. Predictions, the built-in `comparison` functions, `by`
groups, weights, and linear `hypothesis` arguments are all supported. Anything
outside that set falls back to finite differences automatically.

To see the difference the analytic path makes, disable it and compare:

```{r}
#| warning: false
analytic <- function() {
    options(marginaleffects_analytic_jacobian = TRUE)
    avg_predictions(mod)
}
finite_differences <- function() {
    options(marginaleffects_analytic_jacobian = FALSE)
    avg_predictions(mod)
}
microbenchmark(
    analytic(),
    finite_differences(),
    times = 5
)
```

```{r}
#| include: false
options(marginaleffects_analytic_jacobian = TRUE)
```

### Solution C: Parallel processing

For models and calls which are not eligible for the analytic path, another option is to compute standard errors in parallel. 

Parallel processing does not always speed up `marginaleffects` much, because of the overhead involved in passing large datasets between different cores or forked processes. This strategy is more likely to be useful when:

1. The modelling package's `predict()` function is slow.
2. There are *many* parameters in the model.
3. The grid is not large enough to impose too much overhead.

To use parallel processing, we call the `plan` function from [the `future` package](https://future.futureverse.org/) and we set some global options.

:::{.callout-note}
The global option for parallel processing is different for the `inferences()` function. See the documentation of that function for details.
:::

Here is an example with `nycflights13` data and a complex model with many parameters.

```{r}
#| warning: false
library(mgcv) |> suppressPackageStartupMessages()
library(tictoc)
library(future)
library(nycflights13)
library(marginaleffects)
data("flights")
options(future.globals.maxSize = 8000 * 1024^2) # 8 GB memory
packageVersion("marginaleffects")
cores <- 8
plan(multicore, workers = cores)

flights <- flights |>
    transform(date = as.Date(paste(year, month, day, sep = "/"))) |>
    transform(date.num = as.numeric(date - min(date))) |>
    transform(wday = as.POSIXlt(date)$wday) |>
    transform(time = as.POSIXct(paste(hour, minute, sep = ":"), format = "%H:%M")) |>
    transform(time.dt = difftime(time, as.POSIXct('00:00', format = '%H:%M'), units = 'min')) |>
    transform(time.num = as.numeric(time.dt)) |>
    transform(dep_delay = ifelse(dep_delay < 0, 0, dep_delay)) |>
    transform(dep_delay = ifelse(is.na(dep_delay), 0, dep_delay)) |>
    transform(carrier = factor(carrier)) |>
    transform(dest = factor(dest)) |>
    transform(origin = factor(origin))

model <- bam(dep_delay ~ 
    s(date.num, bs = "cr") +
    s(wday, bs = "cc", k = 3) +
    s(time.num, bs = "cr") +
    s(carrier, bs = "re") +
    origin +
    s(distance, bs = "cr") + 
    s(dest, bs = "re"),
    data = flights,
    family = poisson,
    discrete = TRUE,
    nthreads = cores)

# there are many parameters
length(coef(model))
```

Now, compare three calls to `predictions()`.

```{r}
# No standard errors is very fast:
tic()
p1 <- predictions(model, vcov = FALSE)
toc()

# Parallel standard errors is slower
options("marginaleffects_parallel" = TRUE)
tic()
p1 <- predictions(model)
toc()

# Sequential standard errors is even slower
options("marginaleffects_parallel" = FALSE)
tic()
p2 <- predictions(model)
toc()
```
