
# Supported Models


```{r, echo = FALSE, message = FALSE, warning = FALSE}
options(tinytable_html_engine = "tabulator")
library("data.table")
library("tinytable")
dat <- fread("https://raw.githubusercontent.com/vincentarelbundock/marginaleffects/refs/heads/main/website/data/raw/supported_models.csv")
dat <- dat[order(tolower(dat$Package), tolower(dat$Function)), ]
```

In `Python`, `marginaleffects` supports the `statsmodels`, `sklearn`, `pyfixest`, and `linearmodels` modelling packages. 

In `R`, `marginaleffects` supports `r nrow(dat)` model types directly, and dozens more via the `tidymodels` and `mlr3` frameworks. This table shows the list of directly supported model types. 

If you would like to use a model that is not in this list, please [file a request or submit code on Github.](https://github.com/vincentarelbundock/marginaleffects)

```{r, echo = FALSE, message = FALSE, warning = FALSE}
tt(dat) |> theme_html(engine = "tabulator", tabulator_search = "top")
options(tinytable_html_engine = NULL)
```
