
# Tables


```{r, include = FALSE}
options(width = 1000)
## this vignette is in .Rbuildignore because lme4 is not available on old CRAN
## test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

## `tinytable`

By default, `marginaleffects` will print results to the console. However, it is easy to print in many other formats using the `tinytable` package, which supports a variety of output formats, including HTML, LaTeX, markdown, Word, and Typst.

https://vincentarelbundock.github.io/tinytable/

For example,

```{r}
library(marginaleffects)
mod <- glm(am ~ wt + drat, family = binomial, data = mtcars)
mfx <- avg_slopes(mod)
tab <- print(mfx, "tinytable")
class(tab)
```

We can then style this table using `tinytable`'s `style_tt()` function, add grouping columns or rows with `group_tt()`, format the content with `format_tt()`, etc. We can also save the table to file using `save_tt()`. For instance, to save to a Word document:

```r
save_tt(tab, "/path/to/table.docx")
```
 
## `modelsummary`

We can summarize the results of the `comparisons()` or `slopes()` functions using [the `modelsummary` package.](https://github.com/vincentarelbundock/modelsummary)

```{r}
library(modelsummary)
library(marginaleffects)

mod <- glm(am ~ wt + drat, family = binomial, data = mtcars)
mfx <- avg_slopes(mod)

modelsummary(mfx)
```

The same results can be visualized with `modelplot()`:

```{r}
modelplot(mfx)
```

````{comment}
We can display several models or marginal effects side by side:

```{r}
models <- list(
    glm(am ~ drat, family = binomial, data = mtcars),
    glm(am ~ wt + drat, family = binomial, data = mtcars))
models[[1]] <- slopes(models[[1]])
models[[2]] <- slopes(models[[2]], vcov = "HC3")
modelsummary(models)
```
````

When using the `avg_comparisons()` function (or the `avg_slopes()` function with categorical variables), the output will include two columns to uniquely identify the quantities of interest: `term` and `contrast`.

```{r}
dat <- mtcars
dat$gear <- as.factor(dat$gear)
mod <- glm(vs ~ gear + mpg, data = dat, family = binomial)

cmp <- avg_comparisons(mod)
get_estimates(cmp)
```

We can use the `shape` argument of the `modelsummary` function to structure the table properly:

```{r}
modelsummary(cmp, shape = term + contrast ~ model)
```

Cross-contrasts can be a bit trickier, since there are multiple simultaneous groups. Consider this example:

```{r}
mod <- lm(mpg ~ factor(cyl) + factor(gear), data = mtcars)
cmp <- avg_comparisons(
  mod,
  variables = c("gear", "cyl"),
  cross = TRUE)
get_estimates(cmp)
```

As we can see above, there are two relevant grouping columns: `contrast_gear` and `contrast_cyl`. We can simply plug those names in the `shape` argument:

```{r}
modelsummary(
  cmp,
  shape = contrast_gear + contrast_cyl ~ model)
```
