# Uncertainty

This vignette is a complement to the core chapter on hypothesis testing: @sec-uncertainty It presents more advanced examples.

```{r, include = FALSE}
options(width = 1000)
## this vignette is in .Rbuildignore because lme4 is not available on old CRAN
## test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

## Mixed effects models: Satterthwaite and Kenward-Roger corrections

For linear mixed effects models we can apply the Satterthwaite and Kenward-Roger corrections in the same way as in previous chapters:

```{r, message = FALSE}
library(marginaleffects)
library(patchwork)
library(lme4)

dat <- mtcars
dat$cyl <- factor(dat$cyl)
dat$am <- as.logical(dat$am)
mod <- lmer(mpg ~ hp + am + (1 | cyl), data = dat)
```

Marginal effects at the mean with classical standard errors and z-statistic:

```{r}
slopes(mod, newdata = "mean")
```

Marginal effects at the mean with Kenward-Roger adjusted variance-covariance and degrees of freedom:

```{r}
slopes(mod,
    newdata = "mean",
    vcov = "kenward-roger")
```

We can use the same option in any of the package's core functions, including:

```{r}
plot_predictions(mod, condition = "hp", vcov = "satterthwaite")
```

## Numerical derivatives: Sensitivity to step size

```{r}
dat <- get_dataset("penguins", "palmerpenguins")
dat$large_penguin <- ifelse(dat$body_mass_g > median(dat$body_mass_g, na.rm = TRUE), 1, 0)
mod <- glm(large_penguin ~ bill_length_mm * flipper_length_mm + species, data = dat, family = binomial)
```

`marginaleffects` uses numerical derivatives in two contexts:

1. Estimate the partial derivatives reported by `slopes()` function. 
   - Centered finite difference
   - $\frac{f(x + \varepsilon_1 / 2) - f(x - \varepsilon_1 / 2)}{\varepsilon_1}$, where we take the derivative with respect to a predictor of interest, and $f$ is the `predict()` function.
2. Estimate standard errors using the delta method.
   - Forward finite difference
   - $\frac{g(\hat{\beta}) - g(\hat{\beta} + \varepsilon_2)}{\varepsilon_2}$, where we take the derivative with respect to a model's coefficients, and $g$ is a `marginaleffects` function which returns some quantity of interest (slopes, contrasts, predictions, etc.)
     
Note that the step sizes used in those two contexts can differ. If the variables and coefficients have very different scales, it may make sense to use different values for $\varepsilon_1$ and $\varepsilon_2$.

By default, $\varepsilon_1$ is set to `1e-4` times the range of the variable with respect to which we are taking the derivative. By default, $\varepsilon_2$ is set to the maximum value of `1e-8`, or `1e-4` times the smallest absolute coefficient estimate. (These choices are arbitrary, but I have found that in practice, smaller values can produce unstable results.)

$\varepsilon_1$ can be controlled by the `eps` argument of the `slopes()` function. $\varepsilon_2$ can be controlled by setting a global option which tells `marginaleffects` to compute the jacobian using the `numDeriv` package instead of its own internal functions. This allows more control over the step size, and also gives access to other differentiation methods, such as Richardson's. To use `numDeriv`, we define a list of arguments which will be pushed forward to `numDeriv::jacobian`:

```{r}
avg_slopes(mod, variables = "bill_length_mm")

options(marginaleffects_numDeriv = list(method = "Richardson"))
avg_slopes(mod, variables = "bill_length_mm")

options(marginaleffects_numDeriv = list(method = "simple", method.args = list(eps = 1e-3)))
avg_slopes(mod, variables = "bill_length_mm")

options(marginaleffects_numDeriv = list(method = "simple", method.args = list(eps = 1e-5)))
avg_slopes(mod, variables = "bill_length_mm")

options(marginaleffects_numDeriv = list(method = "simple", method.args = list(eps = 1e-7)))
avg_slopes(mod, variables = "bill_length_mm")
```

Notice that the standard errors can vary considerably when using different step sizes. It is good practice for analysts to consider the sensitivity of their results to this setting.

Now, we illustrate the full process of standard error computation, using raw `R` code. First, we choose two step sizes:

```{r}
eps1 <- 1e-5 # slope
eps2 <- 1e-7 # delta method

s <- slopes(mod, newdata = head(dat, 3), variables = "bill_length_mm", eps = eps1)
print(s[, 1:5], digits = 6)
```

We can get the same estimates manually with these steps:

```{r}
linkinv <- mod$family$linkinv

## increment the variable of interest by h
dat_hi <- transform(dat, bill_length_mm = bill_length_mm + eps1)

## model matrices: first 3 rows
mm_lo <- insight::get_modelmatrix(mod, data = dat)[1:3,]
mm_hi <- insight::get_modelmatrix(mod, data = dat_hi)[1:3,]

## predictions
p_lo <- linkinv(mm_lo %*% coef(mod))
p_hi <- linkinv(mm_hi %*% coef(mod))

## slopes
(p_hi - p_lo) / eps1
```

To get standard errors, we build a jacobian matrix where each column holds derivatives of the vector valued slope function, with respect to each of the coefficients. Using the same example:

```{r}
b_lo <- b_hi <- coef(mod)
b_hi[1] <- b_hi[1] + eps2

dydx_lo <- (linkinv(mm_hi %*% b_lo) - linkinv(mm_lo %*% b_lo)) / eps1
dydx_hi <- (linkinv(mm_hi %*% b_hi) - linkinv(mm_lo %*% b_hi)) / eps1
(dydx_hi - dydx_lo) / eps2
```

This gives us the first column of $J$, which we can recover in full from the `marginaleffects` object attribute:

```{r}
J <- components(s, "jacobian")
J
```

To build the full matrix, we would simply iterate through the coefficients, incrementing them one after the other. Finally, we get standard errors via:

```{r}
sqrt(diag(J %*% vcov(mod) %*% t(J)))
```

Which corresponds to our original standard errors:

```{r}
print(s[, 1:5], digits = 7)
```

Reverting to default settings:

```{r}
options(marginaleffects_numDeriv = NULL)
```

Note that our default results for this model are very similar -- but not exactly identical -- to those generated by the `margins`. As should be expected, the results in `margins` are also very sensitive to the value of `eps` for this model:

```{r}
#| eval: false
library(margins)
margins(mod, variables = "bill_length_mm", data = head(dat, 3), unit_ses = TRUE)$SE_dydx_bill_length_mm

margins(mod, variables = "bill_length_mm", data = head(dat, 3), eps = 1e-4, unit_ses = TRUE)$SE_dydx_bill_length_mm

margins(mod, variables = "bill_length_mm", data = head(dat, 3), eps = 1e-5, unit_ses = TRUE)$SE_dydx_bill_length_mm
```


## Bayesian estimates and credible intervals

[See the `brms` vignette](brms.html) for a discussion of bayesian estimates and credible intervals.


<!--
## Uncertainty around unit-level or average estimates

Note that it is normal to observe that confidence intervals around average estimates tend to be tighter than those around unit-level estimates:

```{r}
library(marginaleffects)
mod <- lm(mpg ~ hp * qsec, mtcars)

## On average CIs around unit-level estimates are:
pre1 <- predictions(mod)
as.numeric(mean(pre1$conf.high - pre1$conf.low))

## The CI of the average estimate is:
pre2 <- tidy(predictions(mod))
as.numeric(pre2$conf.high - pre2$conf.low)

## On average CIs around unit-level estimates are:
cmp1 <- comparisons(mod, variables = "hp")
mean(cmp1$conf.high - cmp1$conf.low)

## The CI of the average estimate is:
cmp2 <- comparisons(mod, variables = "hp", comparison = "differenceavg")
cmp2$conf.high - cmp2$conf.low
```

Relevant?
https://en.wikipedia.org/wiki/Jensen%27s_inequality
https://math.stackexchange.com/questions/34009/is-the-variance-of-a-convex-function-a-convex-function/34021#34021?newreg=698ab30a481945878307935ac5d6b629


NO LONGER RELEVANT WITH THE AVERAGE() + RECALL() REFACTOR SINCE WE COMPUTE SEs DIRECTLY ON THE MARGINS W

### Standard Error of the Average Marginal Effect or Contrast

The `avg_*()` functions computes the average marginal effect (or contrast) when they are applied to an object produced by `slopes()` (or `comparisons()`). This is done in 3 steps:

1. Extract the Jacobian used to compute unit-level standard errors.
2. Take the average of that Jacobian.
3. Estimate the standard error of the average marginal effect as the square root of the diagonal of J'VJ, where V is the variance-covariance matrix.

As [explained succinctly on Stack Exchange:](https://stats.stackexchange.com/a/331311/4874)

> we want the variance of the Average Marginal Effect (AME) and hence our transformed function is: $AME =  \frac{1}{N} \sum_{i=1}^N g_i(x_i,\hat{\beta})$ Then using the delta method we have $Var \left( g(\hat{\beta}) \right) = J_g' \Omega_{\hat{\beta}} J_g$ where $\Omega_{\hat{\beta}} = Var(\hat{\beta})$ and $J_g' = \frac{\partial\left[\frac{1}{N}\sum_{i=1}^N g (x_i,\hat{\beta})\right]}{\partial \hat\beta} = \frac{1}{N}{\left[\sum_{i=1}^N \frac{\partial \left (g (x_i,\hat{\beta})\right)}{\partial \hat\beta}\right]}$ Which justifies using the "average Jacobian" in the delta method to calculate variance of the AME.

References:

* Dowd, Bryan E, William H Greene, and Edward C Norton. “Computation of Standard Errors.” Health Services Research 49, no. 2 (April 2014): 731–50. https://doi.org/10.1111/1475-6773.12122.
* https://stats.stackexchange.com/questions/283831/delta-method-for-marginal-effects-of-generalized-linear-model?rq=1
-->
