---
title: "Bayesian analysis with `brms` and `marginaleffects`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marginal Effects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .4,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

ggplot2::theme_set(ggplot2::theme_minimal())
```

Support for `brms` is still experimental. Bug reports and feature requests can be submitted here: https://github.com/vincentarelbundock/marginaleffects/issues

I hope to improve performance considerably in the very near future.

Install the development version of the `marginaleffects` package:

```{r, eval = FALSE}
library(remotes)
install_github("vincentarelbundock/marginaleffects")
```

Load libraries:

```{r, message = FALSE}
library(brms)
library(marginaleffects)
library(ggplot2)
library(ggdist)
```

Fit a logit model with a multiplicative interaction:

```{r, message = FALSE, warning = FALSE, results = "hide"}
mod <- brm(am ~ mpg * vs + hp, family = bernoulli(link = "logit"), data = mtcars)
```

Use `marginaleffects()` to compute marginal effects for each row of the dataset, and use `summary()` to compute "Average Marginal Effects": 

```{r}
mfx <- marginaleffects(mod)
summary(mfx)
```

Compute marginal effects with some regressors fixed at user-specified values, and other regressors held at their means:

```{r}
marginaleffects(mod, newdata = typical(hp = c(110, 130)))
```

The `get_posterior_draws` produces a dataset with `drawid` and `draw` columns.

```{r}
draws <- get_posterior_draws(mfx)

dim(draws)

head(draws)
```

We can use this dataset to plot our results. For example, to plot the posterior density of the marginal effect when the `vs` variable is equal to 0 or 1:

```{r}
marginaleffects(mod, variables = "mpg", newdata = typical(vs = 0:1)) |>
    get_posterior_draws() |>
    ggplot(aes(x = draw, color = factor(vs), fill = factor(vs))) +
    geom_density(alpha = .1) +
    labs(x = "Marginal effect of Miles per Gallon",
         y = "Posterior density")
```


A similar strategy can be used to compute and display model predictions with the `predictions` function:

```{r, message = FALSE, warning = FALSE, results = "hide"}
mod <- brm(mpg ~ hp * vs + am, data = mtcars)
```

```{r}
predictions(mod, newdata = typical(vs = 0:1, hp = c(100, 200))) |>
    get_posterior_draws() |>
    ggplot(aes(y = factor(vs), x = draw)) +
    stat_gradientinterval() +
    facet_wrap(~ hp) +
    labs(x = "Predicted Miles per Gallon",
         y = "Engine shape (vs)")
```
