---
format: 
  html: 
    toc: true
    toc-expand: 2
---


# Comparisons 

```{r, include = FALSE}
options(width = 1000)
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .4,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

library(marginaleffects)
library(patchwork)
library(ggplot2)

theme_set(theme_minimal())
```

In this vignette, we introduce "comparisons", defined as:

> Comparisons are functions of two or more predictions. Example of comparisons include contrasts, differences, risks ratios, odds, etc.

The `comparisons()` function is *extremely* flexible, and it allows users to estimate a vast array of quantities of interest. To describe those quantities, we will break the problem down into in 4 steps:

1. Quantity of interest
2. Grid
3. Average
4. Hypothesis

These steps can be combined, mixed, and matched to define and compute many different estimands.

Throughout this vignette, we will consider a logistic regression model of passenger survival aboard the Titanic. We will consider variations on this questions:

> How does the probability of survival (outcome) change if a passenger travels in 1st class vs. 3rd class?

```{r}
library(marginaleffects)

dat <- "https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Titanic.csv"
dat <- read.csv(dat, na.strings = c("*", ""))

mod <- glm(Survived ~ PClass * SexCode * Age, data = dat, family = binomial)

summary(mod)
```

These coefficient estimates are interesting, but difficult to interpret. In many applications, analysts will want to call the `comparisons()` or `avg_comparisons()` functions to report more meaningful quantities of interest. The rest of explains in detail how to interpret and customize calls like this one:

```{r}
avg_comparisons(mod)
```


# Quantity of interest

To investigate the effect of gender, age, and passenger class on survival aboard the Titanic, we will conduct a series of "comparisons" between predicted outcomes computed for different preditor values. The quantity of interest will vary in two main respects:

* *Predictor type*: How does the explanator of interest change?
* *Comparison type*: How do we compare the two predicted outcomes?

By default, the `comparisons()` function will estimate the "effect" of different changes for different types of predictors:

* Numeric: Increase of 1 unit.
* Logical: Change from `FALSE` to `TRUE`.
* Factor or character: Changes from the reference to each other level.


## Comparison type

The default comparison type in  `comparisons()` is the "difference", which means that the  `comparisons()` function will typically produce quantities which are often interpreted as a measure of "effect" in applications: risk differences, average treatment effects, G-computation estimates, etc. But `comparisons()` is not limited to differences, Indeed, it allows analysts to call on many built-in functions for comparison, and also to supply their own.

### Differences

By default, the `comparisons()` function will compare the predicted values by differencing (subtraction). For example, consider a hypothetical Titanic passenger with these characteristics:

```{r}
passenger <- data.frame(
    PClass = "3rd",
    Age = 20,
    SexCode = 0
)
```

What would happen to the predicted probability of survival if we were to increase the `SexCode` from 0 to 1? Or increase `Age` by 1 unit? Or change `PClass` from "3rd" to "1st"? To answer these questions, we could proceed manually by computing two predictions and then subtracting one from the other:

```{r}
passenger_0 <- subset()
passenger_1 <- transform(passenger, SexCode = 1)

prediction_0 = predict(mod, newdata = passenger_0, type = "response")
prediction_1 = predict(mod, newdata = passenger_1, type = "response")

prediction_1 - prediction_0
```

The result above shows that modifying the `SexCode` variable of this hypothetical passenger from 0 to 1 increases the predicted probability of survival by about `r sprintf("%.f%%", 100 * (prediction_1-prediction_0))`

Instead of computing this estimate by hand, we could call `comparisons()` to get analogous results for all predictors in the model at once:

```{r}
comparisons(mod, newdata = passenger)
```

### Ratios

Instead of taking simple differences between adjusted predictions, it can sometimes be useful to compute ratios or other functions of predictions. For example, [the `adjrr` function the `Stata` software package](https://journals.sagepub.com/doi/pdf/10.1177/1536867X1301300304) can compute "adjusted risk ratios", which are ratios of adjusted predictions. To do this in `R`, we could use the same predictions we computed above:

```{r}
prediction_1 / prediction_0
```

This shows that the predicted probability of survival would be about `r sprintf("%.1f%%", (prediction_1/prediction_0))` larger if our hypothetical passenger were a woman than a man. 

The same result can be obtained by setting the `comparison` argument:

```{r}
comparisons(mod, comparison = "ratio", newdata = passenger)
```


### Built-in functions

Beyond differences and ratios, `comparisons()` supports many other built-in functions to compare predictions, such as log odds ratios, lift, etc. See `?comparisons` for a complete list.

These built-in functions allow us convenient access to complex transformations. For example, this code computes the log odds ratio associated with a change in predictors:

```{r}
comparisons(mod, 
    comparison = "lnor", 
    newdata = passenger)
```

This call adds the `transform` argument to exponentiate the log-odds ratios at the very end of calculations:

```{r}
comparisons(mod, 
    comparison = "lnor", 
    transform = exp,
    newdata = passenger)
```

### Custom functions

```{r}
comparisons(mod, comparison = function(hi, lo) hi / lo, newdata = passenger)
```

This mechanism is powerful, because it lets users create fully customized contrasts. Here is a non-sensical example:

```{r}
avg_comparisons(mod, comparison = function(hi, lo) sqrt(hi) / log(lo + 10))
```



## Predictor types

This section seeks to present how to use the `comparisons()` function depending on the type of predictor. Four will be presented:

1. Numeric
2. Logical and factor 
  - Observations
3. Character
4. Interactions and Cross-Constrasts


### Numeric

We can also compute contrasts for differences in numeric variables. For example, we can see what happens to the adjusted predictions when we increment the `hp` variable by 1 unit (default) or by 5 units about the original value:

```{r}
mod <- lm(mpg ~ hp, data = mtcars)

avg_comparisons(mod)

avg_comparisons(mod, variables = list(hp = 5))
```

Compare adjusted predictions for a change in the regressor between two arbitrary values:

```{r}
avg_comparisons(mod, variables = list(hp = c(90, 110)))
```

Compare adjusted predictions when the regressor changes across the interquartile range, across one or two standard deviations about its mean, or from across its full range:

```{r}
avg_comparisons(mod, variables = list(hp = "iqr"))

avg_comparisons(mod, variables = list(hp = "sd"))

avg_comparisons(mod, variables = list(hp = "2sd"))

avg_comparisons(mod, variables = list(hp = "minmax"))
```

### Logical and factor

Consider a simple model with a logical and a factor variable:

```{r}
library(marginaleffects)

tmp <- mtcars
tmp$am <- as.logical(tmp$am)
mod <- lm(mpg ~ am + factor(cyl), tmp)
```

The `comparisons` function automatically computes contrasts for each level of the categorical variables, relative to the baseline category (`FALSE` for logicals, and the reference level for factors), while holding all other values at their observed values. The `avg_comparisons()` does the same, but then marginalizes by taking the average of unit-level estimates:

```{r}
cmp <- avg_comparisons(mod)
cmp
```

The summary printed above says that moving from the reference category `4` to the level `6` on the `cyl` factor variable is associated with a change of `r sprintf("%.3f", tidy(cmp)$estimate[2])` in the adjusted prediction. Similarly, the contrast from `FALSE` to `TRUE` on the `am` variable is equal to `r sprintf("%.3f", tidy(cmp)$estimate[1])`.

We can obtain different contrasts by using the `comparisons()` function. For example:

```{r}
avg_comparisons(mod, variables = list(cyl = "sequential"))

avg_comparisons(mod, variables = list(cyl = "pairwise"))

avg_comparisons(mod, variables = list(cyl = "reference"))
```

For comparison, this code produces the same results using the `emmeans` package:

```{r}
library(emmeans)
emm <- emmeans(mod, specs = "cyl")
contrast(emm, method = "revpairwise")

emm <- emmeans(mod, specs = "am")
contrast(emm, method = "revpairwise")
```

Note that these commands also work on for other types of models, such as GLMs, on different scales:

```{r}
mod_logit <- glm(am ~ factor(gear), data = mtcars, family = binomial)

avg_comparisons(mod_logit)

avg_comparisons(mod_logit, type = "link")
```

#### Observation-Wise Categorical Marginal Effect

For categorical predictors, [Scholbeck et al. 2023](https://arxiv.org/pdf/2201.08837.pdf) recommend that analysts report what they call the "observation-wise categorical marginal effects." They describe the procedure as follows:

> Recall that the common definition of categorical MEs is based on first changing all observations' value of $x_j$ to each category and then computing the difference in predictions when changing it to the reference category. However, one is often interested in prediction changes if aspects of an actual observation change. We therefore propose an observation-wise categorical ME. We first select a single reference category $c_h$. For each observation whose feature value $x_j \neq c_h$, we predict once with the observed value $x_j$ and once where $x_j$ has been replaced by $c_h$.

To achieve this with `marginaleffects`, we proceed in three simple steps:

1.  Use the `factor()` function to set the reference level of the categorical variable.
2.  Use the `newdata` argument to take the subset of data where the observed $x_j$ is different from the reference level we picked in 1.
3.  Apply the `avg_comparisons()` with the `"revreference"` option.

```{r}
tmp <- transform(mtcars, cyl = factor(cyl, levels = c(6, 4, 8)))

mod <- glm(vs ~ mpg * factor(cyl), data = tmp, family = binomial)

avg_comparisons(mod,
  variables = list(cyl = "revreference"),
  newdata = subset(tmp, cyl != 6))
```

### Character

All functions of the `marginaleffects` package attempt to treat character predictors as factor predictors. However, using factors instead of characters when modeling is *strongly* encouraged, because they are much safer and faster. This is because factors hold useful information about the full list of levels, which makes them easier to track and handle internally by `marginaleffects`. Users are strongly encouraged to convert their character variables to factor before fitting their models and using `slopes` functions.

### Interactions and Cross-Contrasts

In some contexts we are interested in whether the "effect" of a variable changes, as a function of another variable. A very simple strategy to tackle this question is to estimate a model with a multiplicative interaction like this one:

```{r}
mod <- lm(mpg ~ am * factor(cyl), data = mtcars)
```

Calling `avg_comparisons()` with the `by` argument shows that the estimated comparisons differ based on `cyl`:

```{r}
avg_comparisons(mod, variables = "am", by = "cyl")
```

However, using the `hypothesis` argument for pairwise contrasts between the above comparisons reveals that the heterogeneity is not statistically significant:

```{r}
avg_comparisons(mod, variables = "am", by = "cyl", hypothesis = "pairwise")
```

In other contexts, we are interested in a "cross-contrast" or "cross-comparisons"; we would like to know what happens when two (or more) predictors change at the same time. To assess this, we can specify the regressors of interest in the `variables` argument, and set the `cross=TRUE`:

```{r}
avg_comparisons(mod, variables = c("cyl", "am"), cross = TRUE)
```


# Grid 

In GLM models, most quantities of interest are *conditional*, in the sense that they will typically depend on the values of all the predictors in the model. Therefore, we need to decide *where* in the predictor space we want to evaluate the quantity of interest described above.

By default, `comparisons()` will compute estimates for every row of the original dataset that was used to fit a model. There are `r nrow(dat)` observations in the titanic dataset. Therefore, if we just execute the code in the previous section, we will obtain `r nrow(dat)` estimates of the difference between the probability of survival in 3rd and 1st class:

```{r}
mod <- glm(Survived ~ PClass * SexCode * Age, data = dat, family = binomial)

comparisons(mod,
  variables = list(PClass = c("1st", "3rd"))) # Step 1: Quantity
```

Notice that the contrast between 3rd and 1st is different from row to row. This reflects the fact that, in our model, moving from 1st to 3rd would have a different effect on the predicted probability of survival for different individuals. 

We can be more specific in our query. Instead of using the empirical distribution as our "grid", we can specify exactly where we want to evaluate the comparison in the predictor space, by using the `newdata` argument and the `datagrid()` function. For example, say I am interested in:

> The effect of moving from 1st to 3rd class on the probability of survival for a 50 year old man and a 50 year old woman.

I can type:

```{r}
cmp <- comparisons(mod,
  variables = list(PClass = c("1st", "3rd")),  # Step 1: Quantity
  newdata = datagrid(Age = 50, SexCode = 0:1)) # Step 2: Grid
cmp
```

We now know that moving from 1st to 3rd changes by `r sprintf("%.3f", cmp$estimate[1])` the probability of survival for 50 year old men (`SexCode=0`), and by `r sprintf("%.3f", cmp$estimate[2])` the probability of survival for 50 year old women (`SexCode=1`).



### Contrasts at the mean

An alternative which used to be very common but has now fallen into a bit of disfavor is to compute "Contrasts at the mean." The idea is to create a "synthetic" or "hypothetical" individual (row of the dataset) whose characteristics are completely average. Then, we compute and report the contrast for this specific hypothetical individual. 

This can be achieved by setting `newdata="mean"` or to `newdata=datagrid()`, both of which fix variables to their means or modes:

```{r}
comparisons(mod, variables = "mpg", newdata = "mean")
```

Contrasts at the mean can differ substantially from average contrasts.

The advantage of this approach is that it is very cheap and fast computationally. The disadvantage is that the interpretation is somewhat ambiguous. Often times, there simply does not exist an individual who is perfectly average across all dimensions of the dataset. It is also not clear why the analyst should be particularly interested in the contrast for this one, synthetic, perfectly average individual.


## Unit-level contrasts

In models with interactions or non-linear components (e.g., link function), the value of a contrast or marginal effect can depend on the value of all the predictors in the model. As a result, contrasts and marginal effects are fundamentally *unit-level* quantities. The effect of a 1 unit increase in $X$ can be different for Mary or John. Every row of a dataset has a different contrast and marginal effect.

The `mtcars` dataset has 32 rows, so the `comparisons()` function produces 32 contrast estimates:

```{r, warning = FALSE}
library(marginaleffects)
mod <- glm(vs ~ factor(gear) + mpg, family = binomial, data = mtcars)
cmp <- comparisons(mod, variables = "mpg")
nrow(cmp)
```

## Marginal means

Yet another type of contrast is the "Contrast between marginal means." This type of contrast is closely related to the "Contrast at the mean", with a few wrinkles. It is the default approach used by the `emmeans` package for `R`.

Roughly speaking, the procedure is as follows: 

1. Create a prediction grid with one cell for each combination of categorical predictors in the model, and all numeric variables held at their means.
2. Make adjusted predictions in each cell of the prediction grid.
3. Take the average of those predictions (marginal means) for each combination of `btype` (focal variable) and `resp` (group `by` variable).
4. Compute pairwise differences (contrasts) in marginal means across different levels of the focal variable `btype`.

The contrast obtained through this approach has two critical characteristics:

a) It is the contrast for a synthetic individual with perfectly average qualities on every (numeric) predictor.
b) It is a weighted average of unit-level contrasts, where weights assume a perfectly balanced dataset across every categorical predictor.

With respect to (a), the analyst should ask themselves: Is my quantity of interest the contrast for a perfectly average hypothetical individual? With respect to (b), the analyst should ask themselves: Is my quantity of interest the contrast in a model estimated using (potentially) unbalanced data, but interpreted *as if* the data were perfectly balanced? 

For example, imagine that one of the control variables in your model is a variable measuring educational attainment in 4 categories: No high school, High school, Some college, Completed college. The contrast between marginal is a weighted average of contrasts estimated in the 4 cells, and each of those contrasts will be weighted equally in the overall estimate. If the population of interest is highly unbalanced in the educational categories, then the estimate computed in this way will not be most useful.

If the contrasts between marginal means is really the quantity of interest, it is easy to use the `comparisons()` to estimate contrasts between marginal means. The `newdata` determines the values of the predictors at which we want to compute contrasts. We can set `newdata="marginalmeans"` to emulate the `emmeans` behavior. For example, here we compute contrasts in a model with an interaction:

```{r}
dat <- read.csv("https://vincentarelbundock.github.io/Rdatasets/csv/palmerpenguins/penguins.csv")
mod <- lm(bill_length_mm ~ species * sex + island + body_mass_g, data = dat)

avg_comparisons(
    mod,
    newdata = "marginalmeans",
    variables = c("species", "island"))
```

Which is equivalent to this in `emmeans`:

```{r}
library(emmeans)
emm <- emmeans(
    mod,
    specs = c("species", "island"))
contrast(emm, method = "trt.vs.ctrl1")
```

The [`emmeans` section of the Alternative Software vignette](alternative_software.html#emmeans) shows further examples.

The [excellent vignette of the `emmeans` package](https://CRAN.R-project.org/package=emmeans/vignettes/basics.html) discuss the same issues in a slightly different (and more positive) way:

> The point is that the marginal means of cell.means give equal weight to each cell. In many situations (especially with experimental data), that is a much fairer way to compute marginal means, in that they are not biased by imbalances in the data. We are, in a sense, estimating what the marginal means would be, had the experiment been balanced. Estimated marginal means (EMMs) serve that need.

> All this said, there are certainly situations where equal weighting is not appropriate. Suppose, for example, we have data on sales of a product given different packaging and features. The data could be unbalanced because customers are more attracted to some combinations than others. If our goal is to understand scientifically what packaging and features are inherently more profitable, then equally weighted EMMs may be appropriate; but if our goal is to predict or maximize profit, the ordinary marginal means provide better estimates of what we can expect in the marketplace.



# Aggregation

Again, by default `comparisons()` estimates quantities for all the actually observed units in our dataset. Sometimes, it is convenient to marginalize those conditional estimates, in order to obtain an "average contrast": 

```{r}
avg_comparisons(mod,                          # Step 3: Average
  variables = list(PClass = c("1st", "3rd"))) # Step 1: Quantity
```

Alternatively, we could also take the average, but just of the two estimates that we computed above for the 50 year old man and 50 year old woman. 

```{r}
avg_comparisons(mod,                           # Step 3: Average
  variables = list(PClass = c("1st", "3rd")),  # Step 1: Quantity
  newdata = datagrid(Age = 50, SexCode = 0:1)) # Step 2: Grid
```

Notice that this is exactly the same as the average in the estimates from the previous section, which we had stored as `cmp`:

```{r}
cmp$estimate

mean(cmp$estimate)
```

# Uncertainty

By default, the standard errors around contrasts are computed using the delta method on the scale determined by the `type` argument (e.g., "link" or "response"). Some analysts may prefer to proceed differently. For example, in `Stata`, the `adjrr` computes adjusted risk ratios (ARR) in two steps:

1.  Compute the natural log of the ratio between the mean of adjusted predictions with $x+1$ and the mean of adjusted predictions with $x$.
2.  Exponentiate the estimate and confidence interval bounds.

Step 1 is easy to achieve with the `comparison` argument described above. Step 2 can be achieved with the `transform` argument:

```{r}
avg_comparisons(
    mod,
    comparison = function(hi, lo) log(hi / lo),
    transform = exp)
```

Note that we can use the `lnratioavg` shortcut instead of defining the function ourselves.

The order of operations in previous command was:

1.  Compute the custom unit-level log ratios
2.  Exponentiate them
3.  Take the average using the `avg_comparisons()`

There is a very subtle difference between the procedure above and this code:

```{r}
avg_comparisons(
    mod,
    comparison = function(hi, lo) log(hi / lo),
    transform = exp)
```

Since the `exp` function is now passed to the `transform` argument of the `comparisons()` function, the exponentiation is now done only *after* unit-level contrasts have been averaged. This is what `Stata` appears to do under the hood, and the results are slightly different.

```{r}
comparisons(
    mod,
    comparison = function(hi, lo) log(mean(hi) / mean(lo)),
    transform = exp)
```

Note that equivalent results can be obtained using shortcut strings in the `comparison` argument: "ratio", "lnratio", "lnratioavg".

```{r}
comparisons(
    mod,
    comparison = "lnratioavg",
    transform = exp)
```


# Hypothesis

Finally, imagine we are interested in this question:

> Does moving from 1st to 3rd class have a bigger effect on the probability of survival for 50 year old men, or for 50 year old women?

To answer this, we use the `hypothesis` argument:

```{r}
comparisons(mod,
  variables = list(PClass = c("1st", "3rd")),  # Step 1: Quantity
  newdata = datagrid(Age = 50, SexCode = 0:1), # Step 2: Grid
  hypothesis = "b1 = b2")                      # Step 4: Hypothesis
```

This result maps directly onto the estimates we had above. It is the difference in the contrast for 50-men and 50-women:

```{r}
diff(cmp$estimate)
```

This result can be interpreted as a "difference-in-differences": Moving from 1st to 3rd has a much larger negative effect on the probability of survival for a 50 year old woman than for a 50 year old man. This difference is statistically significant.

We can do a similar comparison, but instead of fixing a conditional grid, we can average over subgroups of the empirical distribution, using the `by` argument:

```{r}
avg_comparisons(mod,
  variables = list(PClass = c("1st", "3rd")),  # Step 1: Quantity
  by = "SexCode",                              # Step 3: Average
  hypothesis = "b1 = b2")                      # Step 4: Hypothesis
```

### Average contrasts

By default, the `slopes()` and `comparisons()` functions compute marginal effects and contrasts for every row of the original dataset. These unit-level estimates can be of great interest, as discussed in [another vignette](logit.html). Nevertheless, one may want to focus on one-number summaries: the `avg_*()` functions or the `by` argument compute the "Average Marginal Effect" or "Average Contrast," by taking the mean of all the unit-level estimates.

```{r}
avg_comparisons(mod, variables = "mpg")

comparisons(mod, variables = "mpg", by = TRUE)
```

which are equivalent to:

```{r}
mean(cmp$estimate)
```

We could also show the full distribution of contrasts across our dataset with a histogram:

```{r}
#| fig.asp = .4
library(ggplot2)

cmp <- comparisons(mod, variables = "gear")

ggplot(cmp, aes(estimate)) +
    geom_histogram(bins = 30) +
    facet_wrap(~contrast, scale = "free_x") +
    labs(x = "Distribution of unit-level contrasts")
```

This graph displays the effect of a change of 1 unit in the `gear` variable, for each individual in the observed data.



# Transformations

So far we have focused on simple differences between adjusted predictions. Now, we show how to use ratios, back transformations, and arbitrary functions to estimate a slew of quantities of interest. Powerful transformations and custom contrasts are made possible by using three arguments which act at different stages of the computation process:

-   `comparison`
-   `transform`

Consider the case of a model with a single predictor $x$. To compute average contrasts, we proceed as follows:

1.  Compute adjusted predictions for each row of the dataset for the observed values $x$: $\hat{y}_x$
2.  Compute adjusted predictions for each row of the dataset for the observed values $x + 1$: $\hat{y}_{x+1}$
3.  `comparison`: Compute unit-level contrasts by taking the difference between (or some other function of) adjusted predictions: $\hat{y}_{x+1} - \hat{y}_x$
4.  Compute the average contrast by taking the mean of unit-level contrasts: $1/N \sum_{i=1}^N \hat{y}_{x+1} - \hat{y}_x$
5.  `transform`: Transform the average contrast or return them as-is.

The `comparison` argument of the `comparisons()` function determines how adjusted predictions are combined to create a contrast. By default, we take a simple difference between predictions with `hi` value of $x$, and predictions with a `lo` value of $x$: `function(hi, lo) hi-lo`.

The `transform` argument of the `comparisons()` function applies a custom transformation to the unit-level contrasts.

The `transform` argument applies a custom transformation to the final quantity, as would be returned if we evaluated the same call without `transform`.


# Plotting

The same arguments work in the plotting function `plot_comparisons()` as well, which allows us to plot various custom contrasts. Here is a comparison of Adjusted Risk Ratio and Adjusted Risk Difference in a model of the probability of survival aboard the Titanic:

```{r}
library(ggplot2)
library(patchwork)
titanic <- "https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Titanic.csv"
titanic <- read.csv(titanic)
mod_titanic <- glm(
    Survived ~ Sex * PClass + Age + I(Age^2),
    family = binomial,
    data = titanic)

avg_comparisons(mod_titanic)

p1 <- plot_comparisons(
    mod_titanic,
    variables = "Age",
    condition = "Age",
    comparison = "ratio") +
    ylab("Adjusted Risk Ratio\nP(Survival | Age + 1) / P(Survival | Age)")

p2 <- plot_comparisons(
    mod_titanic,
    variables = "Age",
    condition = "Age") +
    ylab("Adjusted Risk Difference\nP(Survival | Age + 1) - P(Survival | Age)")

p1 + p2
```

# Visual guide

This section illustrates some of the concepts in this vignette visually.

Consider a model with an interaction term. What happens to the dependent variable when the `hp` variable increases by 10 units?

```{r}
library(marginaleffects)

mod <- lm(mpg ~ hp * wt, data = mtcars)

plot_comparisons(
    mod,
    variables = list(hp = 10),
    condition = "wt")
```

```{r}
okabeito <- c('#E69F00', '#56B4E9', '#009E73', '#F0E442', '#0072B2', '#D55E00', '#CC79A7', '#999999', '#000000')
options(ggplot2.discrete.fill = okabeito)
library(ggplot2)
theme_set(theme_minimal())

library(marginaleffects)
library(ggplot2)

set.seed(1024)
n <- 200
d <- data.frame(
  y = rnorm(n),
  cond = as.factor(sample(0:1, n, TRUE)),
  episode = as.factor(sample(0:4, n, TRUE)))

model1 <- lm(y ~ cond * episode, data = d)

p <- predictions(model1, newdata = datagrid(cond = 0:1, episode = 1:3))
ggplot(p, aes(x = cond, y = estimate, shape = episode, color = episode)) +
  geom_point()

## do episodes 1 and 2 differ when `cond=0`
ggplot(p, aes(x = cond, y = estimate, shape = episode, color = episode)) +
  geom_point() +
  geom_segment(aes(x = 1, xend = 1, y = p$estimate[1], yend = p$estimate[2]), color = "black") +
  ggtitle("What is the vertical distance between the linked points?")

comparisons(model1,
  variables = list(episode = 1:2), # comparison of interest
  newdata = datagrid(cond = 0))    # grid

## do cond=0 and cond=1 differ when episode = 1
ggplot(p, aes(x = cond, y = estimate, shape = episode, color = episode)) +
  geom_point() +
  geom_segment(aes(x = 1, xend = 2, y = p$estimate[1], yend = p$estimate[4]), color = okabeito[1]) +
  ggtitle("What is the vertical distance between the linked points?")

comparisons(model1,
  variables = "cond",              # comparison of interest
  newdata = datagrid(episode = 1)) # grid

## Is the difference between episode 1 and 2 larger in cond=0 or cond=1? 
## try this without the `hypothesis` argument to see what we are comparing more clearly
ggplot(p, aes(x = cond, y = estimate, shape = episode, color = episode)) +
  geom_point() +
  annotate("rect", xmin = .9, xmax = 1.1, ymin = p$estimate[1], ymax = p$estimate[2], alpha = .2, fill = "green") +
  annotate("rect", xmin = 1.9, xmax = 2.1, ymin = p$estimate[4], ymax = p$estimate[5], alpha = .2, fill = "orange")  +
  ggtitle("Is the green box taller than the orange box?")

comparisons(model1,
  variables = list(episode = 1:2), # comparison of interest
  newdata = datagrid(cond = 0:1),  # grid
  hypothesis = "b1 = b2")          # hypothesis
```


# Difference-in-Differences(-in-Differences)

Going back to our Titanic example:

```{r}
dat <- "https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Titanic.csv"
dat <- read.csv(dat)
titanic <- glm(Survived ~ PClass * SexCode * Age, data = dat, family = binomial)
```

In this case, a contrast is a difference between predicted probabilities. We can compute that contrast for different types of individuals:

```{r}
comparisons(
  titanic,
  variables = "SexCode",
  newdata = datagrid(PClass = c("1st", "3rd")))
```

One we can notice above, is that the gap in predicted probabilities of survival between men and women is larger in 1st class than in 3rd class. Being a woman matters more for your chances of survival if you travel in first class. Is the difference between those contrasts (diff-in-diff) statistically significant?

To answer this question, we can compute a difference-in-difference using the `hypothesis` argument ([see the Hypothesis vignette for details](hypothesis.html)). For example, using `b1` and `b2` to refer to the contrasts in the first and second rows of the output above, we can test if the difference between the two quantities is different from 0:

```{r}
comparisons(
  titanic,
  hypothesis = "b1 - b2 = 0",
  variables = "SexCode",
  newdata = datagrid(PClass = c("1st", "3rd")))
```

Now, let's say we consider more types of individuals:

```{r}
comparisons(
  titanic,
  variables = "SexCode",
  newdata = datagrid(PClass = c("1st", "3rd"), Age = range))
```

With these results, we could compute a triple difference:

```{r}
comparisons(
  titanic,
  hypothesis = "(b1 - b3) - (b2 - b4) = 0",
  variables = "SexCode",
  newdata = datagrid(PClass = c("1st", "3rd"), Age = range))
```


# Manual computation

Now we show how to use the base `R` `predict()` function to compute some of the same quantities as above. This exercise may be clarifying for some users.

```{r}
grid_50_1_3 <- data.frame(Age = 50, SexCode = 1, PClass = "3rd")
grid_50_1_1 <- data.frame(Age = 50, SexCode = 1, PClass = "1st")
grid_50_0_3 <- data.frame(Age = 50, SexCode = 0, PClass = "3rd")
grid_50_0_1 <- data.frame(Age = 50, SexCode = 0, PClass = "1st")


yhat_50_1_3 <- predict(mod, newdata = grid_50_1_3, type = "response")
yhat_50_1_1 <- predict(mod, newdata = grid_50_1_1, type = "response")
yhat_50_0_3 <- predict(mod, newdata = grid_50_0_3, type = "response")
yhat_50_0_1 <- predict(mod, newdata = grid_50_0_1, type = "response")

## prediction on a grid
predictions(mod, newdata = datagrid(Age = 50, SexCode = 1, PClass = "3rd"))
yhat_50_1_3

## contrast on a grid
comparisons(mod,
  variables = list(PClass = c("1st", "3rd")),
  newdata = datagrid(Age = 50, SexCode = 0:1))

yhat_50_0_3 - yhat_50_0_1
yhat_50_1_3 - yhat_50_1_1

## difference-in-differences 
comparisons(mod,
  variables = list(PClass = c("1st", "3rd")),
  newdata = datagrid(Age = 50, SexCode = 0:1),
  hypothesis = "b1 = b2")

(yhat_50_0_3 - yhat_50_0_1) - (yhat_50_1_3 - yhat_50_1_1)

## average of the empirical distribution of contrasts
avg_comparisons(mod, variables = list(PClass = c("1st", "3rd")), by = "SexCode")

grid_empirical_1_3 <- dat |> subset(SexCode == 1) |> transform(PClass = "3rd")
grid_empirical_1_1 <- dat |> subset(SexCode == 1) |> transform(PClass = "1st")
grid_empirical_0_3 <- dat |> subset(SexCode == 0) |> transform(PClass = "3rd")
grid_empirical_0_1 <- dat |> subset(SexCode == 0) |> transform(PClass = "1st")
yhat_empirical_0_1 <- predict(mod, newdata = grid_empirical_0_1, type = "response")
yhat_empirical_0_3 <- predict(mod, newdata = grid_empirical_0_3, type = "response")
yhat_empirical_1_1 <- predict(mod, newdata = grid_empirical_1_1, type = "response")
yhat_empirical_1_3 <- predict(mod, newdata = grid_empirical_1_3, type = "response")
mean(yhat_empirical_0_3, na.rm = TRUE) - mean(yhat_empirical_0_1, na.rm = TRUE)
mean(yhat_empirical_1_3, na.rm = TRUE) - mean(yhat_empirical_1_1, na.rm = TRUE)
```



# Forward, Backward, Centered, and Custom Differences

By default, the `comparisons()` function computes a "forward" difference. For example, if we ask `comparisons()` to estimate the effect of a 10-unit change in predictor `x` on outcome `y`, `comparisons()` will compare the predicted values with `x` and `x+10`.

```{r}
dat <- mtcars
dat$new_hp <- 49 * (mtcars$hp - min(mtcars$hp)) / (max(mtcars$hp) - min(mtcars$hp)) + 1
mod <- lm(mpg ~ log(new_hp), data = dat)

avg_comparisons(
  mod,
  variables = list(new_hp = 10))
```

We can supply arbitrary functions to create custom differences. These functions must accept a vector of values for the predictor of interest, and return a data frame with the same number of rows as the length, and two columns with the values to compare. For example, we can do:

```{r}
forward_diff <- \(x) data.frame(x, x + 10)
backward_diff <- \(x) data.frame(x - 10, x)
center_diff <- \(x) data.frame(x - 5, x + 5)

avg_comparisons(
  mod,
  variables = list(new_hp = forward_diff))

avg_comparisons(
  mod,
  variables = list(new_hp = backward_diff))

avg_comparisons(
  mod,
  variables = list(new_hp = center_diff))
```

Notice that the last "centered" difference gives the same results as the default `comparisons()` call.

