---
title: "Contrasts" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .4,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

library(marginaleffects)
library(patchwork)
library(ggplot2)

theme_set(theme_minimal())
```

[In a previous vignette](https://vincentarelbundock.github.io/marginaleffects/articles/mfx.html), we introduced the "marginal effect" as a partial derivative. Since derivatives are only properly defined for continuous variables, we cannot use them to interpret the effects of changes in categorical variables. For this, we turn to *contrasts* between [Adjusted predictions.](https://vincentarelbundock.github.io/marginaleffects/articles/predictions.html) In the context of this package, a "Contrast" is defined as:

> The difference between two adjusted predictions, calculated for meaningfully different regressor values (e.g., College graduates vs. Others).

# Simple contrasts

Consider a simple model with a logical and a factor variable:

```{r}
library(marginaleffects)

tmp <- mtcars
tmp$am <- as.logical(tmp$am)
mod <- lm(mpg ~ am + factor(cyl), tmp)
```

The `marginaleffects` function automatically computes contrasts for each level of the categorical variables, relative to the baseline category (`FALSE` for logicals, and the reference level for factors), while holding all other values at their mode or mean:

```{r}
mfx <- marginaleffects(mod)
summary(mfx)
```

The summary printed above says that moving from the reference category `4` to the level `6` on the `cyl` factor variable is associated with a change of `r sprintf("%.3f", tidy(mfx)$estimate[2])` in the adjusted prediction. Similarly, the contrast from `FALSE` to `TRUE` on the `am` variable is equal to `r sprintf("%.3f", tidy(mfx)$estimate[1])`.

We can obtain the same results using the `emmeans` package:

```{r}
library(emmeans)
emm <- emmeans(mod, specs = "cyl")
contrast(emm, method = "revpairwise")

emm <- emmeans(mod, specs = "am")
contrast(emm, method = "revpairwise")
```

# Contrasts with interactions

In models with multiplicative interactions, the contrasts of a categorical variable will depend on the values of the interacted variable:

```{r}
mod_int <- lm(mpg ~ am * factor(cyl), tmp)
```

We can now use the `newdata` argument of the `marginaleffects` function to compute contrasts for different values of the other regressors. As in [the marginal effects vignette](https://vincentarelbundock.github.io/marginaleffects/articles/mfx.html), the `datagrid` function can be handy. Since we only care about the logical `am` contrast, we use the `variables` to indicate the subset of results to report:

```{r}
marginaleffects(mod_int, newdata = datagrid(cyl = tmp$cyl), variables = "am")
```

Once again, we obtain the same results with `emmeans`:

```{r}
emm <- emmeans(mod_int, specs = "am", by = "cyl")
contrast(emm, method = "revpairwise")
```

# Complex queries

As described above, the `marginaleffects` package includes limited support to compute contrasts. Users who require more powerful features are encouraged to consider alternative packages such as [emmeans](https://cran.r-project.org/package=emmeans), [modelbased](https://github.com/easystats/modelbased), or [ggeffects.](https://strengejacke.github.io/ggeffects/index.html) These packages offer useful features such as automatic back-transforms, p value correction for multiple comparisons, and more.
