---
title: "Elasticity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elasticity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

The content of this vignette requires version 0.7.0 of the `marginaleffects` package. If that version is not yet available on CRAN, you can [install it from the Github development website.](https://vincentarelbundock.github.io/marginaleffects/#installation)

In some contexts, it is useful to interpret the results of a regression model in terms of elasticity or semi-elasticity. One strategy to achieve that is to estimate a log-log or a semilog model, where the left and/or right-hand side variables are logged. Another approach is to note that $\frac{\partial ln(x)}{\partial x}=\frac{1}{x}$, and to post-process the marginal effects to transform them into elasticities or semi-elasticities.

For example, say we estimate a linear model of this form:

$$y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \varepsilon$$

Let $\hat{y}$ be the adjusted prediction made by the model for some combination of covariates $x_1$ and $x_2$. The slope with respect to $x_1$ (or "marginal effect") is:

$$\frac{\partial \hat{y}}{\partial x_1}$$

We can estimate the "eyex", "eydx", and "dyex" (semi-)elasticities with respect to $x_1$ as follows:

$$
\eta_1=\frac{\partial \hat{y}}{\partial x_1}\cdot \frac{x_1}{\hat{y}}\\
\eta_2=\frac{\partial \hat{y}}{\partial x_1}\cdot \frac{1}{\hat{y}} \\
\eta_3=\frac{\partial \hat{y}}{\partial x_1}\cdot x_1,
$$

with interpretations roughly as follows:

1. A percentage point increase in $x_1$ is associated to a $\eta_1$ percentage points increase in $y$.
2. A unit increase in $x_1$ is associated to a $\eta_2$ percentage points increase in $y$.
3. A percentage point increase in $x_1$ is associated to a $\eta_3$ units increase in $y$.

For further intuition, consider the ratio of change in $y$ to change in $x$: $\frac{\Delta y}{\Delta x}$. We can turn this ratio into a ratio between *relative* changes by dividing both the numerator and the denominator: $\frac{\frac{\Delta y}{y}}{\frac{\Delta x}{x}}$. This is of course linked to the expression for the $\eta_1$ elasticity above.

With the `marginaleffects` package, these quantities are easy to compute:

```{r}
library(marginaleffects)
mod <- lm(mpg ~ hp + wt, data = mtcars)

marginaleffects(mod) |> summary()

marginaleffects(mod, slope = "eyex") |> summary()

marginaleffects(mod, slope = "eydx") |> summary()

marginaleffects(mod, slope = "dyex") |> summary()
```
