---
title: "Experiments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(width = 1000)
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

# Regression adjustment in experiments

Many analysts who conduct and analyze experiments wish to use regression adjustment with a linear regression model to improve the precision of their estimate of the treatment effect. Unfortunately, regression adjustment can introduce small-sample bias and other undesirable properties (Freedman 2008). Lin (2013) proposes a simple strategy to fix these problems in sufficiently large samples:

1. Center all predictors by subtracting each of their means.
2. Estimate a linear model in which the treatment is interacted with each of the covariates.

The `estimatr` package includes a convenient function to implement this strategy:

```{r}
library(estimatr)
library(marginaleffects)
lalonde <- read.csv("https://vincentarelbundock.github.io/Rdatasets/csv/MatchIt/lalonde.csv")

mod <- lm_lin(
    re78 ~ treat,
    covariates = ~ age + educ + race,
    data = lalonde,
    se_type = "HC3")
summary(mod)
```

We can obtain the same results by fitting a model with the standard `lm` function and using the `comparisons()` function:

```{r}
mod <- lm(re78 ~ treat * (age + educ + race), data = lalonde)
comparisons(
    mod,
    variables = "treat",
    vcov = "HC3") |>
    averages()
```

Notice that the `treat` coefficient and associate standard error in the `lm_lin` regression are exactly the same as the estimates produced by the `comparisons()` function.

## References

* Freedman, David A. “On Regression Adjustments to Experimental Data.” Advances in Applied Mathematics 40, no. 2 (February 2008): 180–93.
* Lin, Winston. “Agnostic Notes on Regression Adjustments to Experimental Data: Reexamining Freedman’s Critique.” Annals of Applied Statistics 7, no. 1 (March 2013): 295–318. https://doi.org/10.1214/12-AOAS583.
