
# Extensions 


```{r, include = FALSE}
options(width = 1000)
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .4,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

This vignette shows how to add support for new models and add new functionality for supported models.

## Support a new model type

It is very easy to add support for new models in `marginaleffects`.  All we need is to set a global option and define 4 very simple functions.

If you add support for a class of models produced by a CRAN package, please consider submitting your code for inclusion in the package: https://github.com/vincentarelbundock/marginaleffects

If you add support for a class of models produced by a package hosted *elsewhere* than CRAN, you can submit it for inclusion in the *unsupported* user-submitted library of extensions: Currently 

* [`countreg` package](https://github.com/vincentarelbundock/marginaleffects/blob/main/sandbox/methods_countreg.R). Thanks to Olivier Beaumais.
* [`censreg` package](https://github.com/vincentarelbundock/marginaleffects/blob/main/sandbox/methods_censReg.R). Thanks to Oleg Komashko.

The rest of this section illustrates how to add support for a very simple lm_manual model.

### Fit function

To begin, we define a function which fits a model. Normally, this function will be supplied by a modeling package published on CRAN. Here, we create a function called `lm_manual()`, which estimates a linear regression model using simple linear algebra operates:

```{r}
lm_manual <- function(f, data, ...) {
    # design matrix
    X <- model.matrix(f, data = data)
    # response matrix
    Y <- data[[as.character(f[2])]]
    # coefficients
    b <- solve(crossprod(X)) %*% crossprod(X, Y)
    Yhat <- X %*% b
    # variance-covariance matrix
    e <- Y - Yhat
    df <- nrow(X) - ncol(X)
    s2 <- sum(e^2) / df
    V <- s2 * solve(crossprod(X))
    # model object
    out <- list(
        d = data,
        f = f,
        X = X,
        Y = Y,
        V = V,
        b = b)
    # class name: lm_manual
    class(out) <- c("lm_manual", "list")
    return(out)
}
```

Important: The custom fit function must assign a new class name to the object it returns. In the example above, the model is assigned to be of class `lm_manual` (see the penultimate line of code in the function).

Our new function replicates the results of `lm()`:

```{r}
model <- lm_manual(mpg ~ hp + drat, data = mtcars)
model$b

model_lm <- lm(mpg ~ hp + drat, data = mtcars)
coef(model_lm)
```

### `marginaleffects` extension

To extend support in `marginaleffects`, the first step is to tell the package that our new class is supported. We do this by defining a global option:

```{r}
library(marginaleffects)

options("marginaleffects_model_classes" = "lm_manual")
```

Then, we define 4 methods:

1. `get_coef()`
    - Mandatory arguments: `model`, `...`
    - Returns: named vector of parameters (coefficients).
1. `set_coef()`
    - Mandatory arguments: `model`, `coefs` (named vector of coefficients), `...`
    - Returns: A new model object in which the original coefficients were replaced by the new vector.
    - [Example](https://github.com/vincentarelbundock/marginaleffects/blob/main/R/set_coef.R)
1. `get_vcov()`
    - Mandatory arguments: `model`, `...`.
    - Optional arguments: `vcov`
    - Returns: A named square variance-covariance matrix.
1. `get_predict()`
    - Mandatory arguments: `model`, `newdata` (data frame), `...`
    - Option arguments: `type` and other model-specific arguments.
    - Returns: A data frame with two columns: a unique `rowid` and a column of `estimate` values.

Note that each of these methods will be named with the suffix `.lm_manual` to indicate that they should be used whenever `marginaleffects` needs to process an object of class `lm_manual`.

```{r}
get_coef.lm_manual <- function(model, ...) {
    b <- model$b
    b <- setNames(as.vector(b), row.names(b))
    return(b)
}

set_coef.lm_manual <- function(model, coefs, ...) {
    out <- model
    out$b <- coefs
    return(out)
}

get_vcov.lm_manual <- function(model, ...) {
    return(model$V)
}

get_predict.lm_manual <- function(model, newdata, ...) {
    newX <- model.matrix(model$f, data = newdata)
    Yhat <- newX %*% model$b
    out <- data.frame(
        rowid = seq_len(nrow(Yhat)),
        estimate = as.vector(Yhat))
    return(out)
}
```

The methods we just defined work as expected:

```{r}
get_coef(model)

get_vcov(model)

get_predict(model, newdata = head(mtcars))
```

Now we can use the `avg_slopes` function:

```{r}
avg_slopes(model, newdata = mtcars, variables = c("hp", "drat"))

predictions(model, newdata = mtcars) |> head()
```

Note that, for custom model, we typically have to supply values for the `newdata` and `variables` arguments explicitly.


## Merge your extension into the main package

If you wrote a working extension, please consider contributing back to the community by submitting your code for inclusion in the main package. To do this, all you need to do is follow the steps above, and then:

1. Add the package to the permanent list of supported models, to avoid calling `options()` in every session: `R/sanitize_model.R`
2. Add the package to the vignette of supported models by editing: `data-raw/supported_models.csv`
3. Add a bullet point to the news file: `NEWS.md`

## Modify or extend supported models

Let's say you want to estimate a model using the `mclogit::mblogit` function.  That package is already supported by `marginaleffects`, but you want to use a `type` (scale) of predictions that is not currently supported: a "centered link scale." 

To achieve this, we would need to override the `get_predict.mblogit()` method. However, it can be unsafe to reassign methods supplied by a package that we loaded with `library`. To be safe, we assign a new model class to our object ("customclass") which will inherit from `mblogit`. Then, we define a `get_predict.customclass` method to make our new kinds of predictions.  

Load libraries, estimate a model:

```{r}
library(mclogit)
library(data.table)

model <- mblogit(
    factor(gear) ~ am + mpg,
    data = mtcars,
    trace = FALSE)
```

Tell `marginaleffects` that we are adding support for a new class model models, and assign a new inherited class name to a duplicate of the model object:

```{r}
options("marginaleffects_model_classes" = "customclass")

model_custom <- model

class(model_custom) <- c("customclass", class(model))
```

Define a new `get_predict.customclass` method. We use the default `predict()` function to obtain predictions. Since this is a multinomial model, `predict()` returns a matrix of predictions with one column per level of the response variable.

Our new `get_predict.customclass` method takes this matrix of predictions, modifies it, and reshapes it to return a data frame with three columns: `rowid`, `group`, and `estimate`:

```{r}
get_predict.customclass <- function(model, newdata, ...) {
    out <- predict(model, newdata = newdata, type = "link")
    out <- cbind(0, out)
    colnames(out)[1] <- dimnames(model$D)[[1]][[1]]
    out <- out - rowMeans(out)
    out <- as.data.frame(out)
    out$rowid <- seq_len(nrow(out))
    out <- data.table(out)
    out <- melt(
        out,
        id.vars = "rowid",
        value.name = "estimate",
        variable.name = "group")
}
```

Finally, we can call any `slopes` function and obtain results. Notice that our object of class `customclass` now produces different results than the default `mblogit` object:

```{r}
avg_predictions(model)

avg_predictions(model_custom)
```
