---
title: "Hypothesis Tests and Custom Contrasts using the Delta Method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hypothesis Tests and Custom Contrasts using the Delta Method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

The code in this vignette requires `marginaleffects` version 0.6.0 or [the development version hosted on Github.](https://vincentarelbundock.github.io/marginaleffects/#installation)

This vignette introduces the `deltamethod()` function, and the `hypothesis` argument of the `comparisons()`, `marginaleffects()`, and `predictions()` function. These features allow users to conduct linear and non-linear hypothesis tests and to compute custom contrasts (linear combinations) between parameters.

# Delta method

Version 0.6.0 of `marginaleffects` includes a very simple yet powerful function called `deltamethod()`. This function emulates the behavior of the well-established `car::deltaMethod` and `car::linearHypothesis` functions, but it supports more models, requires fewer dependencies, and offers some convenience features like shortcuts for robust standard errors.

`deltamethod()` can be used to compute estimates and standard errors of arbitrary functions of model parameters. For example, it can be used to conduct tests of equality between coefficients, or to test the value of some linear or non-linear combination of quantities of interest. `deltamethod()` can also be used to conduct hypothesis tests on other functions of a model's parameter, such as adjusted predictions or marginal effects.

Let's start by estimating a simple model:

```{r}
library(marginaleffects)
mod <- lm(mpg ~ hp + wt + factor(cyl), data = mtcars)
```

When the `FUN` and `hypothesis` arguments of `deltamethod()` equal `NULL` (the default), the function returns a data.frame of raw estimates:

```{r}
deltamethod(mod)
```

Test of equality between coefficients:

```{r}
deltamethod(mod, "hp = wt")
```

Non-linear function of coefficients

```{r}
deltamethod(mod, "exp(hp + wt) = 0.1")
```

The `vcov` argument behaves in the same was as in the `marginaleffects()` function. It allows us to easily compute robust standard errors:

```{r}
deltamethod(mod, "hp = wt", vcov = "HC3")
```

We can use shortcuts like `b1`, `b2`, `...` to identify the position of each parameter in the output of `FUN`. For example, `b2=b3` is equivalent to `hp=wt` because those term names appear in the 2nd and 3rd row when we call `deltamethod(mod)`. 

```{r}
deltamethod(mod, "b2 = b3")
```

Term names with special characters must be enclosed in backticks:

```{r}
deltamethod(mod, "`factor(cyl)6` = `factor(cyl)8`")
```

The `FUN` argument can be used to compute standard errors for arbitrary functions of model parameters. This user-supplied function must accept a single model object, and return a numeric vector or a data.frame with two columns named `term` and `estimate`.

```{r}
mod <- glm(am ~ hp + mpg, data = mtcars, family = binomial)

f <- function(x) predict(x, type = "link", newdata = mtcars)
p <- deltamethod(mod, FUN = f)
head(p)
```

Test of equality between two predictions (row 2 vs row 3):

```{r}
f <- function(x) predict(x, newdata = mtcars)
deltamethod(mod, FUN = f, hypothesis = "b2 = b3")
```

Note that we specified the `newdata` argument in the `f` function. This is because the `predict()` method associated with `lm` objects will automatically the original fitted values when `newdata` is `NULL`, instead of returning the slightly altered fitted values which we need to compute numerical derivatives in the delta method.

We can also use numeric vectors to specify linear combinations of parameters. For example, there are 3 coefficients in the last model we estimated. To test the null hypothesis that the sum of the 2nd and 3rd coefficients is equal to 0, we can do:

```{r}
deltamethod(mod, hypothesis = c(0, 1, 1))
```

See below for more example of how to use string formulas, numeric vectors, or matrices to calculate custom contrasts, linear combinations, and linear or non-linear hypothesis tests.

# `hypothesis` Formulas

Each of the 4 core functions of the package support a `hypothesis` argument which behaves similarly to the `deltamethod()` function. This argument allows users to specify custom hypothesis tests and contrasts, asking questions such as: 

* Are coefficients $\beta_1$ and $\beta_2$ equal?
* Are the marginal effects of $X_1$ and $X_2$ equal?
* Is the marginal effect of $X$ when $W=0$ equal to the marginal effect of $X$ when $W=1$?
* Is a non-linear function of adjusted predictions equal to 100?
* Is the marginal mean in the control group equal to the average of marginal means in 3 treatment arms?
* Cross-level contrasts: In a multinomial model, is the effect of $X$ greater on the 1st or the 2nd outcome level?

## Marginal effects

For example, let's fit a model and compute some [marginal effects at the mean:](https://vincentarelbundock.github.io/marginaleffects/articles/marginaleffects.html#marginal-effect-at-the-mean-mem)

```{r}
library(marginaleffects)

mod <- lm(mpg ~ am + vs, data = mtcars)

marginaleffects(mod, newdata = "mean")
```

Is the marginal effect of `am` different from the marginal effect of `vs`? To answer this question we can run a linear hypothesis test:

```{r}
library(marginaleffects)

mod <- lm(mpg ~ am + vs, data = mtcars)

marginaleffects(
    mod,
    newdata = "mean",
    hypothesis = "am = vs")
```

The `hypothesis` string can include any valid `R` expression, so we can run some silly non-linear tests:

```{r}
marginaleffects(
    mod,
    newdata = "mean",
    hypothesis = "exp(am) - 2 * vs = -400")
```

## Adjusted Predictions

Now consider the case of adjusted predictions:

```{r}
p <- predictions(
    mod,
    newdata = datagrid(am = 0:1, vs = 0:1))
p
```

Since there is no `term` column in the output of the `predictions` function, we must use parameter identifiers like `b1`, `b2`, etc. to determine which estimates we want to compare:

```{r}
predictions(
    mod,
    hypothesis = "b1 = b2",
    newdata = datagrid(am = 0:1, vs = 0:1))

p$predicted[1] - p$predicted[2]
```

In the next section, we will see that we can get equivalent results by using a vector of contrast weights, which will be used to compute a linear combination of estimates:

```{r}
predictions(
    mod,
    hypothesis = c(1, -1, 0, 0),
    newdata = datagrid(am = 0:1, vs = 0:1))
```

There are *many* more possibilities:

```{r}
predictions(
    mod,
    hypothesis = "b1 + b2 = 30",
    newdata = datagrid(am = 0:1, vs = 0:1))

p$predicted[1] + p$predicted[2] - 30

predictions(
    mod,
    hypothesis = "(b2 - b1) / (b3 - b2) = 0",
    newdata = datagrid(am = 0:1, vs = 0:1))
```

## Average contrasts or marginal effects

The standard workflow with the `marginaleffects` package is to first call a function like `marginaleffects()` or `comparisons()` to compute unit-level quantities, and then to call `summary()` to aggregate the unit-level quantities into "Average Marginal Effects" or "Average Contrasts." Unfortunately, the `summary()` and `tidy()` functions do not have a `hypothesis` argument, so we cannot use it to conduct hypothesis tests on aggregated quantities. Instead, we can use the `transform_pre` argument to emulate the behavior of `summary()`, computing average marginal effects in a single step.

First, note that these three commands produce the same results:

```{r}
comparisons(mod) |>
    tidy()

comparisons(
    mod,
    transform_pre = "differenceavg")
```

Notice that in the last one we did not need to use `tidy()` or `summary()`, and yet we still obtained average contrasts. [See the transformations section of the Contrasts vignette for more details.](https://vincentarelbundock.github.io/marginaleffects/articles/contrasts.html)

With these results in hand, we can finally conduct a linear hypothesis test between average marginal effects:

```{r, warning = FALSE}
comparisons(
    mod,
    hypothesis = "am = vs",
    transform_pre = "differenceavg")
```

Computing contrasts between average marginal effects requires a little care to obtain the right scale. In particular, we need to specify both the `variables` and the `transform_pre`:

```{r, warning = FALSE}
comparisons(
    mod,
    hypothesis = "am = vs",
    variables = list(am = "dydx", vs = "dydx"),
    transform_pre = "dydxavg")
```

# `hypothesis` Vectors and Matrices

The `marginalmeans()` function computes [estimated marginal means.](https://vincentarelbundock.github.io/marginaleffects/articles/marginalmeans.html) The `hypothesis` argument of that function offers a powerful mechanism to estimate custom contrasts between marginal means, by way of linear combination.

Consider a simple example:

```{r}
library(marginaleffects)
library(emmeans)
library(nnet)

dat <- mtcars
dat$carb <- factor(dat$carb)
dat$cyl <- factor(dat$cyl)
dat$am <- as.logical(dat$am)

mod <- lm(mpg ~ carb + cyl, dat)
mm <- marginalmeans(mod, variables = "carb")
mm
```

The contrast between marginal means for `carb==1` and `carb==2` is:

```{r}
21.66232 - 21.34058 
```

or

```{r}
21.66232 + -(21.34058)
```

or

```{r}
sum(c(21.66232, 21.34058) * c(1, -1))
```

or 

```{r}
c(21.66232, 21.34058) %*% c(1, -1)
```

The last two commands express the contrast of interest as [a linear combination](https://en.wikipedia.org/wiki/Linear_combination) of marginal means.

## Simple contrast

In the `marginalmeans()` function, we can supply a `hypothesis` argument to compute linear combinations of marginal means. This argument must be a numeric vector of the same length as the number of rows in the output of `marginalmeans()`. For example, in the previous there were six rows, and the two marginal means we want to compare are at in the first two positions:

```{r}
lc <- c(1, -1, 0, 0, 0, 0)
marginalmeans(mod, variables = "carb", hypothesis = lc)
```

## Complex contrast

Of course, we can also estimate more complex contrasts:

```{r}
lc <- c(-2, 1, 1, 0, -1, 1)
marginalmeans(mod, variables = "carb", hypothesis = lc)
```

`emmeans` produces similar results:

```{r}
library(emmeans)
em <- emmeans(mod, "carb")
lc <- data.frame(custom_contrast = c(-2, 1, 1, 0, -1, 1))
contrast(em, method = lc)
```

## Multiple contrasts

Users can also compute multiple linear combinations simultaneously by supplying a numeric matrix to `hypothesis`. This matrix must have the same number of rows as the output of `marginaleffects()`, and each column represents a distinct set of weights for different linear combinations. For example:

```{r}
lc <- matrix(c(
    -2, 1, 1, 0, -1, 1,
    1, -1, 0, 0, 0, 0
    ), ncol = 2)
lc

marginalmeans(mod, variables = "carb", hypothesis = lc)
```

## Contrasts across response levels

In models with multinomial outcomes, one may be interested in comparing outcomes or contrasts across response levels. For example, in this model there are 18 estimated marginal means, across 6 outcome levels (the `group` column):

```{r}
library(nnet)
mod <- multinom(carb ~ mpg + cyl, data = dat, trace = FALSE)
mm <- marginalmeans(mod, type = "probs")
mm
```

Let's contrast the marginal means in the first outcome level when `cyl` equals 4 and 6. These marginal means are located in rows 1 and 7 respectively:

```{r}
lc <- rep(0, nrow(mm))
lc[1] <- -1
lc[7] <- 1
marginalmeans(
    mod,
    type = "probs",
    hypothesis = lc)
```

This is indeed equal to the results we would have obtained manually:

```{r}
2.828726e-01 - 3.678521e-01
```

Now let's say we want to calculate a "contrast in contrasts", that is, the outcome of a 3-step process:

1. Contrast between `cyl=6` and `cyl=4` in the 1st outcome level
2. Contrast between `cyl=6` and `cyl=4` in the 2nd outcome level
3. Contrast between the contrasts defined in steps 1 and 2.

We create the linear combination weights as follows:

```{r}
lc <- rep(0, nrow(mm))
lc[c(1, 8)] <- -1
lc[c(7, 2)] <- 1
```

To make sure that the weights are correct, we can display them side by side with the original `marginalmeans()` output:

```{r}
transform(mm[, 1:3], lc = lc)
```

Compute the results:

```{r}
marginalmeans(mod, type = "probs", hypothesis = lc)
```

# Pairwise contrasts: Difference-in-Differences

Now we illustrate how to use the machinery described above to do pairwise comparisons between contrasts, a type of analysis often associated with a "Difference-in-Differences" research design.

First, we simultate data with two treatment groups and pre/post periods:

```{r, message = FALSE}
library(data.table)

N <- 1000
did <- data.table(
    id = 1:N,
    pre = rnorm(N),
    trt = sample(0:1, N, replace = TRUE))
did$post <- did$pre + did$trt * 0.3 + rnorm(N)
did <- melt(
    did,
    value.name = "y",
    variable.name = "time",
    id.vars = c("id", "trt"))
head(did)
```

Then, we estimate a linear model with a multiple interaction between the time and the treatment indicators. We alco compute contrasts at the mean for each treatment level:

```{r}
did_model <- lm(y ~ time * trt, data = did)

comparisons(
    did_model,
    newdata = datagrid(trt = 0:1),
    variables = "time")
```

Finally, we compute pairwise differences between contrasts. This is the Diff-in-Diff estimate:

```{r}
comparisons(
    did_model,
    variables = "time",
    newdata = datagrid(trt = 0:1),
    hypothesis = "pairwise")
```
