---
title: "Hypothesis Tests and Custom Contrasts using the Delta Method"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hypothesis Tests and Custom Contrasts using the Delta Method}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(width = 1000)
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

The code in this vignette requires `marginaleffects` version 0.6.0 or [the development version hosted on Github.](https://vincentarelbundock.github.io/marginaleffects/#installation)

This vignette introduces the `hypotheses()` function, and the `hypothesis` argument of the `comparisons()`, `slopes()`, and `predictions()` function. These features allow users to conduct linear and non-linear hypothesis tests and to compute custom contrasts (linear combinations) between parameters.

# Null hypothesis

The simplest way to modify a hypothesis test is to change the null hypothesis. By default, all functions in the `marginaleffects` package assume that the null is 0. This can be changed by changing the `hypothesis` argument. 

For example, consider a logistic regression model:

```{r}
library(marginaleffects)
mod <- glm(am ~ hp + drat, data = mtcars, family = binomial)
```

We can compute the predicted outcome for a hypothetical unit where all regressors are fixed to their sample means:

```{r}
predictions(mod, newdata = "mean")
```

The Z statistic and p value reported above assume that the null hypothesis equals zero. We can change the null with the `hypothesis` argument:

```{r}
predictions(mod, newdata = "mean", hypothesis = .5)
```

This can obviously be useful in other contexts. For instance, if we compute risk ratios (at the mean) associated with an increase of 1 unit in `hp`, it makes more sense to test the null hypothesis that the ratio of predictions is 1 rather than 0:

```{r}
comparisons(
    mod,
    newdata = "mean",
    variables = "hp",
    transform_pre = "ratio",
    hypothesis = 1)
```

Warning: Z statistics and p values are computed *before* applying functions in `transform_post`.

# Hypothesis tests with the delta method

Version 0.6.0 of `marginaleffects` includes powerful function called `hypotheses()`. This function emulates the behavior of the well-established `car::deltaMethod` and `car::linearHypothesis` functions, but it supports more models, requires fewer dependencies, and offers some convenience features like shortcuts for robust standard errors.

`hypotheses()` can be used to compute estimates and standard errors of arbitrary functions of model parameters. For example, it can be used to conduct tests of equality between coefficients, or to test the value of some linear or non-linear combination of quantities of interest. `hypotheses()` can also be used to conduct hypothesis tests on other functions of a model's parameter, such as adjusted predictions or marginal effects.

Let's start by estimating a simple model:

```{r}
library(marginaleffects)
mod <- lm(mpg ~ hp + wt + factor(cyl), data = mtcars)
```

When the `FUN` and `hypothesis` arguments of `hypotheses()` equal `NULL` (the default), the function returns a data.frame of raw estimates:

```{r}
hypotheses(mod)
```

Test of equality between coefficients:

```{r}
hypotheses(mod, "hp = wt")
```

Non-linear function of coefficients

```{r}
hypotheses(mod, "exp(hp + wt) = 0.1")
```

The `vcov` argument behaves in the same was as in the `slopes()` function. It allows us to easily compute robust standard errors:

```{r}
hypotheses(mod, "hp = wt", vcov = "HC3")
```

We can use shortcuts like `b1`, `b2`, `...` to identify the position of each parameter in the output of `FUN`. For example, `b2=b3` is equivalent to `hp=wt` because those term names appear in the 2nd and 3rd row when we call `hypotheses(mod)`. 

```{r}
hypotheses(mod, "b2 = b3")
```

Term names with special characters must be enclosed in backticks:

```{r}
hypotheses(mod, "`factor(cyl)6` = `factor(cyl)8`")
```

The `FUN` argument can be used to compute standard errors for arbitrary functions of model parameters. This user-supplied function must accept a single model object, and return a numeric vector or a data.frame with two columns named `term` and `estimate`.

```{r}
mod <- glm(am ~ hp + mpg, data = mtcars, family = binomial)

f <- function(x) predict(x, type = "link", newdata = mtcars)
p <- hypotheses(mod, FUN = f)
head(p)
```

Test of equality between two predictions (row 2 vs row 3):

```{r}
f <- function(x) predict(x, newdata = mtcars)
hypotheses(mod, FUN = f, hypothesis = "b2 = b3")
```

Note that we specified the `newdata` argument in the `f` function. This is because the `predict()` method associated with `lm` objects will automatically the original fitted values when `newdata` is `NULL`, instead of returning the slightly altered fitted values which we need to compute numerical derivatives in the delta method.

We can also use numeric vectors to specify linear combinations of parameters. For example, there are 3 coefficients in the last model we estimated. To test the null hypothesis that the sum of the 2nd and 3rd coefficients is equal to 0, we can do:

```{r}
hypotheses(mod, hypothesis = c(0, 1, 1))
```

See below for more example of how to use string formulas, numeric vectors, or matrices to calculate custom contrasts, linear combinations, and linear or non-linear hypothesis tests.

# `hypotheses` Formulas

Each of the 4 core functions of the package support a `hypothesis` argument which behaves similarly to the `hypotheses()` function. This argument allows users to specify custom hypothesis tests and contrasts, in order to test null hypotheses such as:

* The coefficients $\beta_1$ and $\beta_2$ are equal.
* The marginal effects of $X_1$ and $X_2$ equal.
* The marginal effect of $X$ when $W=0$ is equal to the marginal effect of $X$ when $W=1$.
* A non-linear function of adjusted predictions is equal to 100.
* The marginal mean in the control group is equal to the average of marginal means in the other 3 treatment arms.
* Cross-level contrasts: In a multinomial model, the effect of $X$ on the 1st outcome level is equal to the effect of $X$ on the 2nd outcome level.

## Marginal effects

For example, let's fit a model and compute some [marginal effects at the mean:](https://vincentarelbundock.github.io/marginaleffects/articles/slopes.html#marginal-effect-at-the-mean-mem)

```{r}
library(marginaleffects)

mod <- lm(mpg ~ am + vs, data = mtcars)

mfx <- slopes(mod, newdata = "mean")
mfx
```

Is the marginal effect of `am` different from the marginal effect of `vs`? To answer this question we can run a linear hypothesis test using the `hypotheses` function:

```{r}
hypotheses(mfx, hypothesis = "am = vs")
```

Alternatively, we can specify the hypothesis directly in the original call:

```{r}
library(marginaleffects)

mod <- lm(mpg ~ am + vs, data = mtcars)

slopes(
    mod,
    newdata = "mean",
    hypothesis = "am = vs")
```

The `hypotheses` string can include any valid `R` expression, so we can run some silly non-linear tests:

```{r}
slopes(
    mod,
    newdata = "mean",
    hypothesis = "exp(am) - 2 * vs = -400")
```

## Adjusted Predictions

Now consider the case of adjusted predictions:

```{r}
p <- predictions(
    mod,
    newdata = datagrid(am = 0:1, vs = 0:1))
p
```

Since there is no `term` column in the output of the `predictions` function, we must use parameter identifiers like `b1`, `b2`, etc. to determine which estimates we want to compare:

```{r}
hypotheses(p, hypothesis = "b1 = b2")
```

Or directly:

```{r}
predictions(
    mod,
    hypothesis = "b1 = b2",
    newdata = datagrid(am = 0:1, vs = 0:1))

p$estimate[1] - p$estimate[2]
```

In the next section, we will see that we can get equivalent results by using a vector of contrast weights, which will be used to compute a linear combination of estimates:

```{r}
predictions(
    mod,
    hypothesis = c(1, -1, 0, 0),
    newdata = datagrid(am = 0:1, vs = 0:1))
```

There are *many* more possibilities:

```{r}
predictions(
    mod,
    hypothesis = "b1 + b2 = 30",
    newdata = datagrid(am = 0:1, vs = 0:1))

p$estimate[1] + p$estimate[2] - 30

predictions(
    mod,
    hypothesis = "(b2 - b1) / (b3 - b2) = 0",
    newdata = datagrid(am = 0:1, vs = 0:1))
```

## Average contrasts or marginal effects

The standard workflow with the `marginaleffects` package is to call a function like `predictions()`, `slopes()` or `comparisons()` to compute unit-level quantities; or one of their cousins `avg_predictions()`, `avg_comparisons()`, or `avg_slopes()` to aggregate the unit-level quantities into "Average Marginal Effects" or "Average Contrasts." We can also use the `transform_pre` argument to emulate the behavior of the `avg_*()` functions.

First, note that these three commands produce the same results:

```{r}
comparisons(mod, variables = "vs")$estimate |> mean()

avg_comparisons(mod, variables = "vs")

comparisons(
    mod,
    variables = "vs",
    transform_pre = "differenceavg")
```

[See the transformations section of the Contrasts vignette for more details.](https://vincentarelbundock.github.io/marginaleffects/articles/contrasts.html)

With these results in hand, we can now conduct a linear hypothesis test between average marginal effects:

```{r, warning = FALSE}
comparisons(
    mod,
    hypothesis = "am = vs",
    transform_pre = "differenceavg")
```

Computing contrasts between average marginal effects requires a little care to obtain the right scale. In particular, we need to specify both the `variables` and the `transform_pre`:

```{r, warning = FALSE}
comparisons(
    mod,
    hypothesis = "am = vs",
    variables = c("am", "vs"),
    transform_pre = "dydxavg")
```

# `hypotheses` Vectors and Matrices

The `marginal_means()` function computes [estimated marginal means.](https://vincentarelbundock.github.io/marginaleffects/articles/marginalmeans.html) The `hypothesis` argument of that function offers a powerful mechanism to estimate custom contrasts between marginal means, by way of linear combination.

Consider a simple example:

```{r}
library(marginaleffects)
library(emmeans)
library(nnet)

dat <- mtcars
dat$carb <- factor(dat$carb)
dat$cyl <- factor(dat$cyl)
dat$am <- as.logical(dat$am)

mod <- lm(mpg ~ carb + cyl, dat)
mm <- marginal_means(mod, variables = "carb")
mm
```

The contrast between marginal means for `carb==1` and `carb==2` is:

```{r}
21.66232 - 21.34058 
```

or

```{r}
21.66232 + -(21.34058)
```

or

```{r}
sum(c(21.66232, 21.34058) * c(1, -1))
```

or 

```{r}
c(21.66232, 21.34058) %*% c(1, -1)
```

The last two commands express the contrast of interest as [a linear combination](https://en.wikipedia.org/wiki/Linear_combination) of marginal means.

## Simple contrast

In the `marginal_means()` function, we can supply a `hypothesis` argument to compute linear combinations of marginal means. This argument must be a numeric vector of the same length as the number of rows in the output of `marginal_means()`. For example, in the previous there were six rows, and the two marginal means we want to compare are at in the first two positions:

```{r}
lc <- c(1, -1, 0, 0, 0, 0)
marginal_means(mod, variables = "carb", hypothesis = lc)
```

## Complex contrast

Of course, we can also estimate more complex contrasts:

```{r}
lc <- c(-2, 1, 1, 0, -1, 1)
marginal_means(mod, variables = "carb", hypothesis = lc)
```

`emmeans` produces similar results:

```{r}
library(emmeans)
em <- emmeans(mod, "carb")
lc <- data.frame(custom_contrast = c(-2, 1, 1, 0, -1, 1))
contrast(em, method = lc)
```

## Multiple contrasts

Users can also compute multiple linear combinations simultaneously by supplying a numeric matrix to `hypotheses`. This matrix must have the same number of rows as the output of `slopes()`, and each column represents a distinct set of weights for different linear combinations. The column names of the matrix become labels in the output. For example:

```{r}
lc <- matrix(c(
    -2, 1, 1, 0, -1, 1,
    1, -1, 0, 0, 0, 0
    ), ncol = 2)
colnames(lc) <- c("Contrast A", "Contrast B")
lc

marginal_means(mod, variables = "carb", hypothesis = lc)
```

## Contrasts across response levels

In models with multinomial outcomes, one may be interested in comparing outcomes or contrasts across response levels. For example, in this model there are 18 estimated marginal means, across 6 outcome levels (the `group` column):

```{r}
library(nnet)
mod <- multinom(carb ~ mpg + cyl, data = dat, trace = FALSE)
mm <- marginal_means(mod, type = "probs")
mm
```

Let's contrast the marginal means in the first outcome level when `cyl` equals 4 and 6. These marginal means are located in rows 1 and 7 respectively:

```{r}
lc <- rep(0, nrow(mm))
lc[1] <- -1
lc[7] <- 1
marginal_means(
    mod,
    type = "probs",
    hypothesis = lc)
```

This is indeed equal to the results we would have obtained manually:

```{r}
2.828726e-01 - 3.678521e-01
```

Now let's say we want to calculate a "contrast in contrasts", that is, the outcome of a 3-step process:

1. Contrast between `cyl=6` and `cyl=4` in the 1st outcome level
2. Contrast between `cyl=6` and `cyl=4` in the 2nd outcome level
3. Contrast between the contrasts defined in steps 1 and 2.

We create the linear combination weights as follows:

```{r}
lc <- rep(0, nrow(mm))
lc[c(1, 8)] <- -1
lc[c(7, 2)] <- 1
```

To make sure that the weights are correct, we can display them side by side with the original `marginal_means()` output:

```{r}
transform(mm[, 1:3], lc = lc)
```

Compute the results:

```{r}
marginal_means(mod, type = "probs", hypothesis = lc)
```

# Pairwise contrasts: Difference-in-Differences

Now we illustrate how to use the machinery described above to do pairwise comparisons between contrasts, a type of analysis often associated with a "Difference-in-Differences" research design.

First, we simulate data with two treatment groups and pre/post periods:

```{r, message = FALSE}
library(data.table)

N <- 1000
did <- data.table(
    id = 1:N,
    pre = rnorm(N),
    trt = sample(0:1, N, replace = TRUE))
did$post <- did$pre + did$trt * 0.3 + rnorm(N)
did <- melt(
    did,
    value.name = "y",
    variable.name = "time",
    id.vars = c("id", "trt"))
head(did)
```

Then, we estimate a linear model with a multiple interaction between the time and the treatment indicators. We also compute contrasts at the mean for each treatment level:

```{r}
did_model <- lm(y ~ time * trt, data = did)

comparisons(
    did_model,
    newdata = datagrid(trt = 0:1),
    variables = "time")
```

Finally, we compute pairwise differences between contrasts. This is the Diff-in-Diff estimate:

```{r}
comparisons(
    did_model,
    variables = "time",
    newdata = datagrid(trt = 0:1),
    hypothesis = "pairwise")
```


# Equivalence, non-inferiority, and non-superiority tests

In many contexts, analysts are less interested in rejecting a null hypothesis, and more interested in testing whether an estimate is "inferior", "superior", or "equivalent" to a given threshold or interval. For example, medical researchers may wish to determine if the estimated effect of a new treatment is larger than the effect of prior treatments, or larger than some threshold of "clinical significance." Alternatively, researchers may wish to support a claim that an estimated parameter is "equivalent to" or "not meaningfully different from" a null hypothesis.

To answer these questions, we can use non-inferiority, non-superiority, or equivalence tests like the two-one-sided test (TOST). This article gives a primer and tutorial on TOST:

> Lakens D, Scheel AM, Isager PM. Equivalence Testing for Psychological Research: A Tutorial. Advances in Methods and Practices in Psychological Science. 2018;1(2):259-269. doi:10.1177/2515245918770963

The `hypotheses()` function of the `marginaleffects` package includes an `equivalence` argument which allows users to apply these tests to any of the quantities generated by the package, as well as to arbitrary functions of a model's parameters. To illustrate, we begin by estimating a simple linear regression model:

```{r}
mod <- lm(mpg ~ hp + factor(gear), data = mtcars)
```

The rest of this section considers several quantities estimated by `marginaleffects`.

#### Example 1: Predictions

Consider a single prediction, where all predictors are held at their median or mode:

```{r}
p <- predictions(mod, newdata = "median")
p
```

```{r, include = FALSE}
options(width = 120)
hyp <- hypotheses(p, equivalence = c(17, 18))
```

Now we specify an equivalence interval (or "region") for predictions between 17 and 18:

```{r}
hypotheses(p, equivalence = c(17, 18))
```

The results allow us to draw three conclusions:

1. The p value for the non-superiority test is `r sprintf("%.4f", hyp$p.value.nonsup)`. This suggests that we *cannot* reject the hypothesis that the parameter (`r sprintf("%.4f", hyp$estimate)`) is above the [17,18] interval.
2. The p value for the non-inferiority test is `r sprintf("%.4f", hyp$p.value.noninf)`. This suggests that we *can* reject the hypothesis that the parameter is below the [17,18] interval.
3. The p value for the equivalence test is `r sprintf("%.4f", hyp$p.value.equiv)`. This suggests that we *cannot* reject the hypothesis that the parameter falls outside the equivalence region.

#### Example 2: Model coefficients

The `hypotheses` function also allows users to conduct equivalence, non-inferiority, and non-superiority tests for model coefficients, and for arbitrary functions of model coefficients.

Our estimate of the 4th coefficient in the model is:

```{r}
coef(mod)[4]
```

We can test if this parameter is likely to fall in the [5,7] interval by:

```{r}
hypotheses(mod, equivalence = c(5, 7))[4, ]
```

```{r, include = FALSE}
hyp <- hypotheses(mod, equivalence = c(5, 7))[4, "p.value.equiv"]
```

The p value is `r sprintf("%.4f", hyp)`, so we cannot reject the hypothesis that the `r names(coef(mod))[4]` parameter falls outside the [5,7] interval. 

#### Example 3: Slopes

The same syntax can be used to conduct tests for all the quantities produced by the `marginaleffects` package. For example, imagine that, for substantive or theoretical reasons, an average slope between -0.1 and 0.1 is uninteresting. We can conduct an equivalence test to check if this is the case:

```{r}
avg_slopes(mod, variables = "hp") |>
    hypotheses(equivalence = c(-.1, .1))
```
```{r, include = FALSE}
hyp <- avg_slopes(mod, variables = "hp", equivalence = c(-.1, .1))$p.value.equiv
```

The p value is `r sprintf("%.4f", hyp)`, which suggests that we can reject the hypothesis that the parameter falls outside the equivalence region (or "substantive irrelevance region"). 

#### Example 4: Difference between comparisons (contrasts)

Consider a model with a multiplicative interaction:

```{r}
int <- lm(mpg ~ hp * factor(gear), data = mtcars)
```

The average contrast for a change of 1 unit in `hp` differs based on the value of `gear`:

```{r}
avg_comparisons(int, variables = "hp", by = "gear")
```

Are these contrasts different from one another? Let's look at the pairwise differences between them:

```{r}
avg_comparisons(int, variables = "hp", by = "gear", hypothesis = "pairwise")
```

We consider that these pairwise comparisons are "equivalent to zero" when they fall in the [-.1, .1] interval:

```{r}
avg_comparisons(int, variables = "hp", by = "gear", hypothesis = "pairwise") |>
    hypotheses(equivalence = c(-.1, .1))
```

The `p (Eq.)` column shows that the difference between the average contrasts when `gear` is 3 and `gear` is 5 can be said to be equivalent to the specified interval. However, there are good reasons to think that the other two pairwise comparisons may fall outside the interval.

#### Example 5: Marginal means and `emmeans`

This final example shows the equivalence between results produced by the `emmeans` package and the `marginal_means()` function:

```{r}
library(emmeans)

mod <- lm(log(conc) ~ source + factor(percent), data = pigs)

# {emmeans}
emmeans(mod, specs = "source") |>
    pairs() |>
    test(df = Inf,
         null = 0,
         delta = log(1.25),
         side = "equivalence",
         adjust = "none")

# {marginaleffects}
marginal_means(
    mod,
    variables = "source",
    hypothesis = "pairwise") |>
    hypotheses(equivalence = c(-log(1.25), log(1.25)))
```


#### Example 6: t-test

Now we show that the results produced by `hypotheses()` are identical to the results produced by the `equivalence` package in the case of a simple t-test:

```{r}
library(equivalence)

set.seed(1024)

# simulate data data
N <- 20
dat <- data.frame(
    y = rnorm(N),
    x = sample(c(rep(0, N / 2), rep(1, N / 2)), N))

# fit model
mod <- lm(y ~ x, data = dat)

# test with the {equivalence} package
e <- tost(
    x = dat$y[dat$x == 0],
    y = dat$y[dat$x == 1],
    epsilon = 10)
e

# test with {marginaleffects} package
hypotheses(mod, equivalence = c(-10, 10), df = e$parameter)[2,]
```