---
title: "Hypothesis Tests and Custom Contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hypothesis Tests and Custom Contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

The code in this vignette requires `marginaleffects` version 0.5.1 or [the development version hosted on Github.](https://vincentarelbundock.github.io/marginaleffects/#installation)

# Hypothesis Test Formulas: Linear and Non-Linear

Each of the 4 core functions of the package support a `hypothesis` argument which allows users to specify custom hypothesis tests and contrasts. 


## Marginal effects

For example, let's fit a model and compute some [marginal effects at the mean:](https://vincentarelbundock.github.io/marginaleffects/articles/mfx03_mfx.html#marginal-effect-at-the-mean-mem)

```{r}
library(marginaleffects)

mod <- lm(mpg ~ am + vs, data = mtcars)

marginaleffects(mod, newdata = "mean")
```

Is the marginal effect of `am` different from the marginal effect of `vs`? To answer this question we can run a linear hypothesis test:

```{r}
library(marginaleffects)

mod <- lm(mpg ~ am + vs, data = mtcars)

marginaleffects(
    mod,
    newdata = "mean",
    hypothesis = "am = vs")
```

The `hypothesis` string can include any valid `R` expression, so we can run some silly non-linear tests:

```{r}
marginaleffects(
    mod,
    newdata = "mean",
    hypothesis = "exp(am) - 2 * vs = -400")
```

## Adjusted Predictions

Now consider the case of adjusted predictions:

```{r}
p <- predictions(
    mod,
    newdata = datagrid(am = 0:1, vs = 0:1))
p
```

Since there is no `term` column in the output of the `predictions` function, we must use row identifiers like `r1`, `r2`, etc. to determine which estimates we want to compare:

```{r}
predictions(
    mod,
    hypothesis = "r1 = r2",
    newdata = datagrid(am = 0:1, vs = 0:1))

p$predicted[1] - p$predicted[2]
```

In the next section, we will see that we can get equivalent results by using a vector of contrast weights, which will be used to compute a linear combination of estimates:

```{r}
predictions(
    mod,
    hypothesis = c(1, -1, 0, 0),
    newdata = datagrid(am = 0:1, vs = 0:1))
```

There are *many* more possibilities:

```{r}
predictions(
    mod,
    hypothesis = "r1 + r2 = 30",
    newdata = datagrid(am = 0:1, vs = 0:1))

p$predicted[1] + p$predicted[2] - 30

predictions(
    mod,
    hypothesis = "(r2 - r1) / (r3 - r2) = 0",
    newdata = datagrid(am = 0:1, vs = 0:1))
```

## Average contrasts or marginal effects

The standard workflow with the `marginaleffects` package is to first call a function like `marginaleffects()` or `comparisons()` to compute unit-level quantities, and then to call `summary()` to aggregate the unit-level quantities into "Average Marginal Effects" or "Average Contrasts." Unfortunately, the `summary()` and `tidy()` functions do not have a `hypothesis` argument, so we cannot use it to conduct hypothesis tests on aggregated quantities. Instead, we can use the `transform_pre` argument to emulate the behavior of `summary()`, computing average marginal effects in a single step.

First, note that these three commands produce the same results:

```{r}
comparisons(mod) |>
    tidy()

comparisons(
    mod,
    transform_pre = "differenceavg")
```

Notice that in the last one we did not need to use `tidy()` or `summary()`, and yet we still obtained average contrasts. [See the transformations vignette for more details.](https://vincentarelbundock.github.io/marginaleffects/articles/transformation.html)

With these results in hand, we can finally conduct a linear hypothesis test between average marginal effects:

```{r, warning = FALSE}
comparisons(
    mod,
    hypothesis = "am = vs",
    transform_pre = "differenceavg")
```

Computing contrasts between average marginal effects requires a little care to obtain the right scale. In particular, we need to specify both the `variables` and the `transform_pre`:

```{r, warning = FALSE}
comparisons(
    mod,
    hypothesis = "am = vs",
    variables = list(am = "dydx", vs = "dydx"),
    transform_pre = "dydxavg")
```


# Pairwise contrasts & Difference-in-Differences

Now we illustrate how to use the machinery described above to do pairwise comparisons between contrasts, a type of analysis often associated with a "Difference-in-Differences" research design.

First, we simultate data with two treatment groups and pre/post periods:

```{r, message = FALSE}
library(data.table)

N <- 1000
did <- data.frame(
    id = 1:N,
    pre = rnorm(N),
    trt = sample(0:1, N, replace = TRUE))
did$post <- did$pre + did$trt * 0.3 + rnorm(N)
did <- melt(
    did,
    value.name = "y",
    variable.name = "time",
    id.vars = c("id", "trt"))
head(did)
```

Then, we estimate a linear model with a multiple interaction between the time and the treatment indicators. We alco compute contrasts at the mean for each treatment level:

```{r}
did_model <- lm(y ~ time * trt, data = did)

comparisons(
    did_model,
    newdata = datagrid(trt = 0:1),
    variables = "time")
```

Finally, we compute pairwise differences between contrasts. This is the Diff-in-Diff estimate:

```{r}
comparisons(
    did_model,
    variables = "time",
    newdata = datagrid(trt = 0:1),
    hypothesis = "pairwise")
```

# Custom Contrasts with Vectors and Matrices

The `marginalmeans()` function computes [estimated marginal means.](https://vincentarelbundock.github.io/marginaleffects/articles/mfx04_marginalmeans.html) The `hypothesis` argument of that function offers a powerful mechanism to estimate custom contrasts between marginal means, by way of linear combination.

Consider a simple example:

```{r}
library(marginaleffects)
library(emmeans)
library(nnet)

dat <- mtcars
dat$carb <- factor(dat$carb)
dat$cyl <- factor(dat$cyl)
dat$am <- as.logical(dat$am)

mod <- lm(mpg ~ carb + cyl, dat)
mm <- marginalmeans(mod, variables = "carb")
mm
```

The contrast between marginal means for `carb==1` and `carb==2` is:

```{r}
21.66232 - 21.34058 
```

or

```{r}
21.66232 + -(21.34058)
```

or

```{r}
sum(c(21.66232, 21.34058) * c(1, -1))
```

or 

```{r}
c(21.66232, 21.34058) %*% c(1, -1)
```

The last two commands express the contrast of interest as [a linear combination](https://en.wikipedia.org/wiki/Linear_combination) of marginal means.

## Simple contrast between marginal means

In the `marginalmeans()` function, we can supply a `hypothesis` argument to compute linear combinations of marginal means. This argument must be a numeric vector of the same length as the number of rows in the output of `marginalmeans()`. For example, in the previous there were six rows, and the two marginal means we want to compare are at in the first two positions:

```{r}
lc <- c(1, -1, 0, 0, 0, 0)
marginalmeans(mod, variables = "carb", hypothesis = lc)
```

## Complex contrast between marginal means

Of course, we can also estimate more complex contrasts:

```{r}
lc <- c(-2, 1, 1, 0, -1, 1)
marginalmeans(mod, variables = "carb", hypothesis = lc)
```

`emmeans` produces similar results:

```{r}
library(emmeans)
em <- emmeans(mod, "carb")
lc <- data.frame(custom_contrast = c(-2, 1, 1, 0, -1, 1))
contrast(em, method = lc)
```

# Multiple custom contrasts

Users can also compute multiple linear combinations simultaneously by supplying a numeric matrix to `hypothesis`. This matrix must have the same number of rows as the output of `marginaleffects()`, and each column represents a distinct set of weights for different linear combinations. For example:

```{r}
lc <- matrix(c(
    -2, 1, 1, 0, -1, 1,
    1, -1, 0, 0, 0, 0
    ), ncol = 2)
lc

marginalmeans(mod, variables = "carb", hypothesis = lc)
```

# Linear combinations across response levels

In models with multinomial outcomes, one may be interested in comparing outcomes or contrasts across response levels. For example, in this model there are 18 estimated marginal means, across 6 outcome levels (the `group` column):

```{r}
library(nnet)
mod <- multinom(carb ~ mpg + cyl, data = dat, trace = FALSE)
mm <- marginalmeans(mod, type = "probs")
mm
```

Let's contrast the marginal means in the first outcome level when `cyl` equals 4 and 6. These marginal means are located in rows 1 and 7 respectively:

```{r}
lc <- rep(0, nrow(mm))
lc[1] <- -1
lc[7] <- 1
marginalmeans(
    mod,
    type = "probs",
    hypothesis = lc)
```

This is indeed equal to the results we would have obtained manually:

```{r}
2.828726e-01 - 3.678521e-01
```

Now let's say we want to calculate a "contrast in contrasts", that is, the outcome of a 3-step process:

1. Contrast between `cyl=6` and `cyl=4` in the 1st outcome level
2. Contrast between `cyl=6` and `cyl=4` in the 2nd outcome level
3. Contrast between the contrasts defined in steps 1 and 2.

We create the linear combination weights as follows:

```{r}
lc <- rep(0, nrow(mm))
lc[c(1, 8)] <- -1
lc[c(7, 2)] <- 1
```

To make sure that the weights are correct, we can display them side by side with the original `marginalmeans()` output:

```{r}
transform(mm[, 1:3], lc = lc)
```

Compute the results:

```{r}
marginalmeans(mod, type = "probs", hypothesis = lc)
```
