---
title: "Linear Combinations and Custom Contrasts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear Combinations and Custom Contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

The code in this vignette requires `marginaleffects` version 0.5.1 or the development version hosted on Github.

The `marginalmeans()` function computes [estimated marginal means.](https://vincentarelbundock.github.io/marginaleffects/articles/mfx04_marginalmeans.html) The `lincom` argument of that function offers a powerful mechanism to estimate custom contrasts between marginal means, by way of linear combination.

Consider a simple example:

```{r}
library(marginaleffects)
library(emmeans)
library(nnet)

dat <- mtcars
dat$carb <- factor(dat$carb)
dat$cyl <- factor(dat$cyl)
dat$am <- as.logical(dat$am)

mod <- lm(mpg ~ carb + cyl, dat)
mm <- marginalmeans(mod, variables = "carb")
mm
```

The contrast between marginal means for `carb==1` and `carb==2` is:

```{r}
21.66232 - 21.34058 
```

or

```{r}
21.66232 + -(21.34058)
```

or

```{r}
sum(c(21.66232, 21.34058) * c(1, -1))
```

or 

```{r}
c(21.66232, 21.34058) %*% c(1, -1)
```

The last two commands express the contrast of interest as [a linear combination](https://en.wikipedia.org/wiki/Linear_combination) of marginal means.

# Simple contrast between marginal means

In the `marginalmeans()` function, we can supply a `lincom` argument to compute linear combinations of marginal means. This argument must be a numeric vector of the same length as the number of rows in the output of `marginalmeans()`. For example, in the previous there were six rows, and the two marginal means we want to compare are at in the first two positions:

```{r}
lc <- c(1, -1, 0, 0, 0, 0)
marginalmeans(mod, variables = "carb", lincom = lc)
```

# Complex contrast between marginal means

Of course, we can also estimate more complex contrasts:

```{r}
lc <- c(-2, 1, 1, 0, -1, 1)
marginalmeans(mod, variables = "carb", lincom = lc)
```

`emmeans` produces similar results:

```{r}
library(emmeans)
em <- emmeans(mod, "carb")
lc <- data.frame(custom_contrast = c(-2, 1, 1, 0, -1, 1))
contrast(em, method = lc)
```

# Multiple custom contrasts

Users can also compute multiple linear combinations simultaneously by supplying a numeric matrix to `lincom`. This matrix must have the same number of rows as the output of `marginaleffects()`, and each column represents a distinct set of weights for different linear combinations. For example:

```{r}
lc <- matrix(c(
    -2, 1, 1, 0, -1, 1,
    1, -1, 0, 0, 0, 0
    ), ncol = 2)
lc

marginalmeans(mod, variables = "carb", lincom = lc)
```

# Linear combinations across response levels

In models with multinomial outcomes, one may be interested in comparing outcomes or contrasts across response levels. For example, in this model there are 18 estimated marginal means, across 6 outcome levels (the `group` column):

```{r}
library(nnet)
mod <- multinom(carb ~ mpg + cyl, data = dat, trace = FALSE)
mm <- marginalmeans(mod, type = "probs")
mm
```

Let's contrast the marginal means in the first outcome level when `cyl` equals 4 and 6. These marginal means are located in rows 1 and 7 respectively:

```{r}
lc <- rep(0, nrow(mm))
lc[1] <- -1
lc[7] <- 1
marginalmeans(
    mod,
    type = "probs",
    lincom = lc)
```

This is indeed equal to the results we would have obtained manually:

```{r}
2.828726e-01 - 3.678521e-01
```

Now let's say we want to calculate a "contrast in contrasts", that is, the outcome of a 3-step process:

1. Contrast between `cyl=6` and `cyl=4` in the 1st outcome level
2. Contrast between `cyl=6` and `cyl=4` in the 2nd outcome level
3. Contrast between the contrasts defined in steps 1 and 2.

We create the linear combination weights as follows:

```{r}
lc <- rep(0, nrow(mm))
lc[c(1, 8)] <- -1
lc[c(7, 2)] <- 1
```

To make sure that the weights are correct, we can display them side by side with the original `marginalmeans()` output:

```{r}
transform(mm[, 1:3], lc = lc)
```

Compute the results:

```{r}
marginalmeans(mod, type = "probs", lincom = lc)
```
