---
title: "Marginal Means" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Marginal Means}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .4,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
library("tidyverse")
library("kableExtra")
```

In the context of this package, "marginal means" refer to the values obtained by this three step process:

1. Construct a "grid" of predictor values with all combinations of categorical variables, and where numeric variables are held at their means.
2. Calculate adjusted predictions for each cell in that grid.
3. Take the average of those adjusted predictions across one dimension of the grid to obtain the marginal means.

For example, consider a model with a numeric, a factor, and a logical predictor:

```{r}
library(marginaleffects)

dat <- mtcars
dat$cyl <- as.factor(dat$cyl)
dat$am <- as.logical(dat$am)
mod <- lm(mpg ~ hp + cyl + am, data = dat)
```

Using the `predictions` function, we set the `hp` variable at its mean and compute predictions for all combinations for `am` and `cyl`:

```{r}
predictions(mod, variables = c("am", "cyl"))
```

For illustration purposes, it is useful to reshape the above results:

```{r, echo = FALSE}
pred <- predictions(mod, variables = c("am", "cyl")) %>%
    select(cyl, am, predicted) %>%
    pivot_wider(names_from = "am", values_from = "predicted") %>%
    rowwise() %>%
    mutate(`Marginal mean of cyl` = mean(c(`TRUE`, `FALSE`)))
row <- data.frame(x = "Marginal means of am",
                  y = mean(pred[["TRUE"]]),
                  z = mean(pred[["FALSE"]]))
colnames(row) <- colnames(pred)[1:3]
pred <- bind_rows(pred, row)
for (i in 2:ncol(pred)) {
    pred[[i]] <- sprintf("%.1f", pred[[i]])
}
pred[pred == "NA"] <- ""
kbl(pred) %>% 
    kable_styling() %>%
    add_header_above(c(" " = 1, "am" = 2, " " = 1))
```

The marginal means of `am` and `cyl` are obtained by taking the mean of the adjusted predictions across cells. The `marginalmeans` function gives us the same results easily:

```{r}
marginalmeans(mod)
```

The same results can be obtained using the [very powerful `emmeans` package](https://cran.r-project.org/package=emmeans):

```{r}
library(emmeans)
emmeans(mod, specs = "cyl")
emmeans(mod, specs = "am")
```


# Tidy summaries

The `summary`, `tidy`, and `glance` functions are also available to summarize and manipulate the results:

```{r}
me <- marginalmeans(mod)

tidy(me)

glance(me)

summary(me)
```

Thanks to those tidiers, we can also present the results in the style of a regression table [using the `modelsummary` package:](https://vincentarelbundock.github.io/modelsummary/)

```{r}
library("modelsummary")

modelsummary(me,
             title = "Estimated Marginal Means",
             estimate = "{estimate} ({std.error}){stars}",
             statistic = NULL,
             group = term + group ~ model)
```
