---
title: "Contrasts" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contrasts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .4,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

library(marginaleffects)
library(patchwork)
library(ggplot2)

theme_set(theme_minimal())
```

[In a previous vignette](https://vincentarelbundock.github.io/marginaleffects/articles/mfx.html), we introduced the "marginal effect" as a partial derivative. Since derivatives are only properly defined for continuous variables, we cannot use them to interpret the effects of changes in categorical variables. For this, we turn to *contrasts* between [Adjusted predictions.](https://vincentarelbundock.github.io/marginaleffects/articles/predictions.html) In the context of this package, a "Contrast" is defined as:

> The difference between two adjusted predictions, calculated for meaningfully different regressor values (e.g., College graduates vs. Others).

The `marginaleffects()` function automatically calculates contrasts instead of derivatives for factor, logical, or character variables. The `comparisons()` function gives users more powerful features to compute different contrasts.

# Contrasts: logical and factor variables

Consider a simple model with a logical and a factor variable:

```{r}
library(marginaleffects)
library(magrittr)

tmp <- mtcars
tmp$am <- as.logical(tmp$am)
mod <- lm(mpg ~ am + factor(cyl), tmp)
```

The `marginaleffects` function automatically computes contrasts for each level of the categorical variables, relative to the baseline category (`FALSE` for logicals, and the reference level for factors), while holding all other values at their mode or mean:

```{r}
mfx <- marginaleffects(mod)
summary(mfx)
```

The summary printed above says that moving from the reference category `4` to the level `6` on the `cyl` factor variable is associated with a change of `r sprintf("%.3f", tidy(mfx)$estimate[2])` in the adjusted prediction. Similarly, the contrast from `FALSE` to `TRUE` on the `am` variable is equal to `r sprintf("%.3f", tidy(mfx)$estimate[1])`.

We can obtain different contrasts by using the `comparisons()` function. For example:


```{r}
comparisons(mod, contrast_factor = "sequential") %>% tidy()
comparisons(mod, contrast_factor = "pairwise") %>% tidy()
comparisons(mod, contrast_factor = "reference") %>% tidy()
```

For comparison, this code produces the same results using the `emmeans` package:

```{r}
library(emmeans)
emm <- emmeans(mod, specs = "cyl")
contrast(emm, method = "revpairwise")

emm <- emmeans(mod, specs = "am")
contrast(emm, method = "revpairwise")
```

Note that these commands also work on for other types of models, such as GLMs, on different scales:

```{r}
mod_logit <- glm(am ~ factor(gear), data = mtcars, family = binomial)

comparisons(mod_logit) %>% tidy()

comparisons(mod_logit, type = "link") %>% tidy()
```

# Contrasts: numeric variables

We can also compute contrasts for differences in numeric variables. For example, we can see what happens to the adjusted predictions when we increment the `hp` variable by 1 unit (default) or by 5 units:

```{r}
mod <- lm(mpg ~ hp, data = mtcars)

comparisons(mod) %>% tidy()

comparisons(mod, contrast_numeric = 5) %>% tidy()
```

Compare adjusted predictions for a change in the regressor between two arbitrary values:

```{r}
comparisons(mod, contrast_numeric = c(90, 110)) %>% tidy()
```

Compare adjusted predictions when the regressor changes across the interquartile range, across one or two standard deviations about its mean, or from across its full range:

```{r}
comparisons(mod, contrast_numeric = "iqr") %>% tidy()

comparisons(mod, contrast_numeric = "sd") %>% tidy()

comparisons(mod, contrast_numeric = "2sd") %>% tidy()

comparisons(mod, contrast_numeric = "minmax") %>% tidy()
```

# Interactions between contrasts

In some contexts we would like to know what happens when two predictors change at the same time. In the `marginaleffects` package terminology, this is an "interaction between contrasts."

For example, consider a model with two factor variables: 

```{r}
mod <- lm(mpg ~ am * factor(cyl), data = mtcars)
```

What happens if `am` increases by 1 unit and `cyl` changes from a baseline reference to another level? 

```{r}
comparisons(mod, interaction = TRUE)
```
