---
title: "Multinomial Logit and Discrete Choice Models "
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multinomial Logit and Discrete Choice Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

library(ggplot2)

theme_clean <- function() {
  theme_minimal() +
    theme(panel.grid.minor = element_blank(),
          strip.text = element_text(size = rel(1), hjust = 0),
          strip.background = element_blank(),
          legend.position = "bottom")
}
ggplot2::theme_set(theme_clean())
```

Several packages in the `R` ecosystem allow users to estimate multinomial logit model and discrete choice models. This case study illustrates the use of `marginaleffects` with the `nnet` and `mlogit` packages.

# `nnet` package

The `multinom` function of the `nnet` package allows users to fit log-lienar models via neural networks. The `data` used for this function is a data frame with one observation per row, and the response variable is coded a factor. All the `marginaleffects` package function work seamlessly with this model. For example, we can estimate a model and compute average marginal effects as follows:

```{r, message = FALSE}
library(nnet)
library(tidyverse)

head(mtcars)

mod <- multinom(factor(gear) ~ hp + mpg, data = mtcars, trace = FALSE)

marginaleffects(mod, type = "probs") %>% summary() 
```

Notice that in such models, we get one marginal effect for each term, for each level of the response variable.

# `mlogit` package

The `mlogit` package uses `data` in a slightly different structure, with one row per observation-choice combination. For example, this data on choice of travel mode includes 4 rows per individual, one for each mode of transportation:

```{r, message = FALSE}
library("AER")
library("mlogit")
library(tidyverse)
data("TravelMode", package = "AER")

head(TravelMode)

mod <- mlogit(choice ~ wait + gcost | income + size, TravelMode)

marginaleffects(mod, variables = c("income", "size")) %>% summary()
```

Note that the `marginaleffects` function will always return estimates of zero for regressors before the vertical bar in the formula. This is because the `predict()` function supplied by the `mlogit` package does not produce different predictions for different values of these variables.

To compute different kinds of marginal effects, we can construct customized data frames and feed them to the `newdata` argument of the `marginaleffects` function. 

**Important:** The `newdata` argument for `mlogit` models must be a "balanced" data frame, that is, it must have a number of rows that is a multiple of the number of choices.

If we want to compute the slope of the response function (marginal effecs) when each of the predictors is fixed to its global mean, we can do:

```{r}
nd <- TravelMode %>%
    summarize(across(c("wait", "gcost", "income", "size"),
              function(x) rep(mean(x), 4)))
nd

marginaleffects(mod, newdata = nd, variables = c("income", "size")) |> summary()
```

If we want to compute marginal effects with the `gcost` and `wait` fixed at their mean value, *conditional* on the choice of transportation mode:

```{r}
nd <- TravelMode %>%
    group_by(mode) %>%
    summarize(across(c("wait", "gcost", "income", "size"), mean))
nd

marginaleffects(mod, newdata = nd, variables = c("income", "size")) |> summary()
```

We can also explore more complex alternatives. Here, for example, only one alternative is affected by cost reduction:

```{r}
nd <- datagrid(mode = TravelMode$mode, newdata = TravelMode)
nd <- lapply(1:4, function(i) mutate(nd, gcost = ifelse(1:4 == i, 30, gcost)))
nd <- bind_rows(nd)
nd

marginaleffects(mod, newdata = nd, variables = c("income", "size")) |> summary()
```
