---
title: "Tables and Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tables and Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(width = 1000)
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

Note: This vignette requires a version of `modelsummary` greater than 1.2.0. If this version is not yet available on CRAN, you can install the development version by following the [instructions on the `modelsummary` website.](https://vincentarelbundock.github.io/modelsummary/)

# Marginal effects

We can summarize the results of the `comparisons()` or `slopes()` functions using [the `modelsummary` package.](https://github.com/vincentarelbundock/modelsummary)

```{r}
library(modelsummary)
library(marginaleffects)

mod <- glm(am ~ wt + drat, family = binomial, data = mtcars)
mfx <- slopes(mod)

modelsummary(mfx)
```

The same results can be visualized with `modelplot()`:

```{r}
modelplot(mfx)
```

````{comment}
We can display several models or marginal effects side by side:

```{r}
models <- list(
    glm(am ~ drat, family = binomial, data = mtcars),
    glm(am ~ wt + drat, family = binomial, data = mtcars))
models[[1]] <- slopes(models[[1]])
models[[2]] <- slopes(models[[2]], vcov = "HC3")
modelsummary(models)
```
````

# Contrasts

Note: The code in this section requires a version of the `modelsummary` package greater than 1.2.0. If it is not available from CRAN, you can install the development version by following the instructions on [the `modelsummary` website.](https://vincentarelbundock.github.io/modelsummary/)

When using the `comparisons()` function (or the `slopes()` function with categorical variables), the output will include two columns to uniquely identify the quantities of interest: `term` and `contrast`.

```{r}
dat <- mtcars
dat$gear <- as.factor(dat$gear)
mod <- glm(vs ~ gear + mpg, data = dat, family = binomial)

cmp <- comparisons(mod)
get_estimates(cmp)
```

We can use the `group` argument of the `modelsummary` function to structure the table properly:

```{r}
modelsummary(cmp, group = term + contrast ~ model)
```

Cross-contrasts can be a bit trickier, since there are multiple simultaneous groups. Consider this example:

```{r}
mod <- lm(mpg ~ factor(cyl) + factor(gear), data = mtcars)
cmp <- comparisons(
  mod,
  variables = c("gear", "cyl"),
  cross = TRUE)
get_estimates(cmp)
```

As we can see above, there are two relevant grouping columns: `contrast_gear` and `contrast_cyl`. We can simply plug those names in the `shape` argument:

```{r}
modelsummary(
  cmp,
  shape = contrast_gear + contrast_cyl ~ model)
```

# Marginal means

```{r}
library("marginaleffects")
library("modelsummary")

dat <- mtcars
dat$cyl <- as.factor(dat$cyl)
dat$am <- as.logical(dat$am)
mod <- lm(mpg ~ hp + cyl + am, data = dat)
mm <- marginal_means(mod)

modelsummary(mm,
             title = "Estimated Marginal Means",
             estimate = "{estimate} ({std.error}){stars}",
             statistic = NULL,
             group = term + value ~ model)
```
