---
title: "Tables and Plots"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tables and Plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

# Marginal effects

We can summarize the results of the `comparisons()` or `marginaleffects()` functions using [the `modelsummary` package.](https://github.com/vincentarelbundock/modelsummary)

```{r}
library(modelsummary)
library(marginaleffects)

mod <- glm(am ~ wt + drat, family = binomial, data = mtcars)
mfx <- marginaleffects(mod)

modelsummary(mfx)
```

The same results can be visualized with `modelplot()`:

```{r}
modelplot(mfx)
```

We can display several models or marginal effects side by side:

```{r}
models <- list(
    glm(am ~ drat, family = binomial, data = mtcars),
    glm(am ~ wt + drat, family = binomial, data = mtcars))
models[[1]] <- marginaleffects(models[[1]])
models[[2]] <- marginaleffects(models[[2]], vcov = "HC3")
modelsummary(models)
```

# Contrasts

When using the `comparisons()` function (or the `marginaleffects()` function with categorical variables), the output will include two columns to uniquely identify the quantities of interest: `term` and `contrast`.

```{r}
dat <- mtcars
dat$gear <- as.factor(dat$gear)
mod <- glm(vs ~ gear + mpg, data = dat, family = binomial)

mfx <- marginaleffects(mod)
tidy(mfx)
```

We can use the `group` argument of the `modelsummary` function to structure the table properly:

```{r}
modelsummary(mfx, group = term + contrast ~ model)
```

