---
title: "Multiple Imputation and Missing Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple Imputation and Missing Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(width = 1000)
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

The `marginaleffects` package offers convenience functions to compute and display predictions, contrasts, and marginal effects from models with multiple imputation from the `mice` package. The workflow follows Rubin's rules (Rubin, 1987, p. 76), via the following steps:

1. Impute $M$ data sets.
2. Fit in each of the $M$ imputed data sets.
3. Compute marginal effects in each of the $M$ data sets.
4. Pool results.

To highlight the workflow, we consider a simple linear regression model, although the same workflow should work with any model type which is fit using a formula interface and a `data` argument.

`marginaleffects` support the `mice` imputation package, and any other package which can return a list of imputed data frames. 

# `mice`

First, we insert missing observations randomly in a dataset and we impute the dataset using the `mice` package:

```{r}
library(mice)
library(marginaleffects)
set.seed(1024)

dat <- iris
dat$Sepal.Length[sample(seq_len(nrow(iris)), 40)] <- NA
dat$Sepal.Width[sample(seq_len(nrow(iris)), 40)] <- NA
dat$Species[sample(seq_len(nrow(iris)), 40)] <- NA

dat_mice <- mice(dat, m = 20, printFlag = FALSE, .Random.seed = 1024)
```

Then, we use the standard `mice` syntax to produce an object of class `mira` with all the models:

```{r}
mod_mice <- with(dat_mice, lm(Petal.Width ~ Sepal.Length * Sepal.Width + Species))
```

Finally, we feed the `mira` object to a `marginaleffects` function:

```{r}
mfx_mice <- avg_slopes(mod_mice, by = "Species")
mfx_mice
```

# Other imputation packages: `Amelia`, `missRanger`, or lists of imputed data frames.

Several `R` packages can impute missing data. Indeed, [the `Missing Data CRAN View`](https://cran.r-project.org/web/views/MissingData.html) lists at least a dozen alternatives. Since user interface changes a lot from package to package, `marginaleffects` supports a single workflow which can be used, with some adaptation, to with all imputation packages:

1. Use an external package to create a list of imputed data frames.
2. Apply the `datalist2mids()` function from the `miceadds` package to convert the list of imputed data frames to a `mids` object.
3. Use the `with()` function to fit models, as illustrated in the `mice` section above.
4. Pass the `mids` object to a `marginaleffects` function.
   
Consider two imputation packages, which can both generate lists of imputed datasets: `Amelia` and `missRanger`.

```{r, include = FALSE}
# no startup messages
library(Amelia)
library(miceadds)
library(missRanger)
```

```{r, message = FALSE, warning = FALSE}
library(Amelia)
library(miceadds)
library(missRanger)

# impute data
dat_amelia <- amelia(dat, noms = "Species", p2s = 0)$imputations
mids_amelia <- datlist2mids(dat_amelia)

# convert lists of imputed datasets to `mids` objects
dat_missRanger <- replicate(20, missRanger(dat, verbose = 0), simplify = FALSE)
mids_missRanger <- datlist2mids(dat_missRanger)

# fit models
mod_amelia <- with(mids_amelia, lm(Petal.Width ~ Sepal.Length * Sepal.Width + Species))
mod_missRanger <- with(mids_missRanger, lm(Petal.Width ~ Sepal.Length * Sepal.Width + Species))

# `Amelia` slopes
mfx_amelia <- avg_slopes(mod_amelia, by = "Species")
mfx_amelia

# `missRanger` slopes
mfx_missRanger <- avg_slopes(mod_missRanger, by = "Species")
mfx_missRanger
```

# Comparing results with different imputation software

We can use the `modelsummary` package to compare the results with listwise delection to the results using different imputations software:

```{r}
library(modelsummary)

# listwise deletion slopes
mod_lwd <- lm(Petal.Width ~ Sepal.Length * Sepal.Width + Species, data = dat)
mfx_lwd <- avg_slopes(mod_lwd, by = "Species")

# regression table
models <- list(
    "LWD" = mfx_lwd,
    "mice" = mfx_mice,
    "Amelia" = mfx_amelia,
    "missRanger" = mfx_missRanger)
modelsummary(models, shape = term : contrast + Species ~ model)
```