---
title: "Plots: interactions, predictions, contrasts, and slopes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plots: interactions, predictions, contrasts, and slopes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# this vignette is in .Rbuildignore

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

Note: This vignette requires version 0.7.2 of `marginaleffects`, or [the development version.](https://vincentarelbundock.github.io/marginaleffects/#installation)

The `marginaleffects` package includes several flexible functions to plot estimates and to display interactions:

* Plot the raw output of `slopes()` or `comparisons()`
* Interaction / conditional / heterogenous:
   - `plot_predictions`: Conditional Predictions (equivalent to `predictions()`)
   - `plot_comparisons`: Conditional Contrasts (equivalent to `comparisons()`)
   - `plot_slopes`: Conditional Slopes (aka "Marginal Effects"; equivalent to `slopes()`)

The "interaction" or "conditional" plotting functionality is designed to display how the outcome or how the effect of a predictor changes with respect to another variable.

This vignette focuses on `plot_predictions()`, but the same ideas apply to `plot_slopes()` and `plot_comparisons()` as well.  In the examples below, we will use the `patchwork` package to combine plots and the `ggplot2` package to customize their content and appearance:

```{r}
library(marginaleffects)
library(patchwork)
library(ggplot2)
```

# Interactions: Conditional Adjusted Predictions

Consider a linear model with interactions, where the relationship between `hp` and `mpg` is conditional on the values of `wt` and `cyl`:

```{r}
mod <- lm(mpg ~ hp * wt * factor(cyl), data = mtcars)

plot_predictions(mod, condition = "hp")
```

We can show the predicted values of `mpg` for different values of different predictors:

```{r}
plot_predictions(mod, condition = c("hp", "cyl"))
```

We can include a 3rd conditioning variable, specify what values we want to consider, and use one of several string shortcuts for common reference values ("threenum", "minmax", "quartile", etc.):

```{r, fig.asp = 1}
plot_predictions(mod, condition = list(hp = 110:120, "wt" = "threenum")) /

plot_predictions(mod, condition = list("hp", "wt" = "minmax")) /

plot_predictions(mod, condition = list("hp", "wt" = fivenum)) /

plot_predictions(mod, condition = c("hp", "wt", "cyl"))
```

# Customizing plots

A very useful feature of the plotting functions in this package is that they produce normal `ggplot2` objects. So we can customize them to our heart's content, using `ggplot2` itself, or one of the many packages designed to augment its functionalities:

```{r, fig.asp = .7}
library(ggokabeito)
library(ggrepel)

mt <- mtcars
mt$label <- row.names(mt)

mod <- lm(mpg ~ hp * factor(cyl), data = mt)

plot_predictions(mod, condition = c("hp", "cyl"), vcov = FALSE) +
    geom_point(aes(x = hp, y = mpg, color = factor(cyl)), data = mt) +
    geom_rug(aes(x = hp, y = mpg), data = mt) +
    geom_text_repel(aes(x = hp, y = mpg, label = label),
                    data = subset(mt, hp > 250),
                    nudge_y = 2) +
    theme_classic() +
    scale_color_okabe_ito()
```

All the plotting functions work with all the model supported by the `marginaleffects` package, so we can plot the output of a logistic regression model. This plot shows the probability of survival aboard the Titanic, for different ages and different ticket classes:

```{r, warning = FALSE}
library(ggdist)

dat <- "https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Titanic.csv"
dat <- read.csv(dat)

mod <- glm(Survived ~ Age * PClass, data = dat, family = binomial)

plot_predictions(mod, condition = c("Age", "PClass")) +
    geom_dots(
        alpha = .8,
        scale = .3,
        pch = 18,
        data = dat, aes(
        x = Age,
        y = Survived,
        side = ifelse(Survived == 1, "bottom", "top")))
```

Thanks to [Andrew Heiss](https://bayesf22-notebook.classes.andrewheiss.com/bayes-rules/13-chapter.html) who inspired this plot.

# Fits and smooths

We can compare the model predictors with fits and smoothers using the `geom_smooth()` function from the `ggplot2` package:

```{r}
dat <- "https://vincentarelbundock.github.io/Rdatasets/csv/Stat2Data/Titanic.csv"
dat <- read.csv(dat)
mod <- glm(Survived ~ Age * PClass, data = dat, family = binomial)

plot_predictions(mod, condition = c("Age", "PClass")) +
    geom_smooth(data = dat, aes(Age, Survived), method = "lm", se = FALSE, color = "black") +
    geom_smooth(data = dat, aes(Age, Survived), se = FALSE, color = "black")
```

# Extreme customization

Designing effective data visualizations requires a lot of customization to the specific context and data. The plotting functions in `marginaleffects` offer a powerful way to iterate quickly between plots and models, but they obviously cannot support all the features that users may want. Thankfully, it is very easy to use the `slopes` functions to generate datasets that can then be used in `ggplot2` or any other data visualization tool. Just use the `draw` argument:

```{r}
p <- plot_predictions(mod, condition = c("Age", "PClass"), draw = FALSE)
head(p)
```
