---
title: "Adjusted Predictions" 
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adjusted Predictions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .4,
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

library(marginaleffects)
library(patchwork)
library(ggplot2)

theme_set(theme_minimal())
```

In the context of this package, an "Adjusted Prediction" is defined as:

> The response predicted by a model for some combination of the regressors' values, such as their means or factor levels (a.k.a. "reference grid"). 

An adjusted prediction is thus the regression-adjusted response variable (or link, or other fitted value), for a given combination (or grid) of predictors. This grid may or may not correspond to the actual observations in a dataset.

By default, `predictions` calculates the regression-adjusted predicted values for every observation in the original dataset:


```{r}
library(marginaleffects)

mod <- lm(mpg ~ hp + factor(cyl), data = mtcars)

pred <- predictions(mod)

head(pred)
```

In many cases, this is too limiting, and researchers will want to specify a grid of "typical" values over which to compute adjusted predictions.

# Predictions over a "typical" data grid

There are two main ways to select the reference grid over which we want to compute adjusted predictions. The first is using the `variables` argument. The second is with the `newdata` argument and the [`datagrid()` function](https://vincentarelbundock.github.io/marginaleffects/reference/datagrid.html) that we already introduced in the [marginal effects vignette.](https://vincentarelbundock.github.io/marginaleffects/articles/mfx.html) 

## `variables`: Levels and Tukey's 5 numbers 

The `variables` argument is a handy shortcut to create grids of predictors. Each of the levels of factor/logical/character variables listed in the `variables` argument will be displayed. For numeric variables, `predictions` will compute adjusted predictions at Tukey's 5 summary numbers. All other variables will be set at their means or modes.

```{r}
predictions(mod, variables = c("cyl", "hp"))
```

The `data.frame` produced by `predictions` is "tidy", which makes it easy to manipulate with other `R` packages and functions:

```{r, message = FALSE}
library(kableExtra)
library(tidyverse)

predictions(mod, variables = c("cyl", "hp")) %>%
    select(hp, cyl, predicted) %>%
    pivot_wider(values_from = predicted, names_from = cyl) %>%
    kbl(caption = "A table of Adjusted Predictions") %>%
    kable_styling() %>%
    add_header_above(header = c(" " = 1, "cyl" = 3))
```

## `newdata` and `datagrid`

A second strategy to construct grids of predictors for adjusted predictions is to combine the `newdata` argument and the `datagrid` function. Recall that this function creates a "typical" dataset with all variables at their means or modes, except those we explicitly define:

```{r}
datagrid(cyl = c(4, 6, 8), model = mod)
```

We can also use this `datagrid` function in a `predictions` call (omitting the `model` argument):

```{r}
predictions(mod, newdata = datagrid())

predictions(mod, newdata = datagrid(cyl = c(4, 6, 8)))
```

Users can change the summary function used to summarize each type of variables using the `FUN.numeric`, `FUN.factor`, and related arguments, for example substituting the mean for the median.

## `counterfactual` data grid

An alternative approach to construct grids of predictors is to use `grid.type = "counterfactual"` argument value. This will duplicate the *whole* dataset, with the different values specified by the user. 

For example, the `mtcars` dataset has 32 rows. This command produces a new dataset with 64 rows, with each row of the original dataset duplicated with the two values of the `am` variable supplied (0 and 1):

```{r}
mod <- glm(vs ~ hp + am, data = mtcars, family = binomial)

nd <- datagrid(model = mod, am = 0:1, grid.type = "counterfactual")

dim(nd)
```

Then, we can use this dataset and the `predictions` function to create interesting visualizations:

```{r, fig.asp = 1}
pred <- predictions(mod, newdata = datagrid(am = 0:1, grid.type = "counterfactual")) %>%
    select(am, predicted, rowid_original) %>%
    pivot_wider(id_cols = rowid_original, 
                names_from = am,
                values_from = predicted)

ggplot(pred, aes(x = `0`, y = `1`)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1) +
    labs(x = "Predicted Pr(vs=1), when am = 0",
         y = "Predicted Pr(vs=1), when am = 1")
```

In this graph, each dot represents the predicted probability that `vs=1` for one observation of the dataset, in the counterfactual worlds where `am` is either 0 or 1. 

# Plot: Conditional Predictions

First, we download the `ggplot2movies` dataset from the [RDatasets archive](https://vincentarelbundock.github.io/Rdatasets/articles/data.html). Then, we create a variable called `certified_fresh` for movies with a rating of at least 8. Finally, we discard some outliers and fit a logistic regression model: 

```{r, message = FALSE}
library(tidyverse)
dat <- read.csv("https://vincentarelbundock.github.io/Rdatasets/csv/ggplot2movies/movies.csv") %>%
    mutate(style = case_when(Action == 1 ~ "Action",
                             Comedy == 1 ~ "Comedy",
                             Drama == 1 ~ "Drama",
                             TRUE ~ "Other"),
           style = factor(style),
           certified_fresh = rating >= 8) %>%
    filter(length < 240)

mod <- glm(certified_fresh ~ length * style, data = dat, family = binomial)
```

We can plot adjusted predictions, conditional on the `length` variable using the `plot_cap` function:

```{r}
mod <- glm(certified_fresh ~ length, data = dat, family = binomial)

plot_cap(mod, condition = "length")
```

We can also introduce another condition which will display a categorical variable like `style` in different colors. This can be useful in models with interactions:

```{r}
mod <- glm(certified_fresh ~ length * style, data = dat, family = binomial)

plot_cap(mod, condition = c("length", "style"))
```

Of course, you can also design your own plots or tables by working with the `predictions` output directly:

```{r}
predictions(mod,
            type = c("response", "link"),
            newdata = datagrid(length = 90:120,
                               style = c("Action", "Comedy"))) %>%
    ggplot(aes(length, predicted, color = style))  +
    geom_line() +
    facet_wrap(~type, scales = "free_y")
```

# Prediction types

The `predictions` function computes model-adjusted means on the scale of the output of the `predict(model)` function. By default, `predict` produces predictions on the `"response"` scale, so the adjusted predictions should be interpreted on that scale. However, users can pass a string or a vector of strings to the `type` argument, and `predictions` will consider different outcomes. 

Typical values include `"response"` and `"link"`, but users should refer to the documentation of the `predict` of the package they used to fit the model to know what values are allowable. documentation. 

```{r}
mod <- glm(am ~ mpg, family = binomial, data = mtcars)
pred <- predictions(mod, type = c("response", "link"))
head(pred)
```

Users who need more control over the type of adjusted predictions to compute, including a host of options for back-transformation, may want to consider the [`emmeans` package.](https://cran.r-project.org/package=emmeans)

We can also plot predictions on different outcome scales:

```{r}
plot_cap(mod, condition = "mpg", type = "response")
```

```{r}
plot_cap(mod, condition = "mpg", type = "link")
```
