---
title: "Robust standard errors and more"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Robust standard errors and more}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)
```

The code in this vignette requires installing version 0.17.1 or greater of the `insight` package. On 2022-03-29, this version was not yet available on CRAN, but the development version could be installed from Github:am

```{r}
#| eval = FALSE
library(remotes)
install_github("easystats/insight")
```

Restart `R` completely after installation.

# Robust standard errors 

All the functions in the `marginaleffects` package can compute robust standard errors on the fly for any model type supported by [the `sandwich` package.](https://sandwich.r-forge.r-project.org/) The `vcov` argument supports string shortcuts like `"HC3"`, a one-sided formula to request clustered standard errors, variance-covariance matrices, or functions which return such matrices. Here are a few examples.

Adjusted predictions with classical or heteroskedasticity-robust standard errors:

```{r}
library(marginaleffects)
library(patchwork)
mod <- lm(mpg ~ hp, data = mtcars)

p <- predictions(mod)
head(p, 2)

p <- predictions(mod, vcov = "HC3")
head(p, 2)
```

Marginal effects with cluster-robust standard errors:

```{r}
mfx <- marginaleffects(mod, vcov = ~cyl)
summary(mfx)
```

Comparing adjusted predictions with classical and robust standard errors:

```{r, fig.asp = .4}
p1 <- plot_cap(mod, condition = "hp")
p2 <- plot_cap(mod, condition = "hp", vcov = "HC3")
p1 + p2
```

# Mixed effects models: Satterthwaite and Kenward-Roger corrections

For linear mixed effects models we can apply the Satterthwaite and Kenward-Roger corrections in the same way as above:

```{r, message = FALSE}
library(marginaleffects)
library(patchwork)
library(lme4)

dat <- mtcars
dat$cyl <- factor(dat$cyl)
dat$am <- as.logical(dat$am)
mod <- lmer(mpg ~ hp + am + (1 | cyl), data = dat)
```

Marginal effects at the mean with classical standard errors and z-statistic:

```{r}
marginaleffects(mod, newdata = "mean")
```

Marginal effects at the mean with Kenward-Roger adjusted variance-covariance and degrees of freedom:

```{r}
marginaleffects(mod,
                newdata = "mean",
                vcov = "kenward-roger")
```

We can use the same option in any of the package's core functions, including:

```{r}
plot_cap(mod, condition = "hp", vcov = "satterthwaite")
```

