---
title: "Transformations and Custom Contrasts: Adjusted Risk Ratio Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transformations and Custom Contrasts: Log Adjusted Risk Ratio Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# this vignette is in .Rbuildignore because lme4 is not available on old CRAN
# test machines.

knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 9,
  fig.asp = .4,
  out.width = "100%",
  warning = FALSE,
  message = FALSE,
  comment = "#>"
)

library(ggplot2)

theme_clean <- function() {
  theme_minimal() +
    theme(panel.grid.minor = element_blank(),
          strip.text = element_text(size = rel(1), hjust = 0),
          strip.background = element_blank(),
          legend.position = "bottom")
}
ggplot2::theme_set(theme_clean())
```

[The Contrasts vignette](https://vincentarelbundock.github.io/marginaleffects/articles/mfx03_contrasts.html) defined contrasts as:

> The difference between two adjusted predictions.

This vignette shows how to move beyond such simple differences, by estimating "contrasts" that consist of ratios or arbitrary functions of adjusted predictions. It also shows how to back-transform contrasts to change their scales.

Consider the case of a model with a single predictor $x$. To compute average contrasts, we procede as follows:

1. Compute adjusted predictions for each row of the dataset for the observed values $x$: $\hat{y}_x$
2. Compute adjusted predictions for each row of the dataset for the observed values $x + 1$: $\hat{y}_{x+1}$
3. Compute unit-level contrasts by taking the difference between (or some other function of) adjusted predictions: $\hat{y}_{x+1} - \hat{y}_x$
4. Compute the average contrast by taking the mean of unit-level contrasts: $1/N \sum_{i=1}^N \hat{y}_{x+1} - \hat{y}_x$
5. Return the result directly or transformed.

This vignette shows how to modify steps 3 and 5 to compute custom contrasts: Adjusted Risk Ratios (ARR) and Adjusted Risk Differences (ARD).

# Default contrasts: untransformed differences

In `R` we do:

```{r}
library(marginaleffects)
library(magrittr)

mod <- glm(vs ~ mpg, data = mtcars, family = binomial)

# construct data
mtcars_plus_1 <- mtcars
mtcars_plus_1$mpg <- mtcars_plus_1$mpg + 1

# adjusted predictions
yhat <- predict(mod, newdata = mtcars, type = "response")
yhat_plus_1 <- predict(mod, newdata = mtcars_plus_1, type = "response")

# unit-level contrasts
con <- yhat_plus_1 - yhat

# average contrasts
mean(con)
```

We can use the `comparisons()` and `summary()` functions to obtain the same results:

```{r}
con <- comparisons(mod)
summary(con)
```

# Custom contrasts with `transform_pre`: Ratios

Instead of taking simple differences between adjusted predictions, it can sometimes be useful to compute other forms of contrasts. For example, [the `adjrr` function the `Stata` software package](https://journals.sagepub.com/doi/pdf/10.1177/1536867X1301300304) can compute "adjusted risk ratios", which are ratios of adjusted predictions. To do this in `R`, we use the `transform_pre` argument:

```{r}
comparisons(mod, transform_pre = "ratio") %>% summary()
```

This result is the average adjusted risk ratio, that is, the adjusted predictions when the `mpg` are incremented by 1, divided by the adjusted predictions when `mpg` is at its original value.

The `transform_pre` accepts different values for common types of contrasts: 'difference', 'ratio', 'lnratio', 'ratioavg', 'lnratioavg', 'differenceavg'. These strings are shortcuts for functions that accept two vectors of adjusted predictions and returns a single vector of contrasts. For example, these two commands yield identical results:

```{r}
comparisons(mod, transform_pre = "ratio") %>% summary()

comparisons(mod, transform_pre = function(hi, lo) hi / lo) %>% summary()
```

This mechanism is powerful, because it lets users create fully customized contrasts. Here is a non-sensical example:

```{r}
comparisons(mod, transform_pre = function(hi, lo) sqrt(hi) / log(lo + 10)) %>% summary()
```

# (Back-)Transforming results with `transform_post`: Adjusted Risk Ratios

By default, the standard errors are computed using the delta method on the scale determined by the `type` argument (e.g., "link" or "response"). Some analysts may prefer to proceed differently. For example, in `Stata`, the `adjrr` computes adjusted risk ratios (ARR) in two steps:

1. Compute the natural log of the ratio between the mean of adjusted predictions with $x+1$ and the mean of adjusted predictions with $x$.
2. Exponentiate the estimate and confidence interval bounds.

Step 1 is easy to achieve with the `transform_pre` argument described above. Step 2 can be achieved with the `transform_post` argument:

```{r}
comparisons(
    mod,
    transform_pre = function(hi, lo) log(mean(hi) / mean(lo)),
    transform_post = exp) %>%
    summary()
```

Note that we can use the `lnratioavg` shortcut instead of defining the function ourselves.

The order of operations in previous command was:

1. Compute the custom unit-level contrasts
2. Exponentiate them
3. Take the average using the `summary()` function

There is a very subtle difference between the procedure above and this code:

```{r}
comparisons(
    mod,
    transform_pre = "lnratioavg") %>%
    summary(transform_post = exp)
```

Since the `transform_post` argument is now passed to the `summary()` function, the exponentiation is now done only *after* unit-level contrasts have been average. This is what `Stata` does under the hood, and the results are slightly different.

